---
title: "Cell states along a cotinuum"
subtitle: ""
author: "Jiadong Mao (MIG), Yidi Deng (ANU, Florey Institute)"
date: "6 May 2026"
output:
  xaringan::moon_reader:
    lib_dir: libs
    seal: false
    css: ["default", "afternoon_30min_xaringan.css"]
    nature:
      ratio: "16:9"
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
params:
  render_heavy_nnet: false
---

```{r setup, include=FALSE}
options(htmltools.dir.version = FALSE)

knitr::opts_chunk$set(
    echo = FALSE,
    message = FALSE,
    warning = FALSE,
    fig.align = "center",
    dpi = 180,
    out.width = "100%"
)

ok_pkg <- function(pkg) requireNamespace(pkg, quietly = TRUE)
load_pkg <- function(pkg) {
    if (ok_pkg(pkg)) {
        suppressPackageStartupMessages(library(pkg, character.only = TRUE))
        TRUE
    } else {
        FALSE
    }
}

has_ggplot2 <- load_pkg("ggplot2")
has_patchwork <- load_pkg("patchwork")
has_sce <- ok_pkg("SingleCellExperiment")
has_se <- ok_pkg("SummarizedExperiment")

`%||%` <- function(x, y) if (is.null(x)) y else x
meta <- rmarkdown::metadata
deck_title <- meta$title %||% ""
deck_subtitle <- meta$subtitle %||% ""
deck_author <- meta$author %||% ""
deck_date <- meta$date %||% ""

if (has_ggplot2) {
    theme_set(theme_minimal(base_size = 18))
}

rescale01 <- function(x) {
    rng <- range(x, na.rm = TRUE)
    if (!all(is.finite(rng)) || diff(rng) == 0) return(rep(NA_real_, length(x)))
    (x - rng[1]) / diff(rng)
}

placeholder_plot <- function(title, detail = "Figure placeholder") {
    if (!has_ggplot2) {
        plot.new()
        text(0.5, 0.6, title, cex = 1.2)
        text(0.5, 0.45, detail, cex = 0.9)
        return(invisible(NULL))
    }
    ggplot() +
        annotate("rect", xmin = -Inf, xmax = Inf, ymin = -Inf, ymax = Inf,
                 fill = "#F5F7FA", colour = "#CBD5E1", linewidth = 0.8) +
        annotate("text", x = 0, y = 0.12, label = title,
                 size = 6.5, fontface = "bold", colour = "#111827") +
        annotate("text", x = 0, y = -0.12, label = detail,
                 size = 4.2, colour = "#475569") +
        xlim(-1, 1) + ylim(-0.7, 0.7) +
        theme_void()
}

read_cache <- function(path) {
    if (!file.exists(path)) return(NULL)
    tryCatch(readRDS(path), error = function(e) NULL)
}

plot_umap <- function(df, colour_col, title, continuous = TRUE) {
    p <- ggplot(df, aes(UMAP1, UMAP2, colour = .data[[colour_col]])) +
        geom_point(size = 0.25, alpha = 0.75) +
        labs(title = title, colour = NULL) +
        theme_void(base_size = 15) +
        theme(
            plot.title = element_text(face = "bold", size = 14, hjust = 0.02),
            legend.position = "bottom",
            legend.key.width = grid::unit(1.2, "cm")
        )
    if (continuous) {
        p + scale_colour_distiller(palette = "Spectral", direction = -1)
    } else {
        p + scale_colour_brewer(palette = "Set1")
    }
}

combine2 <- function(p1, p2) {
    if (has_patchwork) p1 | p2 else print(p1)
}

combine4 <- function(p1, p2, p3, p4) {
    if (has_patchwork) (p1 | p2) / (p3 | p4) else print(p1)
}
```

class: title-slide, middle

# `r deck_title`

`r if (nzchar(deck_subtitle)) paste0("## ", deck_subtitle)`

### `r deck_author`

### `r deck_date`

.footnote[
Modules 5-7: Φ-Space -> pseudotime -> NeighbourNet
]

---

class: section-slide, middle

<div class="kicker">Afternoon arc</div>

<div class="claim">The afternoon session asks what we lose when we force continuous biology into discrete bins.</div>

<div class="flow">
  <div class="flow-step">Module 5<br>soft phenotype scores</div>
  <div class="arrow">-></div>
  <div class="flow-step">Module 6<br>maturation ordering</div>
  <div class="arrow">-></div>
  <div class="flow-step">Module 7<br>network rewiring hypotheses</div>
</div>

???
Timing: 2 minutes. Tie back to Session 1: QC, integration, clustering, annotation, DE.

---

# The biological problem

<div class="cols middle">
<div>
<div class="kicker">Human heart development</div>

* The dataset spans **foetal**, **young**, and **adult** heart samples.
* Cardiomyocyte maturation is not a set of clean bins.
* A hard label can be correct and still hide the gradient we care about.
</div>

<div class="panel">
<h3>Morning workflow</h3>
<p>Cluster -> marker genes -> one label per cluster.</p>
<h3>Afternoon question</h3>
<p>Can we preserve gradual identity and use it for trajectory and network analysis?</p>
</div>
</div>

???
Timing: 2 minutes. The core contrast is not that hard labels are bad; they are insufficient for this biological question.

---

# Module 5: Φ-Space turns labels into scores

<div class="cols middle">
<div>
<div class="claim">Ask “how much does this cell resemble each reference phenotype?” rather than “which single label wins?”</div>

* Output: a **cells × phenotype scores** matrix.
* Scores can be collapsed to labels with `getClass()`.
* The reverse is impossible: labels cannot recover the original continuum.
</div>

<div class="panel">
<h3>Method sketch</h3>
<p>Annotated reference expression `X` is mapped onto a dummy phenotype matrix `Y` using PLS regression.</p>
<p>The fitted model is applied to query cells to produce `reducedDim(query, "PhiSpace")`.</p>
</div>
</div>

???
Timing: 3 minutes. Keep PLS high level: components maximize covariance between expression and phenotype labels.

---

# Φ-Space proof object

```{r phispace-proof, fig.width=10.5, fig.height=5.4}
query <- read_cache("results/05_phispace_query.rds")

if (!is.null(query) && has_ggplot2 && has_sce && has_se) {
    phi_scores <- SingleCellExperiment::reducedDim(query, "PhiSpace")
    pc <- prcomp(phi_scores, center = TRUE, scale. = FALSE)
    pc_df <- as.data.frame(pc$x[, 1:2])
    names(pc_df) <- c("PC1", "PC2")
    query_cd <- as.data.frame(SummarizedExperiment::colData(query))
    pc_df$group <- query_cd$group

    age_cols <- intersect(c("fetal", "young", "middle", "old"), colnames(phi_scores))
    cell_type_cols <- setdiff(colnames(phi_scores), age_cols)
    broad <- rep("Cell", nrow(pc_df))
    if (length(cell_type_cols) > 0) {
        broad <- sub(":.*$", "", colnames(phi_scores[, cell_type_cols, drop = FALSE])[
            max.col(phi_scores[, cell_type_cols, drop = FALSE], ties.method = "first")
        ])
    }
    pc_df$broad <- broad

    p_stage <- ggplot(pc_df, aes(PC1, PC2, colour = group)) +
        geom_point(size = 0.25, alpha = 0.65) +
        scale_colour_brewer(palette = "Set1") +
        labs(title = "Phenotype-space PCA by donor stage", colour = NULL) +
        theme_void(base_size = 15) +
        theme(plot.title = element_text(face = "bold", size = 14),
              legend.position = "bottom")

    cm_idx <- broad == "Cardiomyocyte"
    if (any(cm_idx) && length(age_cols) > 0) {
        cm_scores <- as.data.frame(phi_scores[cm_idx, age_cols, drop = FALSE])
        cm_scores$group <- query_cd$group[cm_idx]
        cm_long <- stats::reshape(
            cm_scores,
            varying = age_cols,
            v.names = "score",
            timevar = "age_score",
            times = age_cols,
            direction = "long"
        )
        p_age <- ggplot(cm_long, aes(age_score, score, fill = group)) +
            geom_violin(scale = "width", alpha = 0.75, colour = NA) +
            scale_fill_brewer(palette = "Set1") +
            labs(title = "CM age scores preserve the maturation gradient",
                 x = NULL, y = "Φ-Space score", fill = NULL) +
            theme_minimal(base_size = 15) +
            theme(plot.title = element_text(face = "bold", size = 14),
                  legend.position = "bottom")
    } else {
        p_age <- placeholder_plot("CM age-score figure", "Requires age-group Φ-Space columns")
    }

    combine2(p_stage, p_age)
} else {
    placeholder_plot(
        "Φ-Space biplot / score figure",
        "Render after results/05_phispace_query.rds is available"
    )
}
```

<div class="caption">Use this slide to show that continuous phenotype scores retain developmental signal inside a cell type.</div>

???
Timing: 3 minutes. Explain how to read the stage separation without over-interpreting PCA axes.

---

# Bridge: scores become a maturation coordinate

<div class="cols middle">
<div>
<div class="kicker">The payoff of Module 5</div>

* The age scores give a biological anchor for direction.
* `pt_phi_score` is the trajectory-free baseline: adult-like score minus fetal score.
* Module 6 compares fitted pseudotimes against this simple maturation signal.
</div>

<div>
```{r phispace-pt-umap, fig.width=5.6, fig.height=4.2}
cm <- read_cache("results/06_cm_pseudotime.rds")
if (!is.null(cm) && has_ggplot2 && has_sce && has_se &&
    "UMAP" %in% SingleCellExperiment::reducedDimNames(cm)) {
    umap <- as.data.frame(SingleCellExperiment::reducedDim(cm, "UMAP"))
    names(umap) <- c("UMAP1", "UMAP2")
    umap$pt_phi_score <- SummarizedExperiment::colData(cm)$pt_phi_score
    print(plot_umap(umap, "pt_phi_score", "Trajectory-free Φ-Space maturation score"))
} else {
    placeholder_plot("UMAP by pt_phi_score", "Requires results/06_cm_pseudotime.rds")
}
```
</div>
</div>

???
Timing: 2 minutes. This is the hinge between Modules 5 and 6.

---

# Module 6: pseudotime is an ordering, not a clock

<div class="cols middle">
<div>
<div class="claim">Pseudotime assigns each cell a scalar position along a biological process.</div>

* It is not real elapsed time.
* Its absolute scale is not identified.
* Direction must come from outside the method unless velocity is available.
</div>

<div class="panel">
<h3>In this workshop</h3>
<p>The highest fetal Φ-Space score anchors the early end of the cardiomyocyte maturation trajectory.</p>
</div>
</div>

???
Timing: 2 minutes. Make sure participants do not treat pseudotime values as calendar time.

---

# Two method families, two kinds of uncertainty

<div class="cols-3">
<div class="panel">
<h3>Slingshot</h3>
<p>Cluster centroids -> minimum spanning tree -> principal curves -> arc-length pseudotime.</p>
<p><strong>Main choice:</strong> embedding and clustering.</p>
</div>

<div class="panel">
<h3>DPT via destiny</h3>
<p>Cell-cell affinity graph -> diffusion components -> transition distance from a root.</p>
<p><strong>Main choice:</strong> embedding and root cell.</p>
</div>

<div class="panel">
<h3>Velocity</h3>
<p>Splicing dynamics estimate local time derivatives.</p>
<p><strong>Not run:</strong> needs unspliced counts and has snRNA-seq caveats.</p>
</div>
</div>

???
Timing: 3 minutes. Mention velocity only as context; do not let it take over the talk.

---

# Slingshot sensitivity is the lesson

<div class="cols middle">
<div>
<div class="claim">For principal-curve methods, clustering is a modelling choice.</div>

* Slingshot builds its tree on cluster centroids.
* Changing the clustering can move the trajectory even with the same cells.
* Module 6 therefore crosses two embeddings with two clusterings.
</div>

<div class="panel">
<h3>Four slingshot runs</h3>
<p>Φ-Space or PCA embedding, crossed with Louvain or PhiCellType clustering.</p>
</div>
</div>

???
Timing: 3 minutes. This prepares participants to read the sensitivity grid in the vignette.

---

# Module 6 proof object

```{r pseudotime-proof, fig.width=10.8, fig.height=5.6}
cm <- read_cache("results/06_cm_pseudotime.rds")

if (!is.null(cm) && has_ggplot2 && has_sce && has_se &&
    "UMAP" %in% SingleCellExperiment::reducedDimNames(cm)) {
    umap_df <- as.data.frame(SingleCellExperiment::reducedDim(cm, "UMAP"))
    names(umap_df) <- c("UMAP1", "UMAP2")
    cd <- as.data.frame(SummarizedExperiment::colData(cm))

    needed <- c("pt_sling_phi_phict", "pt_sling_pca_phict", "pt_dpt_phi", "pt_dpt_pca")
    if (all(needed %in% names(cd))) {
        for (nm in needed) umap_df[[paste0(nm, "_01")]] <- rescale01(cd[[nm]])
        p1 <- plot_umap(umap_df, "pt_sling_phi_phict_01", "slingshot / Φ-Space")
        p2 <- plot_umap(umap_df, "pt_sling_pca_phict_01", "slingshot / PCA")
        p3 <- plot_umap(umap_df, "pt_dpt_phi_01", "DPT / Φ-Space")
        p4 <- plot_umap(umap_df, "pt_dpt_pca_01", "DPT / PCA")
        combine4(p1, p2, p3, p4)
    } else {
        placeholder_plot("Canonical four-way pseudotime figure",
                         "Pseudotime columns are missing from the Module 6 cache")
    }
} else {
    placeholder_plot(
        "Canonical four-way pseudotime figure",
        "Render after results/06_cm_pseudotime.rds is available"
    )
}
```

<div class="caption">Rows separate method effects; columns separate embedding effects.</div>

???
Timing: 4 minutes. The ideal result is broad agreement; disagreement is also useful because it exposes method sensitivity.

---

# Bridge: from gene levels to gene relationships

<div class="cols middle">
<div>
<div class="kicker">After pseudotime</div>

* Differential expression asks: **which genes go up or down?**
* Network analysis asks: **which gene relationships change?**
* Along a continuum, one global network can average away rewiring.
</div>

<div class="panel">
<h3>Module 7 question</h3>
<p>As cardiomyocytes mature, do transcription factors and target genes change how they co-vary locally?</p>
</div>
</div>

???
Timing: 2 minutes. This is the transition into NeighbourNet.

---

# Module 7: co-expression scaffold, not causal GRN

<div class="cols middle">
<div>
<div class="claim">NeighbourNet builds cell-specific co-expression networks.</div>

* Edges are local TF-target co-expression evidence.
* Prior knowledge can arrange receptor, TF, and target layers.
* The direction in the plot is biological context, not causality inferred from expression alone.
</div>

<div class="panel">
<h3>Interpretation rule</h3>
<p>Use these networks as structured hypothesis generators. Causal claims need perturbation or external evidence.</p>
</div>
</div>

???
Timing: 3 minutes. This slide prevents over-claiming.

---

# NeighbourNet algorithm at high level

<div class="flow">
  <div class="flow-step">PCA<br>denoise cells</div>
  <div class="arrow">-></div>
  <div class="flow-step">kNN<br>local samples</div>
  <div class="arrow">-></div>
  <div class="flow-step">Local PC<br>regression</div>
  <div class="arrow">-></div>
  <div class="flow-step">Cell-specific<br>networks</div>
  <div class="arrow">-></div>
  <div class="flow-step">nPCA<br>meta-networks</div>
</div>

<div class="cols middle" style="margin-top: 40px;">
<div class="panel">
<h3>Computational shape</h3>
<p><code>cells × target genes × TF predictors</code></p>
</div>
<div class="panel">
<h3>Workshop constraint</h3>
<p>Keep 10 maturation-associated target genes so the vignette remains feasible.</p>
</div>
</div>

???
Timing: 4 minutes. Keep this conceptual; participants will see the function-level workflow in the vignette.

---

# Module 7 proof object

```{r nnet-proof, fig.width=10.5, fig.height=5.2, eval=params$render_heavy_nnet}
seu <- read_cache("results/07_seu_nnet_10targets.rds")
if (!is.null(seu) && has_ggplot2 && ok_pkg("Seurat")) {
    nnet_mod <- Seurat::Misc(seu, "NNet.mod")
    meta_pcs <- nnet_mod$meta.network$pcs
    plot_df <- as.data.frame(meta_pcs[, seq_len(min(2, ncol(meta_pcs))), drop = FALSE])
    names(plot_df) <- paste0("NNet", seq_len(ncol(plot_df)))
    plot_df$group <- seu$group
    plot_df$pt_phi_score <- seu$pt_phi_score

    p_stage <- ggplot(plot_df, aes(NNet1, NNet2, colour = group)) +
        geom_point(size = 0.45, alpha = 0.75) +
        scale_colour_brewer(palette = "Set1") +
        labs(title = "Meta-network space by donor stage", colour = NULL) +
        theme_minimal(base_size = 15) +
        theme(plot.title = element_text(face = "bold", size = 14),
              legend.position = "bottom")

    p_pt <- ggplot(plot_df, aes(NNet1, NNet2, colour = pt_phi_score)) +
        geom_point(size = 0.45, alpha = 0.75) +
        scale_colour_distiller(palette = "Spectral", direction = -1) +
        labs(title = "Meta-network space by Φ-Space maturation", colour = NULL) +
        theme_minimal(base_size = 15) +
        theme(plot.title = element_text(face = "bold", size = 14),
              legend.position = "bottom")

    combine2(p_stage, p_pt)
} else {
    placeholder_plot("NeighbourNet meta-network figure",
                     "Requires results/07_seu_nnet_10targets.rds and Seurat")
}
```

```{r nnet-placeholder, fig.width=10.5, fig.height=5.2, eval=!params$render_heavy_nnet}
placeholder_plot(
    "NeighbourNet figure placeholder",
    "Set params$render_heavy_nnet = TRUE to render from results/07_seu_nnet_10targets.rds"
)
```

<div class="caption">Recommended final figure: meta-network scores plus an adjacency heatmap or radial network plot from Module 7.</div>

???
Timing: 4 minutes. Explain meta-network scores as PCA-like cell scores for recurring network patterns.

---

class: section-slide, middle

<div class="kicker">Take-home</div>

<div class="claim">The afternoon workflow keeps continuous biology continuous long enough to ask better questions.</div>

<div class="flow">
  <div class="flow-step">Soft phenotype scores<br>preserve gradients</div>
  <div class="arrow">-></div>
  <div class="flow-step">Pseudotime<br>tests trajectory robustness</div>
  <div class="arrow">-></div>
  <div class="flow-step">Meta-networks<br>summarise rewiring hypotheses</div>
</div>

???
Timing: 2 minutes. End by directing participants back to the three vignettes.
