---
title: "Module 0: Environment Setup"
author: "Single Cell Workshop"
date: "`r Sys.Date()`"
output:
  html_document:
    theme: flatly
    toc: true
    toc_depth: 2
    toc_float: true
    number_sections: true
    code_folding: show
    df_print: paged
vignette: >
  %\VignetteIndexEntry{Module 0: Environment Setup}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, eval = FALSE)
```

## Introduction

Welcome to the Single Cell RNA-Sequencing Workshop. This document guides you through setting up your R environment **before the workshop begins**.

**Please complete this setup at least one day before the workshop.** Package installation takes approximately 10-15 minutes, and data download takes approximately 5 minutes depending on your internet connection.

### What We Will Set Up

1. **R packages** using `renv` for reproducibility
2. **Workshop data** downloaded from Zenodo (~420 MB)

### System Requirements

| Resource | Minimum | Recommended |
|----------|---------|-------------|
| RAM | 8 GB | 16 GB |
| Disk space | 5 GB free | 10 GB free |
| R version | 4.3+ | 4.5+ |
| RStudio | 2023.06+ | Latest |

Check your R version:

```{r check-version, eval=TRUE}
R.version.string
```

If your R version is older than 4.3, please update from [CRAN](https://cran.r-project.org/) before proceeding.

## Step 1: Clone or Download the Repository

First, obtain the workshop materials:

**Option A: Clone with Git (Recommended)**

```bash
git clone https://github.com/phipsonlab/single_cell_workshop.git
cd single_cell_workshop
```

**Option B: Download ZIP**

1. Visit https://github.com/phipsonlab/single_cell_workshop
2. Click the green "Code" button
3. Select "Download ZIP"
4. Extract the ZIP file

Then open `single_cell_workshop.Rproj` in RStudio.

## Step 2: Install Packages with renv

This workshop uses **renv** for reproducible package management. All package versions are locked in `renv.lock`, ensuring everyone has identical environments.

When you open the project in RStudio, renv should automatically bootstrap itself. If prompted to install renv, select "Yes".

Then run:

```{r renv-restore}
# Install all packages with exact versions from renv.lock
# This may take 10-15 minutes on first run
renv::restore()
```

When prompted "Do you want to proceed?", type `y` and press Enter.

### What renv Does

- Installs packages to a project-specific library (not your global R library)
- Uses exact package versions from `renv.lock`
- Ensures reproducibility across different machines

### Key Packages Installed

| Package | Version | Purpose |
|---------|---------|---------|
| Seurat | 5.4.0 | Core single cell analysis |
| harmony | 1.2.4 | Batch correction |
| glmGamPoi | 1.22.0 | Fast SCTransform (critical) |
| edgeR | 4.8.2 | Differential expression |
| limma | 3.66.0 | Linear models |
| speckle | 1.10.0 | Composition analysis |

## Step 3: Verify Package Installation

After `renv::restore()` completes, verify that all critical packages are installed:

```{r verify-packages, eval=TRUE}
packages <- c("Seurat", "harmony", "glmGamPoi", "edgeR", "limma", "speckle")

check <- sapply(packages, requireNamespace, quietly = TRUE)
missing <- names(check)[!check]

if (length(missing) == 0) {
    message("All critical packages installed successfully!")
} else {
    message("WARNING: Missing packages: ", paste(missing, collapse = ", "))
    message("Try running renv::restore() again.")
}
```

## Step 4: Download Workshop Data

The workshop data files are hosted on Zenodo. Run the following code to download them:

```{r download-data}
# Zenodo record for workshop data
zenodo_url <- "https://zenodo.org/records/18237749/files/"
files <- c("heart-counts.Rds", "cellinfo_updated.Rds")

# Create data directory if needed
if (!dir.exists("data")) dir.create("data")

# Download each file
for (f in files) {
    dest <- file.path("data", f)
    if (file.exists(dest)) {
        message(f, " already exists, skipping")
        next
    }
    message("Downloading ", f, "...")
    download.file(
        url = paste0(zenodo_url, f, "?download=1"),
        destfile = dest,
        mode = "wb"
    )
}
message("Download complete!")
```

**Note:** Total download size is approximately 420 MB.

## Step 5: Final Verification

Run this final check to ensure everything is ready:

```{r final-check, eval=TRUE}
cat("=== Workshop Setup Verification ===\n\n")

# Check R version
cat("R Version:", R.version.string, "\n\n")

# Check critical packages
cat("Package Status:\n")
packages <- c("Seurat", "harmony", "glmGamPoi", "edgeR", "limma", "speckle")
for (pkg in packages) {
    status <- if (requireNamespace(pkg, quietly = TRUE)) "OK" else "MISSING"
    cat(sprintf("  %-12s %s\n", pkg, status))
}

# Check data files
cat("\nData Files:\n")
data_dir <- if (dir.exists("data")) "data" else "../data"
for (f in c("heart-counts.Rds", "cellinfo_updated.Rds")) {
    path <- file.path(data_dir, f)
    status <- if (file.exists(path)) "FOUND" else "NOT FOUND"
    cat(sprintf("  %-25s %s\n", f, status))
}

cat("\n================================\n")
```

## Troubleshooting

### renv::restore() fails

If `renv::restore()` encounters errors:

1. **Restart R** (Session > Restart R)
2. Run `renv::restore()` again
3. If specific packages fail, try installing them manually then run `renv::restore()`

### Memory errors

If you encounter "cannot allocate vector" errors:

1. Close other applications
2. Restart R
3. Try again

### macOS compilation errors

Install Xcode Command Line Tools:

```bash
xcode-select --install
```

### Package conflicts

If you have existing packages causing conflicts:

```r
# Use a clean renv library
renv::rebuild()
```

## Getting Help

If you cannot resolve setup issues:

1. Note the exact error message
2. Run `sessionInfo()` and save the output
3. Contact the workshop organisers before the session

## Session Information

```{r session-info, eval=TRUE}
sessionInfo()
```
