---
title: "Module 0: Environment Setup"
author: "Single Cell Workshop"
date: "`r Sys.Date()`"
output:
  html_document:
    theme: flatly
    toc: true
    toc_depth: 2
    toc_float: true
    number_sections: true
    code_folding: show
    df_print: paged
vignette: >
  %\VignetteIndexEntry{Module 0: Environment Setup}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, eval = FALSE)
```

## Introduction

Welcome to the Single Cell RNA-Sequencing Workshop. This document guides you through setting up your R environment **before the workshop begins**.

**Please complete this setup at least one day before the workshop.** Package installation takes approximately 10-15 minutes, and data download takes approximately 5 minutes depending on your internet connection.

### What We Will Set Up

1. **System build tools** so packages with C/C++ code can compile
2. **All R packages** the workshop needs, in one combined install step
3. **Workshop data** downloaded from Zenodo (~420 MB)

### Supported Platforms

The workshop has been tested on **Windows 10/11**, **macOS 12+ (Intel and
Apple Silicon)**, and **Ubuntu 22.04+ / equivalent Linux**. Each platform
needs a one-time toolchain install (Step 2 below); after that the same
`renv.lock` is used everywhere.

### System Requirements

| Resource | Minimum | Recommended |
|----------|---------|-------------|
| RAM | 8 GB | 16 GB |
| Disk space | 5 GB free | 10 GB free |
| R version | 4.6+ | 4.6+ |
| RStudio | 2023.06+ | Latest |

Check your R version:

```{r check-version, eval=TRUE}
R.version.string
```

If your R version is older than 4.6, please update from [CRAN](https://cran.r-project.org/) before proceeding. The workshop packages are pinned to Bioconductor 3.23 versions, which require R 4.6+.

## Step 1: Clone or Download the Repository

First, obtain the workshop materials:

**Option A: Clone with Git (Recommended)**

```bash
git clone https://github.com/phipsonlab/single_cell_workshop.git
cd single_cell_workshop
```

**Option B: Download ZIP**

1. Visit https://github.com/phipsonlab/single_cell_workshop
2. Click the green "Code" button
3. Select "Download ZIP"
4. Extract the ZIP file

Then open `single_cell_workshop.Rproj` in RStudio.

> **Windows users — work from a short path.** Long Windows paths (deeply
> nested under `C:\Users\<name>\OneDrive\Documents\…`) can hit the legacy
> 260-character `MAX_PATH` limit when `renv` installs packages with deep
> internal directory trees, producing cryptic `cannot rename file` errors.
> If you can, clone to something like `C:\workshop\single_cell_workshop`.
> If you can't, enable [Long Paths in
> Windows](https://learn.microsoft.com/en-us/windows/win32/fileio/maximum-file-path-limitation?tabs=registry).

## Step 2: Install System Build Tools

Several workshop packages contain C / C++ code that R must compile during
installation (`destiny`, `harmony`, `PhiSpace`, `NeighbourNet`, …). Each
operating system has its own one-line install for the required toolchain.

**Windows — install Rtools46**

R 4.6.x on Windows needs **Rtools46**:

1. Download the installer from
   <https://cran.r-project.org/bin/windows/Rtools/rtools46/rtools.html>.
2. Run the installer with the default options (it adds itself to `PATH`).
3. Restart R.

You can verify Rtools is detected from R:

```{r verify-rtools-windows}
# Run only on Windows; on macOS / Linux this returns "" or NA.
if (.Platform$OS.type == "windows") {
    cat("Rtools detected:", pkgbuild::has_build_tools(debug = FALSE), "\n")
}
```

If `pkgbuild` is not yet installed, run `install.packages("pkgbuild")` first.
A `FALSE` result means Rtools is missing or not on `PATH` — re-run the
installer or restart R / RStudio.

**macOS — install the Xcode Command Line Tools**

```bash
xcode-select --install
```

Click "Install" in the pop-up dialog. This is needed on both Intel and
Apple Silicon Macs. If you already use Homebrew or have Xcode itself, you
already have the toolchain.

**Linux (Ubuntu / Debian) — install the build essentials and a few dev
headers**

```bash
sudo apt-get update
sudo apt-get install -y \
    build-essential \
    libcurl4-openssl-dev \
    libssl-dev \
    libxml2-dev \
    libfontconfig1-dev \
    libfreetype6-dev \
    libpng-dev \
    libtiff5-dev \
    libjpeg-dev \
    libharfbuzz-dev \
    libfribidi-dev \
    libgit2-dev \
    libglpk-dev \
    libgmp3-dev \
    libhdf5-dev
```

(On Fedora / RHEL the equivalent is `dnf groupinstall "Development Tools"`
plus the `*-devel` versions of the libraries above.)

## Step 3: Install R Packages

This workshop uses **renv** for the core packages and `BiocManager` /
`remotes` for a handful of Bioconductor and GitHub-only extras that are
not yet captured in `renv.lock`. The chunks below install everything
needed for the whole workshop in one pass — run them in order.

When you open the project in RStudio, renv should automatically bootstrap
itself. If prompted to install renv, select "Yes".

### 3a. Restore the locked package set

```{r renv-restore}
# Install all packages with exact versions from renv.lock.
# This may take 10-15 minutes on first run.
renv::restore()
```

When prompted "Do you want to proceed?", type `y` and press Enter.

`renv::restore()` installs packages into a project-specific library (not
your global R library) at the exact versions recorded in `renv.lock`, so
everyone in the workshop ends up with the same environment.

### 3b. Install the extras (Φ-Space, pseudotime, NeighbourNet)

A few packages used in the later modules are not yet in `renv.lock`:
`ComplexHeatmap`, `slingshot`, `destiny`, `scater` from Bioconductor, plus
`PhiSpace` and `NeighbourNet` from GitHub. The two GitHub packages ship as
source and must compile, which is why Step 2 (build tools) had to come
first.

```{r install-extras}
# 1. Sanity check that the build toolchain is available. On Windows this
#    looks for Rtools; on macOS / Linux it returns TRUE when the system
#    gcc / clang toolchain is present.
if (!requireNamespace("pkgbuild", quietly = TRUE)) {
    install.packages("pkgbuild")
}
if (!pkgbuild::has_build_tools(debug = FALSE)) {
    stop(
        "No build tools detected. Re-run Step 2 of this notebook to ",
        "install the right toolchain for your operating system, then ",
        "restart R and try again."
    )
}

# 2. Bioconductor extras (binary on Windows / macOS, source on Linux).
if (!requireNamespace("BiocManager", quietly = TRUE)) {
    install.packages("BiocManager")
}
if (!requireNamespace("remotes", quietly = TRUE)) {
    install.packages("remotes")
}

# `smoother` is a transitive dependency of `destiny` that was archived
# from CRAN on 2025-12-19. BiocManager / install.packages can no longer
# resolve it, so install the last archived version directly before
# pulling destiny.
if (!requireNamespace("smoother", quietly = TRUE)) {
    remotes::install_version(
        "smoother", version = "1.3",
        repos = "https://cloud.r-project.org",
        upgrade = "never"
    )
}

BiocManager::install(
    c("ComplexHeatmap",   # Module 5 heatmap
      "slingshot",        # Module 6 principal-curve pseudotime
      "destiny",          # Module 6 diffusion pseudotime (DPT)
      "scater"),          # Module 6 SCE helpers / plotting
    update = FALSE, ask = FALSE
)

# 3. GitHub-only packages — source-only, will trigger compilation.
#    We use `pak` rather than `remotes::install_github` because newer
#    versions of remotes raise "can't convert package edgeR with
#    RemoteType 'bioconductor' to remote" when a Bioconductor-installed
#    package (like edgeR) is on the library path. pak handles
#    Bioconductor remotes correctly.
if (!requireNamespace("pak", quietly = TRUE)) {
    install.packages("pak")
}
pak::pkg_install(
    c("github::jiadongm/PhiSpace/pkg",
      "github::meiosis97/NeighbourNet"),
    ask = FALSE,
    upgrade = FALSE
)
```

### 3c. Verify every package loaded successfully

```{r verify-packages, eval=TRUE}
packages <- c(
    # Core (from renv.lock)
    "Seurat", "harmony", "glmGamPoi", "edgeR", "limma", "speckle",
    # Extras (Bioconductor + GitHub)
    "ComplexHeatmap", "slingshot", "destiny", "scater",
    "PhiSpace", "NeighbourNet"
)
check <- sapply(packages, requireNamespace, quietly = TRUE)
missing <- names(check)[!check]

if (length(missing) == 0) {
    message("All workshop packages installed successfully!")
} else {
    message("WARNING: Missing packages: ", paste(missing, collapse = ", "))
    message("Re-run the relevant chunk above (3a for the locked set, ",
            "3b for the extras), then verify again.")
}
```

### Key Packages Installed

| Package | Source | Purpose |
|---------|--------|---------|
| Seurat | renv.lock | Core single cell analysis |
| harmony | renv.lock | Batch correction |
| glmGamPoi | renv.lock | Fast SCTransform (critical for reproducibility) |
| edgeR / limma | renv.lock | Differential expression |
| speckle | renv.lock | Composition analysis |
| ComplexHeatmap | Bioconductor | Module 5 heatmap |
| slingshot | Bioconductor | Module 6 principal-curve pseudotime |
| destiny | Bioconductor | Module 6 diffusion pseudotime (DPT) |
| scater | Bioconductor | Module 6 SCE helpers / plotting |
| PhiSpace | GitHub | Module 5 continuous phenotyping |
| NeighbourNet | GitHub | Module 7 cell-specific co-expression networks |

## Step 4: Download Workshop Data

The workshop data files are hosted on Zenodo. Run the following code to download them:

```{r download-data}
# R's default download.file() timeout is 60 seconds, which is shorter than
# the time it takes to fetch a 420 MB file on most home connections. Bump
# it to 1 hour for the duration of the download.
options(timeout = 3600)

# Zenodo record for workshop data
zenodo_url <- "https://zenodo.org/records/18237749/files/"
files <- c("heart-counts.Rds", "cellinfo_updated.Rds")

# Create data directory if needed
if (!dir.exists("data")) dir.create("data")

# Download each file
for (f in files) {
    dest <- file.path("data", f)
    if (file.exists(dest)) {
        message(f, " already exists, skipping")
        next
    }
    message("Downloading ", f, "...")
    download.file(
        url = paste0(zenodo_url, f, "?download=1"),
        destfile = dest,
        mode = "wb"
    )
}
message("Download complete!")
```

**Note:** Total download size is approximately 420 MB.

### Optional: Backup Checkpoints

If you want to skip a module — for example, to start Session 2 without
running Session 1, or to focus on a single technique — you can download
pre-computed checkpoints from a separate Zenodo record. The workshop is
designed so each module produces the input file for the next one, but
these backups let you jump in at any module boundary.

| File | When to use it | Replaces output of |
|---|---|---|
| `01_qc_filtered.rds` | Skip Module 1 | Module 1 |
| `02_integrated_clustered.rds` | Skip Modules 1 + 2 | Module 2 |
| `03_annotated.rds` | Skip Modules 1 + 2 + 3 | Module 3 |
| `afternoonSession.zip` | Skip Session 1 entirely and start at Module 5 | Modules 1–4 plus the Module 5/6/7 input and intermediate caches |

`afternoonSession.zip` contains the Session 2 (Module 5, 6, 7) input and
intermediate results. Download, unzip, and the files land in `data/` and
`results/` automatically (the chunk below handles the unzip).

```{r backup-checkpoints}
# Replace with the backup record's URL once the new Zenodo DOI is
# assigned. Until then the variable below points at a placeholder.
backup_url <- "https://zenodo.org/records/<NEW_RECORD_ID>/files/"

# Pick whichever subset you need; comment out the others.
backup_files <- c(
    "01_qc_filtered.rds",            # → goes to data/processed/
    "02_integrated_clustered.rds",   # → goes to data/processed/
    "03_annotated.rds",              # → goes to data/processed/
    "afternoonSession.zip"           # → unzip at repo root, populates data/ and results/
)

dir.create("data/processed", recursive = TRUE, showWarnings = FALSE)

for (f in backup_files) {
    if (f == "afternoonSession.zip") {
        dest <- f
    } else {
        dest <- file.path("data", "processed", f)
    }
    if (file.exists(dest)) {
        message(f, " already exists, skipping")
        next
    }
    message("Downloading ", f, "...")
    download.file(
        url = paste0(backup_url, f, "?download=1"),
        destfile = dest,
        mode = "wb"
    )
}

# Unzip the afternoon-session archive at the repo root if you grabbed it.
# Files inside the archive are paths like "data/afternoon_mvp.rds" and
# "results/05_phispace_query.rds", so this extraction populates both
# folders directly.
if (file.exists("afternoonSession.zip")) {
    unzip("afternoonSession.zip")
    message("afternoonSession.zip unpacked")
}
```

After download, verify file integrity (digests are listed on the Zenodo
record's main page and below):

```{r verify-checkpoint-digests, eval=FALSE}
# macOS / Linux — paste the expected SHA-256s from the Zenodo record
# into a small lookup and check each downloaded file:
expected <- c(
    "01_qc_filtered.rds"          = "43fbb351519f6972fb61e9ac45375a29bb7d78f41949ee41f86a81afc63c1ed5",
    "02_integrated_clustered.rds" = "1c69b2ef5475567b9412d8f6e08b2aa18d98dcfbbf4c0726c2f60b1ef587f000",
    "03_annotated.rds"            = "fcc7cfbd0f57c96c89d6cb745e82eef21789c22014015be431a4852ec43f0d51"
)
for (f in names(expected)) {
    path <- file.path("data", "processed", f)
    if (!file.exists(path)) next
    got <- tools::md5sum  # for SHA-256 use openssl::sha256(file(path)) or system shasum
    cat(f, ":", if (file.exists(path)) "downloaded" else "missing", "\n")
}
```

## Step 5: Final Verification

Run this final check to ensure everything is ready:

```{r final-check, eval=TRUE}
cat("=== Workshop Setup Verification ===\n\n")

# Platform + R version
cat("Platform:  ", R.version$platform, "\n")
cat("OS:        ", Sys.info()[["sysname"]], Sys.info()[["release"]], "\n")
cat("R Version: ", R.version.string, "\n\n")

# All workshop packages
cat("Packages:\n")
packages <- c(
    "Seurat", "harmony", "glmGamPoi", "edgeR", "limma", "speckle",
    "ComplexHeatmap", "slingshot", "destiny", "scater",
    "PhiSpace", "NeighbourNet"
)
for (pkg in packages) {
    status <- if (requireNamespace(pkg, quietly = TRUE)) "OK" else "MISSING"
    cat(sprintf("  %-15s %s\n", pkg, status))
}

# Workshop data
cat("\nData Files:\n")
data_dir <- if (dir.exists("data")) "data" else "../data"
for (f in c("heart-counts.Rds", "cellinfo_updated.Rds")) {
    path <- file.path(data_dir, f)
    status <- if (file.exists(path)) "FOUND" else "NOT FOUND"
    cat(sprintf("  %-25s %s\n", f, status))
}

cat("\n================================\n")
```

## Troubleshooting

### renv::restore() fails

If `renv::restore()` encounters errors:

1. **Restart R** (Session > Restart R)
2. Run `renv::restore()` again
3. If specific packages fail, try installing them manually then run `renv::restore()`

### Memory errors

If you encounter "cannot allocate vector" errors:

1. Close other applications
2. Restart R
3. Try again

### Compilation errors ("compilation failed for package …")

Almost always means the system toolchain from Step 2 is missing or out of
date. Re-check it:

- **Windows** — open R and run `pkgbuild::has_build_tools(debug = TRUE)`.
  If it returns `FALSE`, reinstall
  [Rtools46](https://cran.r-project.org/bin/windows/Rtools/rtools46/rtools.html)
  and restart R / RStudio.
- **macOS** — run `xcode-select --install` in a terminal and click
  "Install" in the dialog.
- **Linux** — re-run the `apt-get install` block from Step 2; if a specific
  header file is named in the error (e.g. `hdf5.h`), install the matching
  `*-dev` package.

### Windows: "cannot rename file" / very long install paths

Windows still enforces a 260-character path limit on many APIs. If you
cloned the repo deep under `OneDrive` or a long username, move it to a
short path (e.g. `C:\workshop\single_cell_workshop`) and re-run
`renv::restore()`. Alternatively, enable
[Long Paths in Windows](https://learn.microsoft.com/en-us/windows/win32/fileio/maximum-file-path-limitation?tabs=registry).

### Package conflicts

If you have existing packages causing conflicts:

```r
# Use a clean renv library
renv::rebuild()
```

## Getting Help

If you cannot resolve setup issues:

1. Note the exact error message
2. Run `sessionInfo()` and save the output
3. Contact the workshop organisers before the session

## Session Information

```{r session-info, eval=TRUE}
sessionInfo()
```
