---
title: "Module 0: Environment Setup"
author: "Single Cell Workshop"
date: "`r Sys.Date()`"
output:
  html_document:
    theme: flatly
    toc: true
    toc_depth: 2
    toc_float: true
    number_sections: true
    code_folding: show
    df_print: paged
vignette: >
  %\VignetteIndexEntry{Module 0: Environment Setup}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, eval = FALSE)
```

## Introduction

Welcome to the Single Cell RNA-Sequencing Workshop. This document guides you
through setting up your R environment **before the workshop begins**.

**Please complete this setup at least one day before the workshop.** The data
download takes approximately 5 minutes depending on your internet connection,
and package installation may take an additional 10-15 minutes.

### What We Will Set Up

1. **R packages** using `renv` for reproducibility
2. **Workshop data** downloaded from Zenodo (~420 MB)

### System Requirements

| Resource | Minimum | Recommended |
|----------|---------|-------------|
| RAM | 8 GB | 16 GB |
| Disk space | 5 GB free | 10 GB free |
| R version | 4.3+ | 4.5+ |
| RStudio | 2023.06+ | Latest |

Check your R version:

```{r check-version, eval=TRUE}
R.version.string
```

If your R version is older than 4.3, please update from
[CRAN](https://cran.r-project.org/) before proceeding.

---

## Step 1: Clone or Download the Workshop Repository

First, obtain the workshop materials. You have two options:

### Option A: Clone with Git (Recommended)

If you have Git installed, open a terminal and run:
```bash
git clone https://github.com/phipsonlab/single_cell_workshop.git
cd single_cell_workshop
```

### Option B: Download ZIP

1. Go to https://github.com/phipsonlab/single_cell_workshop
2. Click the green "Code" button
3. Select "Download ZIP"
4. Extract the ZIP file to your desired location

### Open the Project in RStudio

1. Open RStudio
2. Go to **File > Open Project**
3. Navigate to the workshop folder and open `single_cell_workshop.Rproj`

You should see the project name in the top-right corner of RStudio.

---

## Step 2: Install Packages with renv

This workshop uses **renv** to ensure all participants have identical package
versions. This guarantees reproducibility - your results will match exactly
what we demonstrate.

### What is renv?

`renv` is a package manager that creates isolated, reproducible R environments.
The workshop repository includes an `renv.lock` file that specifies the exact
versions of all required packages.

### Install Packages

When you first open the project, renv should automatically bootstrap itself.
If prompted to install renv, type `y` and press Enter.

Then run the following to install all required packages:

```{r install-renv, eval=FALSE}
# This will install all packages with the exact versions used in the workshop
# Installation typically takes 10-15 minutes
renv::restore()
```

When prompted "Do you want to proceed?", type `y` and press Enter.

**Note:** If you encounter errors, see the Troubleshooting section below.

### Verify Installation

After installation completes, verify that key packages load correctly:

```{r verify-packages, eval=TRUE}
# List of essential packages
packages <- c(
    "Seurat",
    "harmony",
    "edgeR",
    "limma",
    "speckle",
    "glmGamPoi",
    "org.Hs.eg.db",
    "ggplot2",
    "patchwork",
    "dplyr",
    "clustree"
)

# Check each package
results <- sapply(packages, requireNamespace, quietly = TRUE)
missing <- names(results)[!results]

if (length(missing) == 0) {
    message("All packages installed successfully!")
} else {
    message("Missing packages: ", paste(missing, collapse = ", "))
    message("Please run renv::restore() again or see Troubleshooting below.")
}
```

### Check Package Versions

For reference, here are the key package versions used in this workshop:

```{r check-versions, eval=TRUE}
key_packages <- c("Seurat", "SeuratObject", "harmony", "glmGamPoi",
                  "edgeR", "limma", "speckle")

cat("Package versions:\n")
cat(paste(rep("-", 40), collapse = ""), "\n")
for (pkg in key_packages) {
    if (requireNamespace(pkg, quietly = TRUE)) {
        cat(sprintf("%-15s %s\n", pkg, as.character(packageVersion(pkg))))
    }
}
```

---
## Step 3: Download Workshop Data from Zenodo

The workshop uses single-nucleus RNA-seq data from human heart tissue. The data
files are hosted on Zenodo and must be downloaded before the workshop.

### About the Data

| File | Size | Contents |
|------|------|----------|
| `heart-counts.Rds` | ~420 MB | Gene expression matrix (33,939 genes x 54,140 cells) |
| `cellinfo_updated.Rds` | ~400 KB | Cell metadata (sample, group, cell type) |

**Expected download time: ~5 minutes** (depending on internet speed)

The data comes from: Sim et al. (2021) "Sex-Specific Control of Human Heart
Maturation by the Progesterone Receptor", *Circulation*.
[DOI: 10.1161/CIRCULATIONAHA.120.051921](https://doi.org/10.1161/CIRCULATIONAHA.120.051921)

### Download the Data

Run the following code to download both files:

```{r download-data, eval=TRUE}
# Create data directory
if (!dir.exists("data")) {
    dir.create("data")
}

# Zenodo record URL
zenodo_base <- "https://zenodo.org/records/18237749/files/"

# Download heart-counts.Rds (~420 MB, takes ~5 minutes)
message("Downloading heart-counts.Rds (~420 MB)...")
message("This will take approximately 5 minutes.")
message("Started at: ", Sys.time())

download.file(
    url = paste0(zenodo_base, "heart-counts.Rds?download=1"),
    destfile = "data/heart-counts.Rds",
    mode = "wb"
)
message("Completed at: ", Sys.time())

# Download cellinfo_updated.Rds (~400 KB, very fast)
message("\nDownloading cellinfo_updated.Rds...")
download.file(
    url = paste0(zenodo_base, "cellinfo_updated.Rds?download=1"),
    destfile = "data/cellinfo_updated.Rds",
    mode = "wb"
)

message("\nDownload complete!")
```

### Verify Download

Check that the files downloaded correctly:

```{r verify-data, eval=TRUE}
data_dir <- "../data"

counts_file <- file.path(data_dir, "heart-counts.Rds")
cellinfo_file <- file.path(data_dir, "cellinfo_updated.Rds")

cat("Checking data files...\n")
cat(paste(rep("-", 50), collapse = ""), "\n")

# Check counts file
if (file.exists(counts_file)) {
    size_mb <- round(file.size(counts_file) / 1024^2, 1)
    cat(sprintf("heart-counts.Rds     : FOUND (%s MB)\n", size_mb))
} else {
    cat("heart-counts.Rds     : NOT FOUND\n")
}

# Check cellinfo file
if (file.exists(cellinfo_file)) {
    size_kb <- round(file.size(cellinfo_file) / 1024, 1)
    cat(sprintf("cellinfo_updated.Rds : FOUND (%s KB)\n", size_kb))
} else {
    cat("cellinfo_updated.Rds : NOT FOUND\n")
}

cat(paste(rep("-", 50), collapse = ""), "\n")

if (file.exists(counts_file) && file.exists(cellinfo_file)) {
    cat("\nData files are ready for the workshop!\n")
} else {
    cat("\nWARNING: Some data files are missing.\n")
    cat("Please run the download code above.\n")
}
```

### Test Data Loading

As a final check, verify the data can be loaded:

```{r test-load, eval=TRUE}
data_dir <- "../data"
counts_file <- file.path(data_dir, "heart-counts.Rds")

if (file.exists(counts_file)) {
    cat("Loading count matrix to verify integrity...\n")
    counts <- readRDS(counts_file)
    cat(sprintf("Matrix dimensions: %d genes x %d cells\n",
                nrow(counts), ncol(counts)))
    cat("Data integrity check: PASSED\n")
    rm(counts)  # Free memory
} else {
    cat("Cannot test - data file not found.\n")
}
```

---

## Final Checklist

Before the workshop, confirm you have completed all steps:

```{r final-check, eval=TRUE}
cat("=== WORKSHOP SETUP CHECKLIST ===\n\n")

# Check 1: R version
r_version <- as.numeric(R.version$major) + as.numeric(R.version$minor) / 10
r_ok <- r_version >= 4.3
cat(sprintf("[%s] R version 4.3+ (you have %s)\n",
            ifelse(r_ok, "OK", "!!"), R.version.string))

# Check 2: Key packages
key_pkgs <- c("Seurat", "harmony", "glmGamPoi", "edgeR", "limma")
pkg_ok <- all(sapply(key_pkgs, requireNamespace, quietly = TRUE))
cat(sprintf("[%s] Essential packages installed\n",
            ifelse(pkg_ok, "OK", "!!")))

# Check 3: Data files
data_dir <- "../data"
counts_exists <- file.exists(file.path(data_dir, "heart-counts.Rds"))
cellinfo_exists <- file.exists(file.path(data_dir, "cellinfo_updated.Rds"))
data_ok <- counts_exists && cellinfo_exists
cat(sprintf("[%s] Workshop data downloaded\n",
            ifelse(data_ok, "OK", "!!")))

cat("\n")
if (r_ok && pkg_ok && data_ok) {
    cat("Your environment is ready for the workshop!\n")
} else {
    cat("Please address the items marked [!!] before the workshop.\n")
}
```

---

## What to Expect in the Workshop

Once your environment is set up, you are ready for the workshop:

### Session 1: Core Single Cell Analysis (Morning, ~3 hours)

| Module | Topic | Duration |
|--------|-------|----------|
| 1 | Quality Control | ~45 min |
| 2 | Normalisation, Integration & Clustering | ~50 min |
| 3 | Cell Type Annotation | ~20 min |
| 4 | Differential Expression & Composition | ~55 min |

### Session 2: Downstream Analysis (Afternoon, ~3 hours)

*Coming soon* - This session will cover additional downstream analyses using
the same heart development dataset.

<!--
TODO: Add Session 2 modules here when ready
| Module | Topic | Duration |
|--------|-------|----------|
| 5 | TBD | ~X min |
| 6 | TBD | ~X min |
-->

We look forward to seeing you at the workshop!

## Session Information

```{r session-info, eval=TRUE}
sessionInfo()
```
