---
title: "Module 0: Environment Setup"
author: "Single Cell Workshop"
date: "`r Sys.Date()`"
output:
  html_document:
    theme: flatly
    toc: true
    toc_depth: 2
    toc_float: true
    number_sections: true
    code_folding: show
    df_print: paged
vignette: >
  %\VignetteIndexEntry{Module 0: Environment Setup}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, eval = FALSE)
```

## Introduction

Welcome to the Single Cell RNA-Sequencing Workshop. This document guides you through setting up your R environment before the workshop begins. Single cell RNA-sequencing (scRNA-seq) has revolutionised our ability to study gene expression at the resolution of individual cells, enabling the discovery of novel cell types and providing insights into the cellular composition of complex tissues.

In this workshop, we will analyse single nuclei RNA-sequencing (snRNA-seq) data from human heart tissue across three developmental stages: foetal, young, and adult. The dataset originates from a study examining sex-specific control of human heart maturation (Sim et al., 2021, *Circulation*).

### Why is Setup Important?

Single cell analysis requires several specialised R packages that work together. Installing these packages correctly before the workshop ensures that:

1. You can follow along with the live coding without delays
2. Package dependencies are resolved correctly
3. Any installation issues can be addressed before the session

**Please complete this setup at least one day before the workshop.** The installation process typically takes 15-30 minutes, depending on your internet connection and system configuration.

## System Requirements

Before proceeding, it is important to verify that your system meets the minimum requirements for running single cell analyses. Single cell datasets can be large, and insufficient resources may cause R to crash or analyses to run very slowly.

| Resource | Minimum | Recommended | Notes |
|----------|---------|-------------|-------|
| RAM | 8 GB | 16 GB | More RAM allows analysis of larger datasets |
| Disk space | 5 GB free | 10 GB free | Needed for packages and intermediate files |
| R version | 4.3+ | 4.4+ | Earlier versions may lack required features |
| RStudio | 2023.06+ | Latest | Provides the integrated development environment |

### Checking Your R Version

To check which version of R you have installed, open RStudio and run the following command in the console:

```{r check-version}
R.version.string
```

If your R version is older than 4.3, we recommend updating R before installing the workshop packages. You can download the latest version from [CRAN](https://cran.r-project.org/).

## Required Package Versions

For reproducibility, this workshop uses specific package versions. The code outputs
in the tutorial notebooks were generated with these exact versions, so using
different versions may produce slightly different results.

**R Environment:**

- R version: **4.5.2**
- Bioconductor: **3.22**

**CRAN Packages:**

| Package | Version | Purpose |
|---------|---------|---------|
| Seurat | 5.4.0 | Core single cell analysis toolkit |
| SeuratObject | 5.3.0 | Data structures for Seurat |
| harmony | 1.2.4 | Batch effect correction |
| ggplot2 | 4.0.1 | Data visualisation |
| patchwork | 1.3.2 | Combining plots |
| dplyr | 1.1.4 | Data manipulation |
| tidyr | 1.3.2 | Data tidying |
| RColorBrewer | 1.1.3 | Colour palettes |
| clustree | 0.5.1 | Cluster resolution visualisation |
| pheatmap | 1.0.13 | Heatmaps |

**Bioconductor Packages:**

| Package | Version | Purpose |
|---------|---------|---------|
| edgeR | 4.8.2 | Differential expression |
| limma | 3.66.0 | Linear models for DE |
| speckle | 1.10.0 | Cell type composition analysis |
| org.Hs.eg.db | 3.22.0 | Human gene annotations |
| AnnotationDbi | 1.72.0 | Annotation database interface |

## Understanding R Package Sources

R packages come from three main sources, each serving different purposes:

1. **CRAN (Comprehensive R Archive Network)**: The official repository for general-purpose R packages. Packages here undergo basic quality checks and are easy to install using `install.packages()`.

2. **Bioconductor**: A specialised repository for bioinformatics packages. Bioconductor packages follow stricter development guidelines and are designed to work together for genomic data analysis. We install these using `BiocManager::install()`.

3. **GitHub**: A code hosting platform where developers share packages that may be in active development or not yet submitted to CRAN/Bioconductor. We install these using `remotes::install_github()`.

We install packages in a specific order (CRAN → Bioconductor → GitHub) to ensure that dependencies are resolved correctly.

## Automatic Package Installation (Recommended)

The following code chunk automatically detects which packages are missing from your system and installs them. This is the easiest way to set up your environment - simply run this chunk and wait for it to complete.

```{r auto-install, eval=FALSE}
# =============================================================================
# Automatic Package Installation with Version Pinning
# This chunk installs packages with specific versions for reproducibility
# =============================================================================

# --------------------------------------------------------------------------
# Step 1: Install remotes and BiocManager
# --------------------------------------------------------------------------
if (!requireNamespace("remotes", quietly = TRUE)) {
    install.packages("remotes")
}
if (!requireNamespace("BiocManager", quietly = TRUE)) {
    install.packages("BiocManager")
}

# Set Bioconductor version to 3.22
BiocManager::install(version = "3.22", ask = FALSE, update = FALSE)

# --------------------------------------------------------------------------
# Step 2: Define required packages with versions
# --------------------------------------------------------------------------
cran_packages <- list(
    Seurat = "5.4.0",
    SeuratObject = "5.3.0",
    harmony = "1.2.4",
    ggplot2 = "4.0.1",
    patchwork = "1.3.2",
    dplyr = "1.1.4",
    tidyr = "1.3.2",
    RColorBrewer = "1.1.3",
    clustree = "0.5.1",
    pheatmap = "1.0.13"
)

bioc_packages <- c(
    "edgeR",
    "limma",
    "org.Hs.eg.db",
    "AnnotationDbi",
    "speckle"
)

# --------------------------------------------------------------------------
# Step 3: Install Bioconductor packages
# --------------------------------------------------------------------------
message("\n=== Installing Bioconductor packages ===\n")
for (pkg in bioc_packages) {
    if (!requireNamespace(pkg, quietly = TRUE)) {
        message("Installing ", pkg, "...")
        BiocManager::install(pkg, ask = FALSE, update = FALSE)
    } else {
        message(pkg, " already installed")
    }
}

# --------------------------------------------------------------------------
# Step 4: Install CRAN packages with specific versions
# --------------------------------------------------------------------------
message("\n=== Installing CRAN packages with pinned versions ===\n")
for (pkg in names(cran_packages)) {
    version <- cran_packages[[pkg]]
    current <- tryCatch(
        as.character(packageVersion(pkg)),
        error = function(e) ""
    )
    if (current != version) {
        message("Installing ", pkg, " version ", version, "...")
        remotes::install_version(pkg, version = version,
                                 repos = "https://cloud.r-project.org",
                                 upgrade = "never", quiet = TRUE)
    } else {
        message(pkg, " ", version, " already installed")
    }
}

# --------------------------------------------------------------------------
# Final verification
# --------------------------------------------------------------------------
message("\n", paste(rep("=", 50), collapse = ""))
all_packages <- c(names(cran_packages), bioc_packages)
final_check <- sapply(all_packages, requireNamespace, quietly = TRUE)
missing_final <- names(final_check)[!final_check]

if (length(missing_final) == 0) {
    message("SUCCESS! All packages are installed.")
    message("Your environment is ready for the workshop!")
} else {
    message("WARNING: The following packages could not be installed:")
    for (pkg in missing_final) {
        message("  - ", pkg)
    }
    message("\nPlease see the Troubleshooting section below.")
}
message(paste(rep("=", 50), collapse = ""))
```

If you prefer to install packages manually, or if the automatic installation fails for some packages, follow the step-by-step instructions below.

---

## Manual Package Installation

### Step 1: Install BiocManager

*BiocManager* is a package that manages installation of Bioconductor packages. It ensures that you get compatible versions of all Bioconductor packages. If you do not already have BiocManager installed, run the following code:

```{r install-biocmanager}
# Check if BiocManager is installed; if not, install it
if (!require("BiocManager", quietly = TRUE))
    install.packages("BiocManager")
```

The `require()` function attempts to load a package and returns `TRUE` if successful, `FALSE` otherwise. The `quietly = TRUE` argument suppresses messages if the package is not found.

### Step 2: Install Bioconductor Packages

The following Bioconductor packages provide essential functionality for our single cell analysis:

| Package | Purpose |
|---------|---------|
| `edgeR` | Differential expression analysis using negative binomial models |
| `limma` | Linear models for differential expression, works with edgeR |
| `org.Hs.eg.db` | Human gene annotation database |
| `AnnotationDbi` | Interface to annotation databases |
| `speckle` | Statistical methods for cell type composition analysis |

```{r install-bioc}
BiocManager::install(c(
    "edgeR",
    "limma",
    "org.Hs.eg.db",
    "AnnotationDbi",
    "speckle"
))
```

When prompted "Update all/some/none? [a/s/n]:", you can type `a` to update all packages, or `n` if you prefer not to update existing packages.

### Step 3: Install CRAN Packages

The following CRAN packages form the core of our analysis toolkit:

| Package | Purpose |
|---------|---------|
| `Seurat` | Comprehensive toolkit for single cell analysis (clustering, visualisation, etc.) |
| `harmony` | Batch effect correction and data integration |
| `ggplot2` | Grammar of graphics for creating publication-quality plots |
| `patchwork` | Combining multiple ggplot2 plots into one figure |
| `dplyr` | Data manipulation (filtering, summarising, etc.) |
| `tidyr` | Data tidying (reshaping data frames) |
| `RColorBrewer` | Colour palettes for visualisation |
| `clustree` | Visualising how clusters change across different resolutions |
| `pheatmap` | Creating heatmaps with clustering |
| `remotes` | Installing packages from GitHub (useful for development versions) |

```{r install-cran}
install.packages(c(
    "Seurat",
    "harmony",
    "ggplot2",
    "patchwork",
    "dplyr",
    "tidyr",
    "RColorBrewer",
    "clustree",
    "pheatmap",
    "remotes"
))
```

## Verify Installation

Once all packages are installed, it is essential to verify that they load correctly. A package may fail to install silently, or there may be version conflicts that prevent loading. The following code attempts to load each required package and reports any failures:

```{r verify-installation, eval=TRUE}
# List of required packages
packages <- c(
    "Seurat",
    "harmony",
    "edgeR",
    "limma",
    "speckle",
    "org.Hs.eg.db",
    "ggplot2",
    "patchwork",
    "dplyr",
    "tidyr",
    "pheatmap",
    "clustree"
)

# Check which packages can be loaded
check_results <- sapply(packages, function(pkg) {
    requireNamespace(pkg, quietly = TRUE)
})

# Report results
missing <- names(check_results)[!check_results]

if (length(missing) > 0) {
    message("=== WARNING ===")
    message("The following packages failed to install or load:")
    for (pkg in missing) {
        message("  - ", pkg)
    }
    message("\nPlease try reinstalling these packages before the workshop.")
} else {
    message("=== SUCCESS ===")
    message("All required packages are installed and can be loaded.")
    message("Your environment is ready for the workshop!")
}
```

### Check Package Versions

Different versions of packages may behave differently. Here, we verify that your
installed package versions match the expected versions used in the workshop
materials. The workshop was developed and tested with these specific versions:

```{r check-versions, eval=TRUE}
# Expected versions for reproducibility
expected_versions <- list(
    # CRAN packages
    Seurat = "5.4.0",
    SeuratObject = "5.3.0",
    harmony = "1.2.4",
    ggplot2 = "4.0.1",
    patchwork = "1.3.2",
    dplyr = "1.1.4",
    tidyr = "1.3.2",
    # Bioconductor packages
    edgeR = "4.8.2",
    limma = "3.66.0",
    speckle = "1.10.0"
)

cat("Package Version Check\n")
cat(paste(rep("=", 55), collapse = ""), "\n")
cat(sprintf("%-20s %-12s %-12s %s\n", "Package", "Expected", "Installed", "Status"))
cat(paste(rep("-", 55), collapse = ""), "\n")

all_match <- TRUE
for (pkg in names(expected_versions)) {
    expected <- expected_versions[[pkg]]
    if (requireNamespace(pkg, quietly = TRUE)) {
        actual <- as.character(packageVersion(pkg))
        status <- if (actual == expected) "OK" else "MISMATCH"
        if (actual != expected) all_match <- FALSE
    } else {
        actual <- "NOT INSTALLED"
        status <- "MISSING"
        all_match <- FALSE
    }
    cat(sprintf("%-20s %-12s %-12s %s\n", pkg, expected, actual, status))
}

cat(paste(rep("=", 55), collapse = ""), "\n")
if (all_match) {
    cat("\nAll package versions match! Your environment is correctly configured.\n")
} else {
    cat("\nWARNING: Some package versions differ from expected.\n")
    cat("The workshop may still work, but outputs might differ slightly.\n")
    cat("Consider re-running the automatic installation to get exact versions.\n")
}
```

## Download Workshop Data

The workshop uses single-nucleus RNA-seq data from human heart tissue. The data
files are hosted on Zenodo and must be downloaded before running the workshop
modules.

### About the Dataset

The dataset contains ~43,000 nuclei (after quality control filtering) from 9
human heart samples across three developmental stages:

| Group | Samples | Age Range |
|-------|---------|-----------|
| Foetal | 3 | 19-20 weeks gestation |
| Young | 3 | 4-14 years |
| Adult | 3 | 35-42 years |

Data source: Sim et al. (2021) "Sex-Specific Control of Human Heart Maturation
by the Progesterone Receptor", *Circulation*.
DOI: [10.1161/CIRCULATIONAHA.120.051921](https://doi.org/10.1161/CIRCULATIONAHA.120.051921)

### Download the Data

Run the following code to download the workshop data files from Zenodo. This
will create a `data/` folder in your working directory containing the required
files.
**Note:** The total download size is approximately 200-300 MB. Download time
depends on your internet connection.

```{r download-data, eval=FALSE}
# Download workshop data from Zenodo
# This will create a data/ folder with the required files

# Zenodo record ID
zenodo_record <- "18237749"
base_url <- paste0("https://zenodo.org/records/", zenodo_record, "/files/")

# Files to download
files <- c("heart-counts.Rds", "cellinfo_updated.Rds")

# Create data directory
if (!dir.exists("data")) {
    dir.create("data")
    message("Created data/ directory")
}

# Download each file
for (f in files) {
    dest_file <- file.path("data", f)
    if (file.exists(dest_file)) {
        message(f, " already exists, skipping...")
        next
    }
    message("Downloading ", f, "...")
    download.file(
        url = paste0(base_url, f, "?download=1"),
        destfile = dest_file,
        mode = "wb"
    )
    message("  Done!")
}

message("\nDownload complete! Data saved to data/ folder.")
```

## Test Data Loading

As a final verification step, we check that the workshop data files are accessible and can be loaded correctly. The workshop uses two data files:

1. **heart-counts.Rds**: A sparse matrix containing gene expression counts (genes × cells)
2. **cellinfo_updated.Rds**: A data frame containing metadata for each cell (sample, group, etc.)

```{r test-data, eval=TRUE}
# Set path to data directory (relative to tutorials folder)
data_dir <- "../data"

# Check that data files exist
counts_file <- file.path(data_dir, "heart-counts.Rds")
cellinfo_file <- file.path(data_dir, "cellinfo_updated.Rds")

cat("Checking for workshop data files...\n")
cat(paste(rep("-", 40), collapse = ""), "\n")

if (file.exists(counts_file)) {
    cat("  heart-counts.Rds    : FOUND\n")

    # Load and report dimensions
    counts <- readRDS(counts_file)
    cat(sprintf("    - Dimensions: %d genes x %d cells\n", nrow(counts), ncol(counts)))
    rm(counts)  # Free memory immediately

} else {
    cat("  heart-counts.Rds    : NOT FOUND\n")
}

if (file.exists(cellinfo_file)) {
    cat("  cellinfo_updated.Rds: FOUND\n")
} else {
    cat("  cellinfo_updated.Rds: NOT FOUND\n")
}

if (file.exists(counts_file) && file.exists(cellinfo_file)) {
    cat("\n=== Data files are ready! ===\n")
} else {
    cat("\n=== WARNING: Some data files are missing ===\n")
    cat("Please check the data directory path.\n")
}
```

## Troubleshooting Common Issues

If you encounter problems during installation, this section provides solutions to the most common issues.

### Issue 1: "Package 'X' is not available for R version Y.Z"

**Cause**: Your R version is too old for the package.

**Solution**: Update R to version 4.3 or later from [CRAN](https://cran.r-project.org/), then restart RStudio and retry the installation.

### Issue 2: "There is no package called 'X'"

**Cause**: The package failed to install, possibly due to missing dependencies or network issues.

**Solution**: Try installing the specific package again:

```{r troubleshoot-missing}
# For CRAN packages:
install.packages("package_name")

# For Bioconductor packages:
BiocManager::install("package_name")

# For GitHub packages:
remotes::install_github("username/package_name")
```

### Issue 3: Compilation Errors on macOS

**Cause**: Missing Xcode Command Line Tools, which provide compilers needed to build some packages.

**Solution**: Open Terminal (not R) and run:

```bash
xcode-select --install
```

Follow the prompts to install, then restart RStudio and retry package installation.

### Issue 4: Memory Errors ("cannot allocate vector of size...")

**Cause**: Insufficient RAM available.

**Solution**:
1. Close other applications to free memory
2. Restart R (Session → Restart R in RStudio)
3. Try installing packages one at a time rather than all at once

### Issue 5: Permission Errors on Windows

**Cause**: R does not have permission to write to the package library.

**Solution**: Run RStudio as Administrator (right-click RStudio icon → "Run as administrator").

### Issue 6: Package Loading Conflicts

**Cause**: Multiple versions of a package are installed, or packages have conflicting dependencies.

**Solution**: Remove and reinstall the problematic package:

```{r troubleshoot-conflicts}
# Remove the package
remove.packages("package_name")

# Restart R, then reinstall
install.packages("package_name")
```

## Getting Help

If you encounter issues that you cannot resolve using this troubleshooting guide, please:

1. **Note the exact error message** - Copy the complete error text
2. **Record your R version** - Run `R.version.string`
3. **List installed package versions** - Run `sessionInfo()`
4. **Contact the workshop organisers** before the session

Having setup issues resolved before the workshop begins ensures that everyone can participate fully in the hands-on exercises.

## What to Expect in the Workshop

Once your environment is set up, you will be ready for the workshop, which covers:

1. **Module 1: Quality Control** - Loading data, calculating QC metrics, filtering low-quality cells
2. **Module 2: Integration & Clustering** - Normalisation, batch correction, dimensionality reduction, and clustering
3. **Module 3: Cell Type Annotation** - Identifying cell types using marker genes
4. **Module 4: Differential Expression** - Finding genes and cell types that differ between conditions

We look forward to seeing you at the workshop!

## Session Information

The output below records your R environment configuration, which is useful for reproducibility and troubleshooting:

```{r session-info, eval=TRUE}
sessionInfo()
```
