---
title: "Module 2: Normalisation, Integration & Clustering"
author: "Single Cell Workshop"
date: "`r Sys.Date()`"
output:
  html_document:
    theme: flatly
    toc: true
    toc_depth: 2
    toc_float: true
    number_sections: true
    code_folding: show
    df_print: paged
vignette: >
  %\VignetteIndexEntry{Module 2: Normalisation, Integration & Clustering}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
    echo = TRUE,
    warning = FALSE,
    message = FALSE,
    fig.align = "center",
    fig.width = 10,
    fig.height = 6
)

# Increase future globals size limit for SCTransform parallelization
options(future.globals.maxSize = 2000 * 1024^2)  # 2 GB
```

## Introduction

After quality control, we have a dataset of high-quality cells. However, before
we can identify cell types, we must address two important challenges:

1. **Technical variation**: Raw count data is affected by differences in
   sequencing depth between cells. A cell with more total counts will appear to
   have higher expression of most genes, even if the underlying biology is
   identical. Normalisation removes this technical variation.

2. **Batch effects**: Cells processed in different batches (different days,
   different samples) may show systematic differences unrelated to biology.
   Integration methods correct for these batch effects while preserving true
   biological variation.

In this module, we normalise the data using SCTransform, integrate samples using
Harmony, and identify cell clusters using graph-based clustering.

The overall workflow follows a standard single cell analysis pipeline:

```
Raw counts → Normalisation → PCA → Integration → UMAP → Clustering
```

Each step builds upon the previous one, progressively transforming the data from
raw gene counts into biologically interpretable cell clusters.

### Why Normalise?

Consider two cells: Cell A has 5,000 total UMI counts, while Cell B has 10,000. If Gene X has 100 counts in Cell A and 200 counts in Cell B, are these cells different? Not necessarily—both cells have 2% of their counts from Gene X. Without normalisation, we would incorrectly conclude that Cell B has higher expression.

**SCTransform** (Hafemeister & Satija, 2019) addresses this by:
1. Modelling the relationship between gene expression and sequencing depth
2. Computing Pearson residuals that remove the effect of sequencing depth
3. Identifying highly variable genes for downstream analysis

### Why Integrate?

Cells from different samples may cluster by sample rather than by cell type due to batch effects. This makes it difficult to compare cell types across conditions. **Harmony** (Korsunsky et al., 2019) corrects batch effects by:
1. Embedding cells in a low-dimensional space (PCA)
2. Iteratively adjusting embeddings so that samples are mixed within cell type clusters
3. Preserving biological variation while removing technical variation

### Learning Objectives

By the end of this module, you will be able to:

1. Understand why normalisation is necessary for scRNA-seq data
2. Apply SCTransform normalisation to single cell data
3. Perform dimensionality reduction with PCA
4. Integrate multiple samples using Harmony
5. Visualise cells using UMAP
6. Identify cell clusters using graph-based clustering

### Setup: Loading Required Libraries

```{r load-libraries}
library(Seurat)
library(harmony)
library(ggplot2)
library(patchwork)
library(dplyr)
library(tidyr)
library(tibble)
library(clustree)
library(RColorBrewer)
```

### Colour Palette

We use a consistent colour scheme throughout the workshop:

```{r define-colors}
# Developmental group colours (lowercase to match data)
group_colors <- c(
    "fetal" = "#E64B35",
    "young" = "#4DBBD5",
    "adult" = "#3C5488"
)

# Sample colours - shades within each group
sample_colors <- c(
    "f1" = "#E64B35", "f2" = "#F39B7F", "f3" = "#FFCAB0",
    "y1" = "#4DBBD5", "y2" = "#91D1C2", "y3" = "#C5E8E0",
    "a1" = "#3C5488", "a2" = "#8491B4", "a3" = "#B4BCD4"
)
```

## Loading Filtered Data

We begin by loading the quality-controlled data from Module 1:

```{r load-data}
# Load the QC-filtered Seurat object
data_dir <- "../data"
seu <- readRDS(file.path(data_dir, "processed/01_qc_filtered.rds"))

# Examine the object dimensions
dim(seu)
```

Let us verify the cell counts per sample:

```{r verify-samples}
table(seu$sample)
```

## Downsampling for Computational Efficiency

The full dataset contains over 40,000 cells. While modern laptops can handle this, processing time can be lengthy during a workshop. We downsample to 10,000 cells to ensure the analysis runs quickly while still capturing the biological diversity of the dataset.

Importantly, we use **stratified sampling** to ensure that all samples are represented proportionally. This preserves the relative contribution of each sample to the final dataset.

```{r downsample}
set.seed(42)  # For reproducibility

# Target number of cells
n_target <- 10000

# Calculate cells to sample from each sample (proportional)
cells_per_sample <- seu@meta.data %>%
    group_by(sample) %>%
    summarise(
        n_cells = n(),
        .groups = "drop"
    ) %>%
    mutate(
        proportion = n_cells / sum(n_cells),
        n_sample = round(proportion * n_target)
    )

# Adjust to ensure we get exactly n_target cells
# Add or remove from the largest group
diff <- n_target - sum(cells_per_sample$n_sample)
if (diff != 0) {
    largest_sample <- which.max(cells_per_sample$n_cells)
    cells_per_sample$n_sample[largest_sample] <-
        cells_per_sample$n_sample[largest_sample] + diff
}

cells_per_sample
```

Now we sample cells from each group:

```{r sample-cells}
# Sample cells from each sample
set.seed(42)

sampled_cells <- lapply(unique(seu$sample), function(s) {
    # Get cells for this sample
    cells <- colnames(seu)[seu$sample == s]
    # Number to sample
    n_to_sample <- cells_per_sample$n_sample[cells_per_sample$sample == s]
    # Sample
    sample(cells, size = min(n_to_sample, length(cells)))
})
sampled_cells <- unlist(sampled_cells)

# Subset the Seurat object
seu <- subset(seu, cells = sampled_cells)

# Verify the downsampling
cat("Downsampling complete:\n")
cat("- Total cells:", ncol(seu), "\n")
cat("- Samples represented:", length(unique(seu$sample)), "\n")
```

```{r verify-downsample}
# Cells per sample after downsampling
table(seu$sample)
```

## Normalisation with SCTransform

Normalisation is one of the most critical steps in single cell analysis. Unlike
bulk RNA-seq, where samples have similar total counts, single cells can vary by
orders of magnitude in their total UMI counts due to differences in cell size,
RNA content, and capture efficiency.

**SCTransform** (Hafemeister & Satija, 2019) uses regularised negative binomial
regression to normalise UMI count data. This approach:

1. Models the relationship between observed counts and sequencing depth for each
   gene using a negative binomial distribution
2. Computes Pearson residuals that represent the deviation from expected
   expression given the sequencing depth
3. Identifies genes with high biological variability (variable features) that
   will drive downstream analyses

The key insight is that highly expressed genes show more variance simply due to
sampling effects. SCTransform accounts for this by modelling the mean-variance
relationship, allowing fair comparison across genes with different expression
levels.

The `vst.flavor = "v2"` option uses an improved algorithm (Choudhary & Satija,
2022) that:

- Uses geometric mean of library size as reference (more robust)
- Implements improved regularisation for better handling of outliers
- Provides more stable results on large datasets

```{r sctransform}
# Run SCTransform normalisation
# This may take 1-2 minutes
set.seed(42)  # For reproducibility
seu <- SCTransform(
    seu,
    vst.flavor = "v2",
    seed.use = 42,
    verbose = FALSE
)
```

SCTransform automatically identifies highly variable genes. Let us examine the top variable features:

```{r variable-features}
# Get the top 10 variable features
top_features <- head(VariableFeatures(seu), 10)
top_features
```

We can visualise the mean-variance relationship:

```{r variance-plot, fig.width=8, fig.height=5}
# Plot variable features
VariableFeaturePlot(seu) +
    ggtitle("Variable Features (SCTransform)") +
    theme_minimal()
```

## Dimensionality Reduction with PCA

Single cell data is high-dimensional—we measure expression of thousands of genes
per cell. This creates challenges for visualisation and analysis:

- **Computational cost**: Calculating distances between cells in 20,000
  dimensions is slow
- **Curse of dimensionality**: In high dimensions, distances become less
  meaningful
- **Noise**: Many genes contain more noise than signal

**Principal Component Analysis (PCA)** addresses these issues by finding linear
combinations of genes (principal components, PCs) that capture the most
variation in the data. The first PC captures the most variance, the second PC
captures the most remaining variance orthogonal to the first, and so on.

Key concepts:

- **Loadings**: The contribution of each gene to a PC. High loading genes
  "define" that component.
- **Embeddings**: The coordinates of each cell in PC space. Used for downstream
  analysis.
- **Variance explained**: How much of the total variation each PC captures.

After PCA, we typically use only the first 20-50 PCs, which capture most of the
biological signal while discarding noise in higher components.

```{r run-pca}
# Run PCA on the SCTransform-normalised data
# By default, this uses the variable features identified by SCTransform
set.seed(42)
seu <- RunPCA(seu, seed.use = 42, verbose = FALSE)
```

### Visualising PCA Results

Let us examine the top genes contributing to the first two principal components:

```{r pca-loadings, fig.width=10, fig.height=6}
# Visualise genes contributing to PC1 and PC2
VizDimLoadings(seu, dims = 1:2, reduction = "pca")
```

We can plot cells in PCA space, coloured by sample:

```{r pca-by-sample, fig.width=10, fig.height=5}
# PCA plot coloured by sample
DimPlot(
    seu,
    reduction = "pca",
    group.by = "sample",
    shuffle = TRUE
) +
    ggtitle("PCA - Before Integration") +
    theme_minimal()
```

Notice how cells from different samples may cluster separately. This could reflect:
- True biological differences between samples
- Technical batch effects that should be removed

### Selecting the Number of Dimensions

We use an elbow plot to determine how many PCs to use for downstream analysis. The elbow occurs where adding more PCs provides diminishing returns:

```{r elbow-plot, fig.width=8, fig.height=5}
ElbowPlot(seu, ndims = 30) +
    ggtitle("PCA Elbow Plot") +
    theme_minimal()
```

The elbow plot helps identify where adding more PCs provides diminishing returns.
For this dataset, the curve flattens around 15-20 PCs, suggesting that the first
20 dimensions capture most of the meaningful biological variation:

```{r set-dims}
n_dims <- 20
```

## Integration with Harmony

When analysing data from multiple samples or batches, cells often cluster by
their sample of origin rather than by cell type. This happens because each
sample has unique technical characteristics (slight differences in dissociation,
library preparation, sequencing) that create systematic biases.

**Harmony** (Korsunsky et al., 2019) is a fast and effective integration method
that corrects these batch effects while preserving biological variation. The
algorithm works as follows:

1. **Soft clustering**: Cells are assigned probabilistically to multiple
   clusters in PCA space
2. **Calculate centroids**: For each cluster, compute the centroid separately
   for each batch
3. **Adjust embeddings**: Move cells toward the global centroid, removing
   batch-specific shifts
4. **Iterate**: Repeat until convergence

The key advantage of Harmony is that it operates directly on the PCA embeddings,
making it fast and memory-efficient. Unlike methods that require re-analysis of
raw counts, Harmony can integrate datasets in seconds to minutes.

Important considerations:

- **Over-integration**: Aggressive correction can merge distinct cell types. We
  check this by examining marker genes after clustering.
- **Under-integration**: Insufficient correction leaves batch effects. We verify
  by checking sample mixing in UMAP.
- **Biological variation**: Some sample-specific patterns are real biology (e.g.,
  disease vs healthy). Harmony should preserve these while removing technical
  artefacts.

```{r run-harmony}
# Run Harmony integration
# group.by.vars specifies which variable contains batch information
# Here we integrate by sample to remove sample-specific technical effects
set.seed(42)  # Harmony has stochastic components
seu <- RunHarmony(
    seu,
    group.by.vars = "sample",
    reduction.use = "pca",
    dims.use = 1:n_dims,
    verbose = FALSE
)
```

### Comparing Before and After Integration

Let us compare the embeddings before and after Harmony:

```{r compare-integration, fig.width=12, fig.height=5}
# Before integration (PCA)
p1 <- DimPlot(
    seu,
    reduction = "pca",
    group.by = "sample",
    cols = sample_colors,
    shuffle = TRUE
) +
    ggtitle("Before Integration (PCA)") +
    theme_minimal() +
    NoLegend()

# After integration (Harmony)
p2 <- DimPlot(
    seu,
    reduction = "harmony",
    group.by = "sample",
    cols = sample_colors,
    shuffle = TRUE
) +
    ggtitle("After Integration (Harmony)") +
    theme_minimal()

p1 + p2
```

After Harmony, cells from different samples should be more mixed, while cells of the same type should remain close together.

## UMAP Visualisation

While PCA is excellent for capturing global structure and variance, it uses
linear projections that cannot represent complex, non-linear relationships
between cells. **UMAP** (Uniform Manifold Approximation and Projection; McInnes
et al., 2018) addresses this limitation.

UMAP works by:

1. **Building a high-dimensional graph**: Connect each cell to its nearest
   neighbours in the high-dimensional space (Harmony embeddings)
2. **Optimising a low-dimensional representation**: Find 2D coordinates that
   preserve the neighbour relationships as much as possible

Key properties of UMAP:

- **Local structure preserved**: Cells that are close in the original space
  remain close in UMAP
- **Global structure approximate**: Distances between distant groups are less
  meaningful
- **Clusters appear as islands**: Distinct cell types typically form separate
  "islands" in UMAP space

**Important caveats**:

- UMAP is primarily for **visualisation**, not for quantitative analysis
- The size and density of clusters in UMAP do not reflect the number of cells
- Do not over-interpret distances between clusters

We run UMAP on the Harmony-corrected embeddings to ensure batch effects do not
distort the visualisation:

```{r run-umap}
set.seed(42)  # UMAP is stochastic - seed ensures reproducible embeddings
seu <- RunUMAP(
    seu,
    reduction = "harmony",
    dims = 1:n_dims,
    seed.use = 42,
    verbose = FALSE
)
```

### UMAP by Sample and Group

```{r umap-sample, fig.width=12, fig.height=5}
# UMAP coloured by sample
p1 <- DimPlot(
    seu,
    reduction = "umap",
    group.by = "sample",
    cols = sample_colors,
    shuffle = TRUE
) +
    ggtitle("UMAP - By Sample") +
    theme_minimal()

# UMAP coloured by developmental group
p2 <- DimPlot(
    seu,
    reduction = "umap",
    group.by = "group",
    cols = group_colors,
    shuffle = TRUE
) +
    ggtitle("UMAP - By Group") +
    theme_minimal()

p1 + p2
```

The cells from different samples should now be well-mixed within cell type clusters, indicating successful batch correction.

## Graph-Based Clustering

Clustering identifies groups of cells with similar gene expression profiles.
Unlike traditional k-means clustering, single cell analysis typically uses
**graph-based clustering**, which does not require specifying the number of
clusters in advance.

### Why Graph-Based Clustering?

Traditional clustering methods like k-means have limitations for single cell
data:

- They require pre-specifying the number of clusters (k)
- They assume spherical clusters of similar size
- They are sensitive to outliers

Graph-based clustering overcomes these limitations by representing cells as a
network and identifying densely connected communities. This approach naturally
handles:

- Variable numbers of cell types
- Clusters of different sizes
- Complex cluster shapes

### The Two-Step Process

Seurat implements graph-based clustering in two steps:

**Step 1: Build a K-Nearest Neighbour (KNN) Graph**

Each cell is connected to its K nearest neighbours in the reduced dimensional
space (Harmony embeddings). Two cells are connected if they have similar gene
expression profiles. The result is a graph where:

- Nodes = cells
- Edges = connections between similar cells
- Edge weights = strength of similarity

**Step 2: Community Detection (Louvain Algorithm)**

The Louvain algorithm identifies "communities" of densely connected cells. It
optimises **modularity**—a measure of how much more densely connected cells
within a community are compared to random expectation. The algorithm:

1. Starts with each cell in its own community
2. Iteratively moves cells between communities to improve modularity
3. Aggregates communities and repeats until no further improvement

### Building the Neighbour Graph

The `FindNeighbors()` function constructs a **shared nearest neighbour (SNN)**
graph. Unlike simple KNN, two cells are connected based on how many neighbours
they share, not just whether they are neighbours. This makes clustering more
robust to noise.

```{r find-neighbors}
seu <- FindNeighbors(
    seu,
    reduction = "harmony",
    dims = 1:n_dims,
    verbose = FALSE
)
```

### Finding Clusters

The **resolution** parameter is the main tuning parameter for clustering. It
controls the granularity of the identified clusters:

- **Lower resolution (0.1-0.3)**: Produces fewer, larger clusters. Identifies
  major cell types but may merge subtypes.
- **Medium resolution (0.4-0.6)**: Balances major types and subtypes. Often a
  good starting point.
- **Higher resolution (0.8-1.2)**: Produces more, smaller clusters. Can reveal
  rare subtypes but may over-split continuous populations.

There is no "correct" resolution—the appropriate choice depends on the
biological question. If you are interested in major cell types, use lower
resolution. If you want to identify rare subtypes, use higher resolution.

We test a range of resolutions from 0.1 to 1.0 to understand how clusters behave
across the resolution spectrum:

```{r find-clusters}
# Test resolutions from 0.1 to 1.0 in steps of 0.1
resolutions <- seq(0.1, 1.0, by = 0.1)

set.seed(42)  # Louvain clustering has stochastic components
for (res in resolutions) {
    seu <- FindClusters(
        seu,
        resolution = res,
        verbose = FALSE
    )
}

# List the clustering results
grep("SCT_snn_res", colnames(seu@meta.data), value = TRUE)
```

Let us see how many clusters each resolution produces:

```{r cluster-counts}
# Count clusters at each resolution
cat("Number of clusters at each resolution:\n")
for (res in resolutions) {
    col_name <- paste0("SCT_snn_res.", res)
    n_clusters <- length(unique(seu@meta.data[[col_name]]))
    cat(sprintf("  Resolution %.1f: %d clusters\n", res, n_clusters))
}
```

### Visualising Cluster Resolution with Clustree

**Clustree** (Zappia & Oshlack, 2018) is a diagnostic tool that visualises how
clusters evolve across resolutions. Each node represents a cluster at a given
resolution, and edges show how cells flow between clusters.

```{r clustree, fig.width=12, fig.height=10}
clustree(seu, prefix = "SCT_snn_res.")
```

**How to interpret the clustree plot:**

1. **Read from top to bottom**: Resolution increases downward. At the top
   (res=0.1), there are few clusters. As you move down, clusters split.

2. **Thick edges** indicate stable clusters: Most cells stay together as
   resolution increases. These likely represent robust, biologically distinct
   populations.

3. **Thin, branching edges** indicate unstable clusters: Cells are split across
   multiple downstream clusters. This may represent over-clustering of a
   continuous population, or genuine subtypes being resolved.

4. **Look for the "Goldilocks zone"**: A resolution where most clusters are
   stable (thick edges going straight down) but before excessive splitting
   begins.

5. **Node size** is proportional to the number of cells in that cluster.

### Selecting a Resolution

Based on the clustree visualisation, we select a working resolution. Key
considerations:

- Are the major cell types separated? (Should happen at low resolution)
- Are biologically meaningful subtypes visible? (May require medium resolution)
- Is there excessive splitting of continuous populations? (Sign of too high
  resolution)

The appropriate resolution depends on the dataset and biological question. For
this heart development dataset, we observe that:

- Resolution 0.1-0.2 groups cells into broad categories but may merge distinct
  populations
- Resolution 0.3-0.4 produces a stable cluster structure
- Higher resolutions progressively split clusters further

We select resolution 0.4 for this analysis, which produces 17 clusters with good
separation of major cell types. Notably, several clusters are strongly enriched
for fetal cells (>95%), reflecting the distinct transcriptional programmes active
during early heart development. The remaining clusters contain cells from multiple
developmental stages, representing shared cell types like fibroblasts, endothelial
cells, and immune cells that are present throughout development:

```{r select-resolution}
# Set the active identity to our chosen resolution
selected_res <- 0.4
Idents(seu) <- paste0("SCT_snn_res.", selected_res)

# Store in a standard column
seu$seurat_clusters <- Idents(seu)

# How many clusters?
cat("Clustering results:\n")
cat("- Resolution:", selected_res, "\n")
cat("- Number of clusters:", length(unique(seu$seurat_clusters)), "\n")
```

### UMAP with Clusters

Now we visualise the clusters on the UMAP. Each cluster is assigned a number (0, 1, 2, ...) rather than a biological label—we will assign cell type names in the next module.

```{r umap-clusters, fig.width=10, fig.height=6}
DimPlot(
    seu,
    reduction = "umap",
    group.by = "seurat_clusters",
    label = TRUE,
    label.size = 5
) +
    ggtitle(paste0("UMAP - Clusters (Resolution ", selected_res, ")")) +
    theme_minimal() +
    NoLegend()
```

Each cluster should form a distinct region in UMAP space. Clusters that overlap spatially may represent closely related cell types or states.

### Cluster Composition by Sample

A key quality check is examining whether clusters contain cells from multiple samples. If a cluster is dominated by a single sample, it may represent a technical artefact rather than a true cell type.

```{r cluster-composition, fig.width=12, fig.height=5}
# Calculate composition
composition <- seu@meta.data %>%
    group_by(seurat_clusters, sample) %>%
    summarise(n = n(), .groups = "drop") %>%
    group_by(seurat_clusters) %>%
    mutate(proportion = n / sum(n))

# Stacked bar plot with our sample colour palette
ggplot(composition, aes(x = seurat_clusters, y = proportion, fill = sample)) +
    geom_bar(stat = "identity") +
    scale_fill_manual(values = sample_colors) +
    labs(
        x = "Cluster",
        y = "Proportion",
        fill = "Sample",
        title = "Cluster Composition by Sample"
    ) +
    theme_minimal()
```

If integration worked well, each cluster should contain cells from multiple samples (mixed colours).

### Cluster Composition by Group

```{r cluster-by-group, fig.width=12, fig.height=5}
# Calculate composition by group
composition_group <- seu@meta.data %>%
    group_by(seurat_clusters, group) %>%
    summarise(n = n(), .groups = "drop") %>%
    group_by(seurat_clusters) %>%
    mutate(proportion = n / sum(n))

# Stacked bar plot with our group colour palette
ggplot(composition_group, aes(x = seurat_clusters, y = proportion, fill = group)) +
    geom_bar(stat = "identity") +
    scale_fill_manual(values = group_colors) +
    labs(
        x = "Cluster",
        y = "Proportion",
        fill = "Group",
        title = "Cluster Composition by Developmental Group"
    ) +
    theme_minimal()
```

Some clusters may be enriched for specific developmental stages, reflecting biological differences in cell type composition.

## Save Checkpoint

We save the integrated and clustered object for use in the next module:

```{r save-checkpoint}
# Create output directory if needed
output_dir <- "../data/processed"
if (!dir.exists(output_dir)) {
    dir.create(output_dir, recursive = TRUE)
}

# Save the object
output_file <- file.path(output_dir, "02_integrated_clustered.rds")
saveRDS(seu, output_file)

message("Saved: ", output_file)
```

## Summary

In this module, we:

- Downsampled to `r ncol(seu)` cells using stratified sampling
- Normalised with SCTransform v2
- Reduced dimensions with PCA (`r n_dims` components)
- Integrated samples with Harmony
- Visualised with UMAP
- Identified `r length(unique(seu$seurat_clusters))` clusters at resolution `r selected_res`

The clustered object is now ready for cell type annotation in Module 3.

## Session Information

```{r session-info}
sessionInfo()
```
