---
title: "Module 3: Cell Type Annotation"
author: "Single Cell Workshop"
date: "`r Sys.Date()`"
output:
  html_document:
    theme: flatly
    toc: true
    toc_depth: 2
    toc_float: true
    number_sections: true
    code_folding: show
    df_print: paged
vignette: >
  %\VignetteIndexEntry{Module 3: Cell Type Annotation}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
    echo = TRUE,
    warning = FALSE,
    message = FALSE,
    fig.align = "center",
    fig.width = 10,
    fig.height = 6
)
```

## Introduction

In this module, we identify cell types by analysing marker genes that distinguish
each cluster. Cell type annotation is a critical step in single cell analysis,
as it transforms abstract cluster numbers into biologically meaningful cell
populations.

There are two main approaches to cell type annotation:

1. **Marker-based annotation** (manual): Examine differentially expressed genes
   in each cluster and match them to known cell type markers from the literature.
   This approach requires domain knowledge but provides full control and
   interpretability.

2. **Reference-based annotation** (automated): Use tools like Azimuth, SingleR,
   or scArches to transfer labels from a reference dataset. This is faster but
   may miss context-specific populations.

In this workshop, we use marker-based annotation to teach the fundamental
concepts. The human heart contains several major cell types:

- **Cardiomyocytes**: The contractile muscle cells, marked by TNNT2, TTN, MYH7
- **Fibroblasts**: Structural cells producing extracellular matrix (DCN, COL1A1)
- **Endothelial cells**: Line blood vessels (PECAM1, VWF, CDH5)
- **Pericytes/Smooth muscle**: Support vascular structures (ACTA2, RGS5)
- **Immune cells**: Macrophages and lymphocytes (PTPRC, CD68, CD163)
- **Epicardial cells**: Cover the heart surface (WT1, TBX18)
- **Neural cells**: Neurons and Schwann cells (NRXN1, PLP1)

## Load Libraries and Data

```{r load-libraries}
library(Seurat)
library(ggplot2)
library(dplyr)
library(tidyr)
library(patchwork)
library(RColorBrewer)
library(pheatmap)
```

We load the integrated and clustered Seurat object from Module 2.

```{r load-data}
# Load the clustered data
seu <- readRDS("../data/processed/02_integrated_clustered.rds")

# Verify the data
cat("Loaded Seurat object:\n")
cat("- Cells:", ncol(seu), "\n")
cat("- Genes:", nrow(seu), "\n")
cat("- Clusters:", length(unique(seu$seurat_clusters)), "\n")
```

Let us first examine the cluster distribution to understand what we are
annotating.

```{r cluster-overview}
# Cluster sizes
cluster_sizes <- table(seu$seurat_clusters)

cat("Cluster sizes:\n")
for (i in seq_along(cluster_sizes)) {
    cat(sprintf("  Cluster %s: %d cells\n", names(cluster_sizes)[i], cluster_sizes[i]))
}
```

## Define Colour Palette

We maintain consistent colours throughout our analysis.

```{r define-colors}
# Group colours (developmental stage - lowercase to match data)
group_colors <- c(
    "fetal" = "#E64B35",
    "young" = "#4DBBD5",
    "adult" = "#3C5488"
)

# Generate a palette for clusters
n_clusters <- length(unique(seu$seurat_clusters))
cluster_colors <- colorRampPalette(brewer.pal(12, "Paired"))(n_clusters)
names(cluster_colors) <- levels(seu$seurat_clusters)
```

## Visualise Current Clusters

Before annotation, let us examine the clusters and their composition by
developmental stage.

```{r umap-clusters, fig.width=12, fig.height=5}
# UMAP coloured by cluster and by group
p1 <- DimPlot(seu, group.by = "seurat_clusters", label = TRUE,
              label.size = 4, repel = TRUE, cols = cluster_colors) +
    ggtitle("Clusters") +
    theme(legend.position = "none")

p2 <- DimPlot(seu, group.by = "group", cols = group_colors) +
    ggtitle("Developmental Stage")

p1 + p2
```

Understanding the developmental composition of each cluster provides important
context for annotation.

```{r cluster-composition, fig.width=10, fig.height=5}
# Calculate cluster composition by group
comp_data <- seu@meta.data %>%
    group_by(seurat_clusters, group) %>%
    summarise(n = n(), .groups = "drop") %>%
    group_by(seurat_clusters) %>%
    mutate(pct = 100 * n / sum(n))

ggplot(comp_data, aes(x = seurat_clusters, y = pct, fill = group)) +
    geom_bar(stat = "identity", position = "stack") +
    scale_fill_manual(values = group_colors) +
    labs(x = "Cluster", y = "Percentage", fill = "Stage",
         title = "Developmental Stage Composition per Cluster") +
    theme_minimal() +
    theme(axis.text.x = element_text(size = 10))
```

We observe that some clusters are highly enriched for specific developmental
stages. For example, clusters 0, 6, 8, 9, 12, 15, 16, and 17 are nearly
exclusively fetal, while cluster 2 contains predominantly young and adult cells,
and cluster 18 is exclusively young. This developmental segregation often
reflects biological differences in cell state, particularly among cardiomyocytes
where fetal cells have distinct transcriptional profiles from mature cells.

## Find Cluster Markers

We use `FindAllMarkers()` to identify genes that distinguish each cluster from
all other cells. This is the foundation of marker-based annotation.

```{r find-markers}
# Set default assay to SCT
DefaultAssay(seu) <- "SCT"

# Prepare for marker finding with SCT assay
seu <- PrepSCTFindMarkers(seu)

# Find markers for all clusters
# We use only.pos = TRUE to find upregulated markers
# min.pct = 0.25 requires the gene to be expressed in at least 25% of cells
# logfc.threshold = 0.5 requires at least 0.5 log2 fold change
markers <- FindAllMarkers(
    seu,
    only.pos = TRUE,
    min.pct = 0.25,
    logfc.threshold = 0.5,
    test.use = "wilcox"
)

cat("Found", nrow(markers), "marker genes across all clusters\n")
```

Let us examine the top markers for each cluster.

```{r top-markers}
# Get top 5 markers per cluster
top_markers <- markers %>%
    group_by(cluster) %>%
    slice_max(order_by = avg_log2FC, n = 5) %>%
    select(cluster, gene, avg_log2FC, pct.1, pct.2, p_val_adj)

# Display top markers for each cluster
for (cl in levels(seu$seurat_clusters)) {
    cl_markers <- top_markers %>% filter(cluster == cl)
    cat("\n--- Cluster", cl, "---\n")
    for (i in seq_len(nrow(cl_markers))) {
        cat(sprintf("  %s: log2FC=%.2f, pct.1=%.0f%%, pct.2=%.0f%%\n",
                    cl_markers$gene[i],
                    cl_markers$avg_log2FC[i],
                    cl_markers$pct.1[i] * 100,
                    cl_markers$pct.2[i] * 100))
    }
}
```

### Visualise Top Markers with DotPlot

The DotPlot below shows the top 5 differentially expressed genes for each
cluster. This provides an unbiased view of what distinguishes each cluster
before we examine the cell type markers from the original study.

```{r dotplot-top-markers, fig.width=16, fig.height=8}
# Get top 5 unique markers per cluster, ordered by cluster
top5_genes <- top_markers %>%
    arrange(cluster, desc(avg_log2FC)) %>%
    pull(gene) %>%
    unique()

DotPlot(seu, features = top5_genes, group.by = "seurat_clusters") +
    RotatedAxis() +
    scale_colour_gradient2(low = "blue", mid = "white", high = "red",
                           midpoint = 0) +
    labs(title = "Top 5 Marker Genes per Cluster (from FindAllMarkers)") +
    theme(axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5, size = 7),
          axis.text.y = element_text(size = 10))
```

### Marker Heatmap

A heatmap of these top markers provides a comprehensive view of the expression
patterns across clusters.

```{r marker-heatmap-top, fig.width=12, fig.height=10}
# Get top 3 markers per cluster for heatmap
top3_markers <- markers %>%
    group_by(cluster) %>%
    slice_max(order_by = avg_log2FC, n = 3) %>%
    pull(gene) %>%
    unique()

# Downsample for visualisation (50 cells per cluster max)
set.seed(42)
cells_subset <- seu@meta.data %>%
    mutate(cell_id = rownames(seu@meta.data)) %>%
    group_by(seurat_clusters) %>%
    slice_sample(n = 50) %>%
    pull(cell_id)

# Create heatmap
DoHeatmap(subset(seu, cells = cells_subset),
          features = top3_markers,
          group.by = "seurat_clusters",
          size = 3) +
  scale_fill_gradient2(low = "blue", mid = "white", high = "red",
                       midpoint = 0) +
  theme(axis.text.y = element_text(size = 6))
```

## Cell Type Marker Analysis

While the differentially expressed genes provide a starting point, we typically
compare against known marker genes for the cell types expected in our tissue.
The human heart contains several well-characterised cell populations.

### Cell Type Marker Genes

The marker genes below were identified by Sim et al. 2021 (Circulation) from
their analysis of the full human heart development dataset. These markers
distinguish the major cell populations across foetal, young, and adult hearts.

#### Visualise with DotPlot

The DotPlot is a powerful visualisation that shows both the percentage of cells
expressing a gene (dot size) and the average expression level (colour intensity).

```{r dotplot-celltype-markers, fig.width=16, fig.height=12}
# Define markers and cell types from Sim et al. 2021
marker_info <- data.frame(
    gene = c(
        # Cardiomyocytes
        "TNNT2", "ACTN2", "TNNI3", "RYR2", "MYH6", "TTN", "MYBPC3", "PLN", "MYL2",
        # Fibroblasts
        "COL1A1", "COL3A1", "DCN", "PDGFRA", "POSTN", "TCF21", "VIM",
        # Pericytes
        "KCNJ8", "VTN", "ABCC9", "HEYL",
        # Endothelial
        "PECAM1", "KDR", "CDH5", "EMCN", "FLT1",
        # Smooth Muscle
        "TAGLN", "MYH11", "MYLK", "LMOD1",
        # Macrophages
        "CD68", "CD163", "CD74", "F13A1", "LGALS3",
        # T Cells
        "CD3D", "CD3G", "CD3E", "IL7R",
        # Schwann/Neural
        "PLP1", "CNP", "S100B",
        # Epicardial
        "UPK3B", "WT1",
        # Proliferating
        "MKI67", "TOP2A"
    ),
    celltype = c(
        rep("Cardiomyocytes", 9),
        rep("Fibroblasts", 7),
        rep("Pericytes", 4),
        rep("Endothelial", 5),
        rep("Smooth Muscle", 4),
        rep("Macrophages", 5),
        rep("T Cells", 4),
        rep("Schwann", 3),
        rep("Epicardial", 2),
        rep("Proliferating", 2)
    )
)

# Define colors for each cell type
celltype_colors <- c(
    "Cardiomyocytes" = "#D62728",
    "Fibroblasts" = "#1F77B4",
    "Pericytes" = "#2CA02C",
    "Endothelial" = "#9467BD",
    "Smooth Muscle" = "#FF7F0E",
    "Macrophages" = "#8C564B",
    "T Cells" = "#E377C2",
    "Schwann" = "#7F7F7F",
    "Epicardial" = "#17BECF",
    "Proliferating" = "#333333"
)

# Filter to available genes
marker_info <- marker_info[marker_info$gene %in% rownames(seu), ]
marker_genes <- marker_info$gene
gene_colors <- celltype_colors[marker_info$celltype]
marker_info$celltype_f <- factor(marker_info$celltype, levels = names(celltype_colors))

# Order clusters by cell type identity for diagonal pattern
# Preferred order based on cell type (CM, Fibroblasts, Pericytes, Endothelial, Immune, Neural, etc.)
preferred_order <- c("0", "6", "8", "17", "7", "9", "2", "11", "15", "1", "12", "3", "5", "13", "18", "4", "14", "16", "10")
# Filter to only clusters that exist in our data
actual_clusters <- as.character(unique(seu$seurat_clusters))
cluster_order <- preferred_order[preferred_order %in% actual_clusters]
# Add any clusters not in preferred order (shouldn't happen, but safety)
cluster_order <- c(cluster_order, setdiff(actual_clusters, cluster_order))
seu$cluster_ordered <- factor(seu$seurat_clusters, levels = cluster_order)

# Calculate cell type label positions
celltype_positions <- marker_info %>%
    mutate(pos = row_number()) %>%
    group_by(celltype_f) %>%
    summarise(start = min(pos) - 0.5, end = max(pos) + 0.5,
              mid = mean(pos), .groups = "drop")

# Create DotPlot with cell type annotations
DotPlot(seu, features = marker_genes, group.by = "cluster_ordered") +
    RotatedAxis() +
    scale_colour_gradient2(low = "#2166AC", mid = "white", high = "#B2182B",
                           midpoint = 0, name = "Average\nExpression") +
    scale_size_continuous(name = "Percent\nExpressed", range = c(0.5, 6)) +
    labs(x = NULL, y = "Cluster",
         title = "Cell Type Marker Expression (Sim et al. 2021)") +
    theme_bw() +
    theme(
        plot.title = element_text(size = 16, face = "bold", hjust = 0.5),
        axis.text.x = element_text(angle = 45, hjust = 1, size = 10,
                                   color = gene_colors, face = "italic"),
        axis.text.y = element_text(size = 12),
        axis.title.y = element_text(size = 14, face = "bold"),
        panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(),
        panel.border = element_rect(color = "grey40", linewidth = 0.8),
        legend.position = "right",
        legend.title = element_text(size = 11, face = "bold"),
        legend.text = element_text(size = 10),
        plot.margin = margin(10, 10, 45, 10)
    ) +
    # Add cell type color bars
    annotate("rect",
             xmin = celltype_positions$start, xmax = celltype_positions$end,
             ymin = -1.8, ymax = -1.2,
             fill = celltype_colors[as.character(celltype_positions$celltype_f)]) +
    # Add cell type labels inside bars
    annotate("text",
             x = celltype_positions$mid, y = -1.5,
             label = celltype_positions$celltype_f,
             size = 3.5, angle = 0, hjust = 0.5, fontface = "bold",
             color = "white") +
    coord_cartesian(clip = "off", ylim = c(0.5, length(cluster_order) + 0.5))
```

The DotPlot reveals a clear diagonal pattern where each cell type shows specific
marker expression. Gene names are coloured by their cell type, matching the
annotation bar at the bottom:

- **Cardiomyocytes** (red): TNNT2, ACTN2, TNNI3, RYR2, MYH6, TTN, MYBPC3, PLN, MYL2
- **Fibroblasts** (blue): COL1A1, COL3A1, DCN, PDGFRA, POSTN, TCF21, VIM
- **Pericytes** (green): KCNJ8, VTN, ABCC9, HEYL
- **Endothelial** (purple): PECAM1, KDR, CDH5, EMCN, FLT1
- **Smooth Muscle** (orange): TAGLN, MYH11, MYLK, LMOD1
- **Macrophages** (brown): CD68, CD163, CD74, F13A1, LGALS3
- **T Cells** (pink): CD3D, CD3G, CD3E, IL7R
- **Schwann** (grey): PLP1, CNP, S100B
- **Epicardial** (teal): UPK3B, WT1
- **Proliferating** (black): MKI67, TOP2A

### Feature Plots for Key Markers

FeaturePlots show the spatial distribution of marker expression on the UMAP,
helping us understand the relationship between clusters.

```{r featureplot-cm, fig.width=12, fig.height=8}
# Cardiomyocyte markers
FeaturePlot(seu,
            features = c("TNNT2", "TTN", "MYH7", "MYH6"),
            ncol = 2, order = TRUE) &
    scale_colour_gradient(low = "lightgrey", high = "darkred") &
    theme_minimal()
```

The cardiomyocyte markers reveal an interesting pattern: MYH6 (atrial isoform)
is enriched in clusters 2 and 9, while MYH7 (ventricular isoform) is more
expressed in clusters 0, 11, 15, and 17. This suggests we can distinguish atrial
from ventricular cardiomyocytes.

```{r featureplot-stromal, fig.width=12, fig.height=8}
# Stromal cell markers
FeaturePlot(seu,
            features = c("DCN", "PECAM1", "KCNJ8", "ACTA2"),
            ncol = 2, order = TRUE) &
    scale_colour_gradient(low = "lightgrey", high = "darkblue") &
    theme_minimal()
```

```{r featureplot-immune, fig.width=12, fig.height=8}
# Immune and other markers
FeaturePlot(seu,
            features = c("PTPRC", "F13A1", "NRXN1", "MKI67"),
            ncol = 2, order = TRUE) &
    scale_colour_gradient(low = "lightgrey", high = "darkgreen") &
    theme_minimal()
```

## Assign Cell Type Annotations

Based on the marker analysis above, we can now assign cell type labels to each
cluster. We consider both the marker gene expression and the developmental
stage composition.

**Note:** These annotations are based on the 10,000 cell subset used for this
workshop. The full dataset (~54,000 cells) may reveal additional rare
populations or finer subclusters.

The table below summarises our annotations with the supporting marker evidence:

| Cluster | Cell Type | Key Markers | Notes |
|---------|-----------|-------------|-------|
| 0 | Fetal CM | TNNT2+, TTN+, MYH7+ | Fetal ventricular cardiomyocytes (100% fetal) |
| 1 | Fibroblasts | DCN+, COL1A1+ | Present across all developmental stages |
| 2 | Atrial CM | TTN+, MYH6++ | Mature atrial CM (young/adult enriched) |
| 3 | Pericytes | KCNJ8+, RGS5+, ABCC9+ | Vascular support cells |
| 4 | Macrophages | CD163+, F13A1+, PTPRC+ | Tissue-resident macrophages |
| 5 | Endothelial | PECAM1+, VWF+, CDH5+ | Main endothelial population |
| 6 | Fetal CM | TNNT2+, TTN+ | Fetal ventricular cardiomyocytes (100% fetal) |
| 7 | Activated CM | TTN+, TNNT2+, HLA-DR+ | CM with immune activation/MHC class II (90% young) |
| 8 | Fetal CM | TNNT2+, TTN+ | Fetal ventricular cardiomyocytes (100% fetal) |
| 9 | Fetal Atrial CM | TTN+, MYH6++ | Fetal atrial cardiomyocytes (100% fetal) |
| 10 | Proliferating | MKI67+, TOP2A+ | Proliferating cells (fetal enriched) |
| 11 | Ventricular CM | TNNT2++, TTN++, MYH7++ | High-expressing ventricular CM |
| 12 | Fetal Fibroblasts | DCN+, COL1A1++ | Fetal fibroblasts (99% fetal) |
| 13 | Endothelial | PECAM1+ | Endothelial subset |
| 14 | Immune | PTPRC+, CD3E+ | Mixed immune/lymphocytes |
| 15 | Ventricular CM | TTN+, MYH7+ | Ventricular CM (fetal enriched) |
| 16 | Neurons | NRXN1++ | Neural cells (100% fetal) |
| 17 | Fetal CM | TNNT2+, TTN++, MYH7+ | Fetal ventricular cardiomyocytes (98% fetal) |
| 18 | Young Endothelial | PECAM1+, VWF++ | Endothelial cells (100% young) |


```{r define-annotations}
# Define cell type annotations for all 19 clusters
cluster_annotations <- c(
  "0" = "Fetal CM",
  "1" = "Fibroblasts",
  "2" = "Atrial CM",
  "3" = "Pericytes",
  "4" = "Macrophages",
  "5" = "Endothelial",
  "6" = "Fetal CM",
  "7" = "Activated CM",
  "8" = "Fetal CM",
  "9" = "Fetal Atrial CM",
  "10" = "Proliferating",
  "11" = "Ventricular CM",
  "12" = "Fetal Fibroblasts",
  "13" = "Endothelial",
  "14" = "Immune",
  "15" = "Ventricular CM",
  "16" = "Neurons",
  "17" = "Fetal CM",
  "18" = "Young Endothelial"
)

cell_type_df <- data.frame(
  cell_type = cluster_annotations[as.character(seu$seurat_clusters)],
  row.names = colnames(seu)
)
seu <- AddMetaData(seu, metadata = cell_type_df)

# Broad categories for summary analyses
broad_annotations <- c(
  "0" = "Cardiomyocytes",
  "1" = "Fibroblasts",
  "2" = "Cardiomyocytes",
  "3" = "Pericytes",
  "4" = "Immune",
  "5" = "Endothelial",
  "6" = "Cardiomyocytes",
  "7" = "Cardiomyocytes",
  "8" = "Cardiomyocytes",
  "9" = "Cardiomyocytes",
  "10" = "Proliferating",
  "11" = "Cardiomyocytes",
  "12" = "Fibroblasts",
  "13" = "Endothelial",
  "14" = "Immune",
  "15" = "Cardiomyocytes",
  "16" = "Neural",
  "17" = "Cardiomyocytes",
  "18" = "Endothelial"
)

cell_type_broad_df <- data.frame(
  cell_type_broad = broad_annotations[as.character(seu$seurat_clusters)],
  row.names = colnames(seu)
)
seu <- AddMetaData(seu, metadata = cell_type_broad_df)
```

## Visualise Annotated Cell Types

With cell type annotations assigned, we now examine the results to verify our
annotations make biological sense and to understand the cellular composition
of the heart across developmental stages.

### Annotated UMAP

The UMAP visualisation with cell type labels provides the key output of our
annotation workflow. We show both detailed annotations (distinguishing fetal
from adult cardiomyocytes, atrial from ventricular) and broad categories.

```{r umap-annotated, fig.width=14, fig.height=6}
# Define colours for cell types (14 unique detailed types)
celltype_colors <- c(
    "Fetal CM" = "#E64B35",
    "Fibroblasts" = "#00A087",
    "Atrial CM" = "#3C5488",
    "Ventricular CM" = "#4DBBD5",
    "Activated CM" = "#F4A582",
    "Pericytes" = "#F39B7F",
    "Endothelial" = "#8491B4",
    "Macrophages" = "#91D1C2",
    "Fetal Atrial CM" = "#B09C85",
    "Fetal Fibroblasts" = "#00CED1",
    "Neurons" = "#CD534C",
    "Immune" = "#868686",
    "Proliferating" = "#9370DB",
    "Young Endothelial" = "#7570B3"
)

# Broad categories
broad_colors <- c(
    "Cardiomyocytes" = "#E64B35",
    "Fibroblasts" = "#00A087",
    "Endothelial" = "#8491B4",
    "Pericytes" = "#F39B7F",
    "Immune" = "#91D1C2",
    "Neural" = "#CD534C",
    "Proliferating" = "#9370DB"
)

p1 <- DimPlot(seu, group.by = "cell_type", label = TRUE,
              label.size = 3, repel = TRUE) +
    scale_color_manual(values = celltype_colors) +
    ggtitle("Detailed Cell Types") +
    theme(legend.position = "none")

p2 <- DimPlot(seu, group.by = "cell_type_broad", label = TRUE,
              label.size = 4, repel = TRUE) +
    scale_color_manual(values = broad_colors) +
    ggtitle("Broad Cell Types")

p1 + p2
```

The annotated UMAP reveals a well-organised cellular landscape:

- **Cardiomyocytes** form the dominant population, with fetal cells (red tones)
  clustering separately from mature atrial and ventricular populations
- **Stromal cells** (fibroblasts, pericytes, endothelial) occupy distinct
  regions of the UMAP space
- **Immune and neural cells** form smaller but clearly defined clusters

### Cluster Size Distribution

Before examining developmental patterns, we assess how many cells belong to
each cluster. This helps interpret whether observed patterns reflect genuine
biology or sampling effects.

```{r cluster-sizes, fig.width=12, fig.height=5}
# Cluster sizes with cell type labels
cluster_size_data <- seu@meta.data %>%
    group_by(seurat_clusters, cell_type) %>%
    summarise(n = n(), .groups = "drop")

ggplot(cluster_size_data, aes(x = seurat_clusters, y = n, fill = cell_type)) +
  geom_bar(stat = "identity") +
  scale_fill_manual(values = celltype_colors) +
  labs(x = "Cluster", y = "Number of Cells", fill = "Cell Type",
       title = "Cell Count per Cluster") +
  theme_minimal() +
  theme(legend.position = "right",
        legend.text = element_text(size = 8))
```

Cluster 0 (Fetal CM) is the largest population with ~2,400 cells, consistent
with fetal samples contributing the most cells. Clusters 16 (Neurons) and 18
(Young Endothelial) are among the smallest, representing specialised populations.

### Developmental Composition per Cluster

A critical question in developmental studies is: which cell populations are
stage-specific versus shared across development? This stacked bar chart shows
the fetal/young/adult breakdown within each cluster.

```{r dev-composition-per-cluster, fig.width=12, fig.height=6}
# Calculate developmental composition per cluster
cluster_dev_comp <- seu@meta.data %>%
    group_by(seurat_clusters, group) %>%
    summarise(n = n(), .groups = "drop") %>%
    group_by(seurat_clusters) %>%
    mutate(pct = 100 * n / sum(n))

ggplot(cluster_dev_comp, aes(x = seurat_clusters, y = pct, fill = group)) +
  geom_bar(stat = "identity", position = "stack") +
  scale_fill_manual(values = group_colors) +
  labs(x = "Cluster", y = "Percentage", fill = "Stage",
       title = "Developmental Stage Composition per Cluster") +
  theme_minimal() +
  theme(axis.text.x = element_text(size = 10))
```

This figure reveals striking developmental patterns:

- **Fetal-specific clusters** (0, 6, 8, 9, 12, 16, 17): These cardiomyocyte,
  fibroblast, and neural populations are almost exclusively fetal, reflecting
  the distinct transcriptional state of immature heart cells
- **Mature CM clusters** (2, 11, 15): Atrial and ventricular CM clusters contain
  cells from all stages, with cluster 2 enriched for young and adult cells
- **Activated CM** (7): Cardiomyocytes with MHC class II expression, predominantly
  from young samples, potentially reflecting immune activation
- **Young-enriched endothelial** (18): Endothelial cells exclusively from young
  samples
- **Shared populations** (1, 3, 4, 5, 13, 14): Fibroblasts, pericytes, immune,
  and endothelial cells are present across all developmental stages
- **Proliferating cells** (10): Nearly exclusively fetal, consistent with the
  high proliferative capacity of fetal cardiac cells

### Cell Type Composition by Developmental Stage

We can also view this relationship from the opposite perspective: what is the
cellular composition within each developmental stage?

```{r celltype-by-stage, fig.width=10, fig.height=6}
# Calculate composition
stage_comp <- seu@meta.data %>%
    group_by(group, cell_type_broad) %>%
    summarise(n = n(), .groups = "drop") %>%
    group_by(group) %>%
    mutate(pct = 100 * n / sum(n))

ggplot(stage_comp, aes(x = group, y = pct, fill = cell_type_broad)) +
    geom_bar(stat = "identity", position = "stack") +
    scale_fill_manual(values = broad_colors) +
    labs(x = "Developmental Stage", y = "Percentage", fill = "Cell Type",
         title = "Cell Type Composition by Developmental Stage") +
    theme_minimal()
```

This complementary view shows:

- **Cardiomyocytes** dominate all stages but show distinct subtypes (fetal vs mature)
- **Proliferating cells** are predominantly fetal, consistent with the known
  proliferative capacity of fetal cardiomyocytes
- **Stromal populations** (fibroblasts, endothelial, pericytes) are relatively
  stable across development

### Split UMAP by Developmental Stage

Finally, we split the UMAP by developmental stage to directly visualise which
cell types are present at each time point. This view clearly shows how the
cellular landscape changes from fetal to adult heart.

```{r split-umap, fig.width=14, fig.height=5}
DimPlot(seu, group.by = "cell_type", split.by = "group",
        label = FALSE, cols = celltype_colors) +
  ggtitle("Cell Types by Developmental Stage") +
  theme(legend.position = "right",
        legend.text = element_text(size = 8))
```

The split view highlights several key observations:

- **Fetal heart** is dominated by Fetal CM populations, with proliferating
  cells visible (cluster 10)
- **Young heart** shows a transition state with both mature atrial/ventricular
  CM and some residual fetal-like populations
- **Adult heart** shows predominantly mature cardiomyocyte subtypes (Atrial
  and Ventricular CM)
- **Stromal and immune populations** are consistent across all stages

## Save Annotated Object

```{r save-checkpoint}
# Create results directory if it doesn't exist
dir.create("../results", recursive = TRUE, showWarnings = FALSE)

# Save the annotated Seurat object
saveRDS(seu, "../data/processed/03_annotated.rds")

# Also save the markers table
write.csv(markers, "../results/cluster_markers.csv", row.names = FALSE)

message("Saved annotated object to: data/processed/03_annotated.rds")
message("Saved marker genes to: results/cluster_markers.csv")
```

## Summary

In this module, we performed cell type annotation on the clustered data. We:

- Identified marker genes for each cluster using `FindAllMarkers()`
- Compared cluster markers against known heart cell type markers from Sim et al. 2021
- Visualised marker expression using DotPlots and FeaturePlots
- Assigned cell type labels based on marker gene expression and developmental
  composition
- Created both detailed (14 types) and broad (7 types) cell type annotations

The annotated dataset is now ready for differential expression and composition
analysis in Module 4.

## Session Information

```{r session-info}
sessionInfo()
```
