---
title: "Module 5: Continuous Phenotyping with Φ-Space"
author: "Single Cell Workshop"
date: "`r Sys.Date()`"
output:
  html_document:
    theme: flatly
    toc: true
    toc_depth: 2
    toc_float: true
    number_sections: true
    code_folding: show
    df_print: paged
vignette: >
  %\VignetteIndexEntry{Module 5: Continuous Phenotyping with Φ-Space}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
    echo = TRUE,
    warning = FALSE,
    message = FALSE,
    fig.align = "center",
    fig.width = 10,
    fig.height = 6
)
```

## Introduction

In Module 3 we annotated cell types by looking up **marker genes** and assigning
each cluster a single discrete label. That approach is intuitive and directly
interpretable, but it rests on two assumptions that are not always true for the
heart maturation dataset we are analysing:

1. **Cells belong to one cell type.** A cluster is given a single label, and
   every cell inside it inherits that label.
2. **Cell identity is discrete.** Cells are either cardiomyocytes or they are
   not; there is no notion of a cell that is *partly* a fetal cardiomyocyte and
   *partly* an adult one.

For a mature tissue with well-separated cell types, these assumptions are
reasonable. But our dataset spans foetal, young and adult heart samples, so
many cells — especially cardiomyocytes — are in the middle of a maturation
continuum. A discrete label forces each cell into one bin and throws away the
very gradient we want to study in Module 6 (pseudotime) and Module 7 (gene
regulatory networks).

In this module we use **Φ-Space**
([Mao, Deng & Lê Cao, 2025, *Genome Biology*](https://doi.org/10.1186/s13059-025-03755-8))
to re-annotate the heart data with *continuous* phenotype scores instead of
hard labels.

### From hard labels to soft scores

The core idea of Φ-Space is simple: instead of asking *"which cell type is this
cell?"*, ask *"how much does this cell resemble each reference phenotype?"*.
For every query cell and every phenotype in the reference atlas, Φ-Space
returns a real-valued score. A cell that sits halfway between "foetal
cardiomyocyte" and "adult cardiomyocyte" will score positively on both; a cell
committed firmly to one identity will score high on one and low on the other.

Under the hood, Φ-Space fits a **partial least squares (PLS)** regression from
gene expression in the reference onto a dummy matrix of phenotype labels. The
fitted model is then applied to the query, producing one score per cell per
phenotype. This matrix of scores — the **phenotype space embedding** — is
stored as a `reducedDim` and becomes the substrate for every downstream
analysis: visualisation, clustering, marker selection, and trajectory
inference.

### Why Φ-Space is a good fit for this dataset

Five features of Φ-Space line up unusually well with the Sim et al. heart data:

1. **Continuous biology deserves continuous annotation.** Cardiomyocyte
   maturation is gradual. Φ-Space preserves the gradient; hard classification
   collapses it.
2. **Multiple phenotype layers at once.** We can score each cell simultaneously
   against cell type *and* developmental stage (and sex, the original study's
   focus). Most tools predict one label at a time.
3. **Bulk references are allowed.** There is no single cross-age
   snRNA-seq atlas of the human heart. Φ-Space accepts bulk RNA-seq
   references, which opens up developmental bulk time-course atlases as viable
   options.
4. **No extra batch correction between reference and query.** PLS inherently
   removes unwanted variation. You do not need to run Harmony *between* query
   and reference — only to match normalisation.
5. **The embedding feeds directly into Modules 6 and 7.** Because scores are
   stored in `reducedDim`, they can be used as input for slingshot (pseudotime)
   and NeighbourNet (gene regulatory networks) without any extra wrangling.

### Learning objectives

By the end of this module you will be able to:

- Explain the difference between hard and soft classification for cell type
  annotation, and identify situations where each is appropriate.
- Prepare an annotated reference dataset and a query dataset for Φ-Space,
  including matched normalisation and gene name harmonisation.
- Run `PhiSpace()` to obtain a continuous phenotype space embedding for the
  query cells.
- Visualise and interpret the phenotype space embedding as a heatmap, a UMAP,
  and a set of per-phenotype score maps.
- Derive discrete labels from continuous scores when needed, and understand
  what is gained and lost in that step.
- Export the phenotype space embedding in a form that Modules 6 and 7 can
  consume.

## Load Libraries and Data

```{r load-libraries, eval=FALSE}
# TODO: finalise once reference dataset and Seurat↔SCE conversion path are fixed
library(PhiSpace)
library(SingleCellExperiment)
library(Seurat)
library(ggplot2)
library(dplyr)
library(patchwork)
```

```{r load-data, eval=FALSE}
# TODO: load the processed Seurat object produced by Module 4
# seu <- readRDS("../data/processed/04_annotated.rds")
# query_sce <- as.SingleCellExperiment(seu)
```

## Choosing a Reference Dataset

<!-- TODO:
  - Criteria for a good reference for this dataset:
      * cross-age coverage (foetal through adult heart)
      * annotated cell types at a useful granularity
      * publicly available
  - Candidate references to discuss with instructor:
      * Litviňuková et al. 2020 (Nature) — adult heart cell atlas
      * Kanemaru et al. 2023 (Nature) — adult human heart
      * Developmental heart bulk RNA-seq time course (TBD)
  - Decide: single reference or multi-reference list?
-->

## Preparing Reference and Query

<!-- TODO:
  - Match normalisation: RankTransf() / logTransf() / scranTransf()
  - KeepCommonGenes() between reference and query
  - zeroFeatQC() on query
  - subsample() the reference for balance and speed
  - Convert Seurat ↔ SCE as required
-->

## Running Φ-Space

<!-- TODO:
  - PhiSpace() call with phenotypes = c("cell_type", "stage") if multi-layer
  - Show how to access reducedDim(query_sce, "PhiSpace")
  - Discuss ncomp default and when to tune with tunePhiSpace()
-->

## Exploring the Phenotype Space

<!-- TODO:
  - Heatmap of scores grouped by query cluster / morning-module cell type
  - PCA on reference scores, project query → 2D "meta-phenotype" plot
  - UMAP on phenotype space embedding
  - Per-phenotype score UMAPs for a few highlighted phenotypes
    (e.g. adult cardiomyocyte, foetal cardiomyocyte, fibroblast)
-->

## Hard vs Soft Classification

<!-- TODO:
  - getClass() to derive a single label per cell
  - Compare to Module 3 marker-based labels in a confusion table
  - Discuss which cells disagree and why — teaching moment about
    transitional states
-->

## Handing Off to Module 6

<!-- TODO:
  - Save updated SCE / Seurat object with PhiSpace embedding in reducedDim
  - Recap: cardiomyocytes are the subset we'll carry into slingshot
  - Brief preview of pseudotime
-->

## Summary

<!-- TODO: bullet recap of what we did and what stays with us into Modules 6-7 -->

## Session Info

```{r session-info, eval=FALSE}
sessionInfo()
```
