---
title: "Module 5: Continuous Phenotyping with Φ-Space"
author: "Single Cell Workshop"
date: "`r Sys.Date()`"
output:
  html_document:
    theme: flatly
    toc: true
    toc_depth: 2
    toc_float: true
    number_sections: true
    code_folding: show
    df_print: paged
vignette: >
  %\VignetteIndexEntry{Module 5: Continuous Phenotyping with Φ-Space}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
    echo = TRUE,
    warning = FALSE,
    message = FALSE,
    fig.align = "center",
    fig.width = 10,
    fig.height = 6
)
```

## Introduction

In Module 3 we annotated cell types by looking up **marker genes** and assigning
each cluster a single discrete label. That approach is intuitive and directly
interpretable, but it rests on two assumptions that are not always true for the
heart maturation dataset we are analysing:

1. **Cells belong to one cell type.** A cluster is given a single label, and
   every cell inside it inherits that label.
2. **Cell identity is discrete.** Cells are either cardiomyocytes or they are
   not; there is no notion of a cell that is *partly* a fetal cardiomyocyte and
   *partly* an adult one.

For a mature tissue with well-separated cell types, these assumptions are
reasonable. But our dataset spans foetal, young and adult heart samples, so
many cells — especially cardiomyocytes — are in the middle of a maturation
continuum. A discrete label forces each cell into one bin and throws away the
very gradient we want to study in Module 6 (pseudotime) and Module 7 (gene
regulatory networks).

In this module we use **Φ-Space**
([Mao, Deng & Lê Cao, 2025, *Genome Biology*](https://doi.org/10.1186/s13059-025-03755-8))
to re-annotate the heart data with *continuous* phenotype scores instead of
hard labels.

### From hard labels to soft scores

The core idea of Φ-Space is simple: instead of asking *"which cell type is this
cell?"*, ask *"how much does this cell resemble each reference phenotype?"*.
For every query cell and every phenotype in the reference atlas, Φ-Space
returns a real-valued score. A cell that sits halfway between "foetal
cardiomyocyte" and "adult cardiomyocyte" will score positively on both; a cell
committed firmly to one identity will score high on one and low on the other.

Under the hood, Φ-Space fits a **partial least squares (PLS)** regression from
gene expression in the reference onto a dummy matrix of phenotype labels. The
fitted model is then applied to the query, producing one score per cell per
phenotype. This matrix of scores — the **phenotype space embedding** — is
stored as a `reducedDim` and becomes the substrate for every downstream
analysis: visualisation, clustering, marker selection, and trajectory
inference.

### Why Φ-Space is a good fit for this dataset

Five features of Φ-Space line up unusually well with the Sim et al. heart data:

1. **Continuous biology deserves continuous annotation.** Cardiomyocyte
   maturation is gradual. Φ-Space preserves the gradient; hard classification
   collapses it.
2. **Multiple phenotype layers at once.** We can score each cell simultaneously
   against cell type *and* developmental stage. Most tools predict one label
   at a time.
3. **Bulk references are allowed.** There is no single cross-age snRNA-seq
   atlas of the human heart that covers fetal through adult equally well.
   Φ-Space also accepts bulk RNA-seq references, which opens up developmental
   bulk time-course atlases as viable options.
4. **No extra batch correction between reference and query.** PLS inherently
   removes unwanted variation. You do not need to run Harmony *between* query
   and reference — only to match normalisation.
5. **The embedding feeds directly into Modules 6 and 7.** Because scores are
   stored in `reducedDim`, they can be used as input for slingshot (pseudotime)
   and NeighbourNet (gene regulatory networks) without any extra wrangling.

### Learning objectives

By the end of this module you will be able to:

- Explain the difference between hard and soft classification for cell type
  annotation, and identify situations where each is appropriate.
- Prepare an annotated reference dataset and a query dataset for Φ-Space,
  including matched normalisation and gene name harmonisation.
- Run `PhiSpace()` to obtain a continuous phenotype space embedding for the
  query cells.
- Visualise and interpret the phenotype space embedding through heatmaps and a
  PCA biplot.
- Derive discrete labels from continuous scores when needed, and understand
  what is gained and lost in that step.
- Export the phenotype space embedding in a form that Modules 6 and 7 can
  consume.

## Load Libraries and Data

```{r load-libraries}
library(PhiSpace)
library(SingleCellExperiment)
library(Seurat)
library(ggplot2)
library(dplyr)
library(tibble)
library(patchwork)
library(ComplexHeatmap)
```

We need two data objects:

- a **query** — the heart snRNA-seq data from the morning session, trimmed to
  the essentials (raw counts, sample / stage / sex metadata, ~10,000 cells).
- a **reference** — an annotated left-ventricle atlas covering the full
  developmental axis (see the next section for details).

```{r load-data}
seu <- readRDS("../data/afternoon_mvp.rds")
reference <- readRDS("../references/reference_gao2026.rds")

cat("Query (Seurat):\n")
cat("  - cells:", ncol(seu), "\n")
cat("  - genes:", nrow(seu), "\n")
cat("  - metadata:", paste(colnames(seu@meta.data), collapse = ", "), "\n\n")

cat("Reference (SingleCellExperiment):\n")
cat("  - cells:", ncol(reference), "\n")
cat("  - genes:", nrow(reference), "\n")
cat("  - assays:", paste(assayNames(reference), collapse = ", "), "\n")
cat("  - cell types:", length(unique(reference$final_cell_type)), "\n")
cat("  - age groups:", paste(unique(reference$age_group), collapse = ", "), "\n")
```

Φ-Space works with `SingleCellExperiment` (SCE) objects, so we convert the
query from Seurat to SCE. The raw counts travel across as the `counts` assay;
the cell-level metadata travels across as `colData`.

```{r seurat-to-sce}
query <- as.SingleCellExperiment(seu)
assayNames(query)
head(colData(query)[, c("sample", "group", "sex")])
```

We will also set up a cache folder for intermediate results. Computations that
take more than a few seconds are cached to `../results/` so that re-rendering
this vignette is fast, and so that workshop participants can download the
pre-computed caches directly if they prefer.

```{r cache-setup}
results_dir <- "../results"
dir.create(results_dir, showWarnings = FALSE, recursive = TRUE)
```

## Choosing a Reference Dataset

A Φ-Space annotation is only as good as the reference it is built on. For the
Sim et al. heart data we need a reference that:

- **covers the full developmental axis** — foetal, paediatric and adult heart,
  so that every query cell has something biologically sensible to score
  against;
- **annotates cell types at a useful granularity** — at least the major heart
  lineages (cardiomyocyte, fibroblast, endothelial, pericyte, immune,
  epicardial, neuronal);
- **is publicly available** and free of major disease confounders.

We use the **Gao et al. 2026** (*Genome Biology*) integrative snRNA-seq atlas
of the human left ventricle. It combines 299 donors from eight published
studies, spans 8 post-conceptional weeks through ≥60 years, and provides
harmonised cell type and age-group annotations. We have pre-filtered it to
non-diseased donors and balanced the major cell types by down-sampling; the
processing script lives at
[`references/prepare_gao_2026.R`](../references/prepare_gao_2026.R).

Two columns from the reference drive everything that follows:

```{r reference-metadata}
table(reference$final_cell_type)
table(reference$age_group)
```

Notice the reference's `age_group` is `fetal / young / middle / old`, while
our query's `group` is `fetal / young / adult`. That is fine — Φ-Space will
score each query cell against all four reference age groups, and we will use
those four scores as a fine-grained read-out of where on the maturation axis
each cell sits.

## Preparing Reference and Query

Three preparation steps are required before we can run Φ-Space:

1. **Match normalisation.** Φ-Space does not do batch correction between
   reference and query; it relies on both being on the same numerical scale.
2. **Match gene names.** PLS can only use genes present in both datasets.
3. **Quality control.** Remove all-zero genes from the query, which would
   otherwise carry no signal but still consume model capacity.

### Match normalisation with rank transformation

The reference is stored with a `logcounts` assay; the query carries raw
`counts`. Rather than hand-matching library-size normalisations across two
very different pipelines, we apply **rank transformation** to both. Rank
transformation replaces each cell's expression vector with the ranks of its
genes, making the result invariant to library size, scale, and monotone
transformations — exactly what we need when matching across platforms.

```{r rank-transform}
reference <- RankTransf(reference, assayname = "logcounts")
query     <- RankTransf(query,     assayname = "counts")
assayNames(reference)
assayNames(query)
```

Both objects now carry a `rank` assay, and that is what Φ-Space will use.

### Keep common genes and remove zero-expression genes

```{r common-genes}
common <- intersect(rownames(reference), rownames(query))
cat("Reference genes:", nrow(reference), "\n")
cat("Query genes:    ", nrow(query), "\n")
cat("Common:         ", length(common), "\n")

reference <- reference[common, ]
query     <- query[common, ]
query     <- zeroFeatQC(query)
dim(query)
```

The gene-symbol overlap is comfortably above ten thousand, which is plenty for
PLS.

### Balancing the reference

A good reference is balanced across cell types — if 90 % of the reference cells
are cardiomyocytes, the model will over-learn cardiomyocyte variation and
under-learn everything else. Our Gao 2026 reference has already been
down-sampled per cell type during processing (see
[`prepare_gao_2026.R`](../references/prepare_gao_2026.R)), with each rare cell
type capped at 100 cells and abundant types capped at a few thousand. If you
ever work with a less curated reference, `PhiSpace::subsample()` is the
function you want — it draws a stratified subset by any `colData` column.

## Running Φ-Space

We now fit the PLS model on the reference and project the query into the
phenotype space. We ask for two phenotype layers at once:

- `final_cell_type` (13 levels) — what kind of cell is this?
- `age_group` (4 levels) — where on the maturation axis does it sit?

The resulting embedding will have 13 + 4 = 17 columns: one score per query
cell per phenotype level.

This step is the computational bottleneck of the module (roughly one minute
on a laptop), so we cache the output.

```{r run-phispace}
phispace_cache <- file.path(results_dir, "05_phispace_query.rds")

if (file.exists(phispace_cache)) {
    query <- readRDS(phispace_cache)
    message("Loaded cached PhiSpace result from ", phispace_cache)
} else {
    query <- PhiSpace(
        reference  = reference,
        query      = query,
        phenotypes = c("final_cell_type", "age_group"),
        refAssay   = "rank",
        regMethod  = "PLS",
        center     = TRUE,
        scale      = FALSE
    )
    # Drop the intermediate rank assay before caching — it is no longer
    # needed downstream and roughly halves the file size.
    assay(query, "rank") <- NULL
    saveRDS(query, phispace_cache)
}
```

The phenotype-space embedding is stored as a `reducedDim`:

```{r inspect-scores}
phi_scores <- reducedDim(query, "PhiSpace")
dim(phi_scores)
colnames(phi_scores)
round(phi_scores[1:4, 1:6], 3)
```

Each row is a query cell, each column is a reference phenotype level. Values
are column-wise normalised so that the median of each column is zero and the
maximum absolute deviation is one — high positive values mean the cell
resembles that phenotype, values near zero mean average, and negative values
mean the cell is confidently *not* that phenotype.

### On the number of components

Φ-Space defaults `ncomp` to the total number of phenotype levels (here 17).
This is a sensible starting point that we have found works well in practice.
For a more principled choice — especially with large references — you can
cross-validate `ncomp` and feature count with `tunePhiSpace()`; we skip it
here in the interest of time, but the function is a one-liner when you need
it.

## Exploring the Phenotype Space

### Heatmap of scores grouped by developmental stage

A good first look is a heatmap with query cells as rows (grouped by
developmental stage) and reference phenotypes as columns. The expectation is
clear: cells from the foetal group should score high on foetal-related
reference phenotypes, and similarly for young and adult.

```{r heatmap-scores, fig.height=7}
split_phenotype <- c(
    rep("cell type", length(unique(reference$final_cell_type))),
    rep("age group", length(unique(reference$age_group)))
)

ht <- Heatmap(
    phi_scores,
    name             = "Φ-Space score",
    row_split        = query$group,
    column_split     = split_phenotype,
    cluster_row_slices = FALSE,
    show_row_names   = FALSE,
    show_row_dend    = FALSE,
    show_column_dend = FALSE,
    column_names_rot = 45,
    row_title_rot    = 0
)
draw(ht, heatmap_legend_side = "right")
```

Two patterns should jump out:

- Within the `age group` columns, the diagonal block structure tells you that
  Φ-Space has correctly recovered the developmental stage of each query cell
  without ever being told what stage it was from.
- Within the `cell type` columns, the dominant column for most cells is
  `Cardiomyocyte`, as expected for a heart sample — but a subset of cells
  score high on `Fibroblast`, `Endothelial`, `Myeloid`, etc.

### PCA biplot of the phenotype space

The phenotype-space embedding is 17-dimensional. To get a two-dimensional view
that still reflects the geometry of the embedding, we apply PCA to the score
matrix. Φ-Space ships a `getPC()` helper that returns scores, loadings and
variance-explained in one call.

```{r phispace-pca}
pc_res <- getPC(phi_scores, ncomp = 5, center = TRUE, scale = FALSE)
round(pc_res$props * 100, 1)
```

PC1 and PC2 together explain a large fraction of the variance in phenotype
space. We now plot the embedding as a **biplot**: one panel shows the query
cells (the `scores`), the other panel shows the reference phenotypes (the
`loadings`) in the same 2D basis. Reading the two panels together tells you
*why* cells sit where they do — a cell in the positive-PC1 / positive-PC2
quadrant is being pulled there by the phenotypes that load positively on both.

```{r biplot, fig.width=14, fig.height=6}
scores_df <- as.data.frame(pc_res$scores[, 1:2])
colnames(scores_df) <- c("PC1", "PC2")
scores_df$group <- query$group

loadings_df <- as.data.frame(pc_res$loadings[, 1:2])
colnames(loadings_df) <- c("PC1", "PC2")
loadings_df$phenotype <- rownames(loadings_df)
loadings_df$layer <- ifelse(
    loadings_df$phenotype %in% unique(reference$age_group),
    "age group", "cell type"
)

pc1_lab <- sprintf("PC1 (%.1f%%)", pc_res$props[1] * 100)
pc2_lab <- sprintf("PC2 (%.1f%%)", pc_res$props[2] * 100)

p_scores <- ggplot(scores_df, aes(PC1, PC2, colour = group)) +
    geom_point(size = 0.6, alpha = 0.7) +
    scale_colour_brewer(palette = "Set1") +
    labs(title = "Query cells", x = pc1_lab, y = pc2_lab, colour = "Stage") +
    theme_bw(base_size = 13) +
    theme(legend.position = "bottom") +
    guides(colour = guide_legend(override.aes = list(size = 3, alpha = 1)))

p_loadings <- ggplot(loadings_df, aes(PC1, PC2, colour = layer)) +
    geom_segment(aes(x = 0, y = 0, xend = PC1, yend = PC2),
                 arrow = arrow(length = unit(0.015, "npc")), alpha = 0.5) +
    geom_text(aes(label = phenotype), size = 3.5, vjust = -0.4) +
    scale_colour_manual(values = c("age group" = "#d95f02", "cell type" = "#1b9e77")) +
    labs(title = "Reference phenotypes (loadings)",
         x = pc1_lab, y = pc2_lab, colour = "Phenotype layer") +
    theme_bw(base_size = 13) +
    theme(legend.position = "bottom")

p_scores + p_loadings
```

This biplot is the single most useful visualisation in Φ-Space. Read it like
this:

- **Query cells cluster by stage.** The three stage colours should separate
  along the PC that is loaded most heavily by `fetal`, `young`, `middle`,
  `old` — that is the maturation axis we will hand off to slingshot in
  Module 6.
- **Cell-type phenotypes point to where each lineage lives.** A query cell
  near the `Cardiomyocyte` arrow is a cardiomyocyte; one near `Fibroblast`
  is a fibroblast. Cells in between the two arrows are transitional /
  ambiguous and would be assigned inconsistently by a hard classifier.
- **The foetal→adult axis is visible even within cardiomyocytes.** Look at
  where the `Cardiomyocyte` loading sits relative to `fetal` and `old` —
  query cardiomyocytes should spread along that direction, which is precisely
  the trajectory Module 6 will trace.

## Hard vs Soft Classification

Sometimes you genuinely do need a single label per cell — for example, to
tabulate cell-type composition per donor. Φ-Space provides `getClass()` for
exactly this: it picks the column with the highest score for each cell.

```{r hard-labels}
cell_type_cols <- unique(reference$final_cell_type)
query$PhiCellType <- getClass(phi_scores[, cell_type_cols])

age_cols <- unique(reference$age_group)
query$PhiAge <- getClass(phi_scores[, age_cols])

table(query$PhiCellType)
table(query$group, query$PhiAge)
```

The `group` × `PhiAge` cross-table is a useful sanity check: most foetal
query cells should land in the `fetal` reference bin, and most adult query
cells in `old` — but the off-diagonal entries are where the teaching is.
Those cells are the ones whose continuous scores straddle two age groups.

Let's look at the cardiomyocytes more closely — they are the main interest
for Module 6:

```{r cm-score-distribution, fig.width=10, fig.height=4}
cm_scores <- phi_scores[query$PhiCellType == "Cardiomyocyte", age_cols, drop = FALSE]
cm_df <- as.data.frame(cm_scores) %>%
    tibble::rownames_to_column("cell") %>%
    mutate(group = query$group[match(cell, colnames(query))]) %>%
    tidyr::pivot_longer(all_of(age_cols), names_to = "age_phenotype",
                        values_to = "score") %>%
    mutate(age_phenotype = factor(age_phenotype,
                                  levels = c("fetal", "young", "middle", "old")))

ggplot(cm_df, aes(age_phenotype, score, fill = group)) +
    geom_violin(scale = "width", alpha = 0.8) +
    scale_fill_brewer(palette = "Set1") +
    labs(x = "Reference age group", y = "Φ-Space score",
         fill = "Query stage",
         title = "Age-group scores for query cardiomyocytes") +
    theme_bw(base_size = 13)
```

Notice that even within a single query stage there is substantial spread
along the age-group axis. A hard classifier hides this spread; the continuous
scores are exactly what slingshot will exploit to trace a maturation
trajectory.

## Handing Off to Module 6

We save the annotated query object so that Module 6 can pick up where we
leave off.

```{r save-for-module-6}
final_cache <- file.path(results_dir, "05_phispace_query.rds")
if (!file.exists(final_cache)) saveRDS(query, final_cache)
cat("Annotated query at", final_cache, "\n")
```

The object now carries:

- `reducedDim(query, "PhiSpace")` — the 17-column score matrix.
- `colData(query)$PhiCellType` — discrete cell-type label.
- `colData(query)$PhiAge` — discrete maturation-stage label.
- all original metadata (`sample`, `group`, `sex`, QC metrics).

In Module 6 we will **subset this object to cardiomyocytes** (using
`PhiCellType == "Cardiomyocyte"`) and apply **slingshot** to the PhiSpace
score matrix. The continuous age-group scores will act as the biological
anchor that tells slingshot which end of the trajectory is foetal and which
end is adult.

## Summary

- Φ-Space replaces a hard cell-type label with a continuous score per
  reference phenotype per cell — a natural fit for a dataset that spans a
  developmental gradient.
- The workflow is four steps: match normalisation (we used rank
  transformation on both reference and query), keep common genes, run
  `PhiSpace()` with one or more phenotype layers, inspect the scores.
- Continuous scores can always be collapsed to a discrete label with
  `getClass()`, but the inverse is not possible — preserve the scores for
  any downstream analysis that benefits from soft assignments (trajectory
  inference, co-phenotype analysis, gradient visualisation).
- The phenotype-space embedding we saved is the input to Module 6
  (pseudotime) and Module 7 (gene regulatory networks).

## Session Info

```{r session-info}
sessionInfo()
```
