---
title: "Module 5: Continuous Phenotyping with Φ-Space"
author: "Single Cell Workshop"
date: "`r Sys.Date()`"
output:
  html_document:
    theme: flatly
    toc: true
    toc_depth: 2
    toc_float: true
    number_sections: true
    code_folding: show
    df_print: paged
vignette: >
  %\VignetteIndexEntry{Module 5: Continuous Phenotyping with Φ-Space}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
# Module 5 needs two data files that live outside the default workshop
# download: the heart query object (`afternoon_mvp.rds`) and the Gao 2026
# reference (`reference_gao2026.rds`). If either is missing (e.g. on a
# fresh clone or in CI before the Zenodo upload), skip execution but still
# render the prose so the site build succeeds.
data_ok <- all(file.exists(
    "../data/afternoon_mvp.rds",
    "../data/reference_gao2026.rds"
))

knitr::opts_chunk$set(
    echo = TRUE,
    warning = FALSE,
    message = FALSE,
    fig.align = "center",
    fig.width = 10,
    fig.height = 6,
    eval = data_ok
)
```

```{asis, echo = !data_ok}
> **Note.** This module's code chunks are not evaluated because
> `data/afternoon_mvp.rds` and/or `data/reference_gao2026.rds` are not
> present. Download them from the workshop's Zenodo record (see Module 0)
> and re-render to reproduce the analysis.
```

## Introduction

In Module 3 we annotated cell types by looking up **marker genes** and assigning
each cluster a single discrete label. That approach is intuitive and directly
interpretable, but it rests on two assumptions that are not always true for the
heart maturation dataset we are analysing:

1. **Cells belong to one cell type.** A cluster is given a single label, and
   every cell inside it inherits that label.
2. **Cell identity is discrete.** Cells are either cardiomyocytes or they are
   not; there is no notion of a cell that is *partly* a fetal cardiomyocyte and
   *partly* an adult one.

For a mature tissue with well-separated cell types, these assumptions are
reasonable. But our dataset spans foetal, young and adult heart samples, so
many cells — especially cardiomyocytes — are in the middle of a maturation
continuum. A discrete label forces each cell into one bin and throws away the
very gradient we want to study in Module 6 (pseudotime) and Module 7
(cell-specific co-expression networks).

In this module we use **Φ-Space**
([Mao, Deng & Lê Cao, 2025, *Genome Biology*](https://doi.org/10.1186/s13059-025-03755-8))
to re-annotate the heart data with *continuous* phenotype scores instead of
hard labels.

### From hard labels to soft scores

The core idea of Φ-Space is simple: instead of asking *"which cell type is this
cell?"*, ask *"how much does this cell resemble each reference phenotype?"*.
For every query cell and every phenotype in the reference atlas, Φ-Space
returns a real-valued score. A cell that sits halfway between "foetal
cardiomyocyte" and "adult cardiomyocyte" will score positively on both; a cell
committed firmly to one identity will score high on one and low on the other.

Under the hood, Φ-Space fits a **partial least squares (PLS)** regression from
gene expression in the reference onto a dummy matrix of phenotype labels. The
fitted model is then applied to the query, producing one score per cell per
phenotype. This matrix of scores — the **phenotype space embedding** — is
stored as a `reducedDim` and becomes the substrate for every downstream
analysis: visualisation, clustering, marker selection, and trajectory
inference.

### Why Φ-Space is a good fit for this dataset

Five features of Φ-Space line up unusually well with the Sim et al. heart data:

1. **Continuous biology deserves continuous annotation.** Cardiomyocyte
   maturation is gradual. Φ-Space preserves the gradient; hard classification
   collapses it.
2. **Multiple phenotype layers at once.** We can score each cell simultaneously
   against cell type *and* developmental stage. Most tools predict one label
   at a time.
3. **Bulk references are allowed.** There is no single cross-age snRNA-seq
   atlas of the human heart that covers fetal through adult equally well.
   Φ-Space also accepts bulk RNA-seq references, which opens up developmental
   bulk time-course atlases as viable options.
4. **No extra batch correction between reference and query.** PLS inherently
   removes unwanted variation. You do not need to run Harmony *between* query
   and reference — only to match normalisation.
5. **The embedding feeds directly into Modules 6 and 7.** Because scores are
   stored in `reducedDim`, they can be used as input for slingshot
   (pseudotime) and NeighbourNet target-gene selection without any extra
   wrangling.

### Learning objectives

By the end of this module you will be able to:

- Explain the difference between hard and soft classification for cell type
  annotation, and identify situations where each is appropriate.
- Prepare an annotated reference dataset and a query dataset for Φ-Space,
  including matched normalisation and gene name harmonisation.
- Run `PhiSpace()` to obtain a continuous phenotype space embedding for the
  query cells.
- Visualise and interpret the phenotype space embedding through heatmaps and a
  PCA biplot.
- Derive discrete labels from continuous scores when needed, and understand
  what is gained and lost in that step.
- Export the phenotype space embedding in a form that Modules 6 and 7 can
  consume.

## Load Libraries and Data

```{r load-libraries}
library(PhiSpace)
library(SingleCellExperiment)
library(Seurat)
library(ggplot2)
library(dplyr)
library(tibble)
library(patchwork)
library(ComplexHeatmap)
```

We need two data objects:

- a **query** — the heart snRNA-seq data from the morning session, trimmed to
  the essentials (raw counts, sample / stage / sex metadata, ~10,000 cells).
- a **reference** — an annotated left-ventricle atlas covering the full
  developmental axis (see the next section for details).

```{r load-data}
seu <- readRDS("../data/afternoon_mvp.rds")
reference <- readRDS("../data/reference_gao2026.rds")

cat("Query (Seurat):\n")
cat("  - cells:", ncol(seu), "\n")
cat("  - genes:", nrow(seu), "\n")
cat("  - metadata:", paste(colnames(seu@meta.data), collapse = ", "), "\n\n")

cat("Reference (SingleCellExperiment):\n")
cat("  - cells:", ncol(reference), "\n")
cat("  - genes:", nrow(reference), "\n")
cat("  - assays:", paste(assayNames(reference), collapse = ", "), "\n")
cat("  - cell types:", length(unique(reference$final_cell_type)), "\n")
cat("  - age groups:", paste(unique(reference$age_group), collapse = ", "), "\n")
```

Φ-Space works with `SingleCellExperiment` (SCE) objects, so we convert the
query from Seurat to SCE. The raw counts travel across as the `counts` assay;
the cell-level metadata travels across as `colData`.

```{r seurat-to-sce}
query <- as.SingleCellExperiment(seu)
assayNames(query)
head(colData(query)[, c("sample", "group", "sex")])
```

We will also set up a cache folder for intermediate results. Computations that
take more than a few seconds are cached to `../results/` so that re-rendering
this vignette is fast, and so that workshop participants can download the
pre-computed caches directly if they prefer.

```{r cache-setup}
results_dir <- "../results"
dir.create(results_dir, showWarnings = FALSE, recursive = TRUE)
```

## Choosing a Reference Dataset

A Φ-Space annotation is only as good as the reference it is built on. For the
Sim et al. heart data we need a reference that:

- **covers the full developmental axis** — foetal, paediatric and adult heart,
  so that every query cell has something biologically sensible to score
  against;
- **annotates cell types at a useful granularity** — at least the major heart
  lineages (cardiomyocyte, fibroblast, endothelial, pericyte, immune,
  epicardial, neuronal);
- **is publicly available** and free of major disease confounders.

We use the **Gao et al. 2026** (*Genome Biology*) integrative snRNA-seq atlas
of the human left ventricle. It combines 299 donors from eight published
studies, spans 8 post-conceptional weeks through ≥60 years, and provides
harmonised cell type and age-group annotations. We have pre-filtered it to
non-diseased donors and balanced the major cell types by down-sampling; the
processing script lives at
[`references/prepare_gao_2026.R`](../references/prepare_gao_2026.R).

Two columns from the reference drive everything that follows:

```{r reference-metadata}
table(reference$final_cell_type)
table(reference$age_group)
```

Notice the reference's `age_group` is `fetal / young / middle / old`, while
our query's `group` is `fetal / young / adult`. That is fine — Φ-Space will
score each query cell against all four reference age groups, and we will use
those four scores as a fine-grained read-out of where on the maturation axis
each cell sits.

### Zooming in: splitting cardiomyocyte and fibroblast into subtypes

Looking at `final_cell_type` above, **Cardiomyocyte** (~1,900 cells) and
**Fibroblast** (~1,300 cells) dominate the reference. They are also the two
lineages we care about most biologically: cardiomyocytes carry the
foetal→adult maturation signal we will trace in Module 6, and fibroblasts
reshape the extracellular matrix across development. The Gao 2026 authors
further annotated these two lineages into biologically meaningful subclusters
(e.g. `Fetal_enriched_metabolic`, `Stressed`, `Interstitial`,
`Perivascular` …) — stored in the `cell_type_plus_subcluster` column as
`CellType:Subtype` strings.

We therefore build a **fine cell-type column** that keeps every non-CM,
non-Fibroblast cell on its broad label, and replaces the CM / Fibroblast
labels with the Gao subclusters. Two things happen at once:

1. **We zoom in on the cells we care about.** When a small number of cell
   types carry most of the biological signal, it is worth asking Φ-Space to
   resolve them more finely. Every CM subcluster becomes its own phenotype
   column in the score matrix, so we can tell a `Fetal_enriched_metabolic`
   CM from a `Stressed` CM rather than collapsing both into "Cardiomyocyte".
2. **The reference becomes more balanced.** Splitting the two giant classes
   into ~8 + ~7 smaller classes brings their per-class counts closer to the
   rarer lineages (Pericyte, Myeloid, Endothelial, …). PLS over-fits the
   majority class when the reference is unbalanced, and this split softens
   that skew for free.

```{r split-subtypes}
ct <- reference$final_cell_type
sub <- reference$cell_type_plus_subcluster
# Replace only the two populous lineages with their subclusters;
# leave the rest on their broad label.
reference$cell_type_fine <- ifelse(
    ct %in% c("Cardiomyocyte", "Fibroblast"),
    sub,
    ct
)

sort(table(reference$cell_type_fine), decreasing = TRUE)
```

Note the alternative we did **not** take: **balanced subsampling**. Instead
of splitting the majority classes, you could down-sample them with
`PhiSpace::subsample(..., key = "final_cell_type", minCellNum = ...)` so every
cell type contributes a comparable number of reference cells. That is the
right move when you are happy with the broad-class annotation and simply want
fairer PLS training; here we chose to split instead because the CM / Fibroblast
subclusters are interesting in their own right. The two approaches can also
be combined — split first, then subsample within each fine label.

### A note on very rare phenotypes

Even after splitting, some fine labels are tiny (e.g.
`Cardiomyocyte:Senescent` ≈ 4 cells, `Fibroblast:Inflammatory` ≈ 6 cells).
Training PLS with a phenotype represented by a handful of cells is a bad
bargain: the column it produces in the score matrix is mostly noise, and it
still consumes a component in the PLS fit.

`PhiSpace()` exposes a `cellTypeThreshold` argument for exactly this case —
set it to, say, `20`, and any phenotype level with fewer than 20 reference
cells is dropped before model fitting. We use it below to prune the long
tail of the subtype split while keeping every lineage that is biologically
well-represented.

## Preparing Reference and Query

Three preparation steps are required before we can run Φ-Space:

1. **Match normalisation.** Φ-Space does not do batch correction between
   reference and query; it relies on both being on the same numerical scale.
2. **Match gene names.** PLS can only use genes present in both datasets.
3. **Quality control.** Remove all-zero genes from the query, which would
   otherwise carry no signal but still consume model capacity.

### Match normalisation with scran

The reference is stored with a `logcounts` assay (scran-normalised during
reference preparation — see
[`HPC_reference_processing.md`](../references/HPC_reference_processing.md));
the query carries raw `counts`. To put both on the same numerical scale we
apply **scran normalisation** to the query as well, via PhiSpace's
`scranTransf()` wrapper. `scranTransf()` computes pool-based size factors
with `scran::computeSumFactors()` and then calls `scuttle::logNormCounts()`,
storing the result in a `logcounts` assay — the same assay name and the
same normalisation family the reference already uses.

An earlier draft of this module used rank transformation instead. Rank is
fast and invariant to library size, which is attractive when reference and
query come from very different platforms. But for this dataset both sides
are snRNA-seq, so the cross-platform argument is weaker, and scran's
size-factor normalisation preserves more of the count-level variation that
PLS can exploit. The trade-off is that scran is slower, so we cache the
intermediate object.

```{r scran-transform}
ref_cache   <- file.path(results_dir, "05_reference_scran.rds")
query_cache <- file.path(results_dir, "05_query_scran.rds")

if (file.exists(ref_cache)) {
    reference <- readRDS(ref_cache)
    message("Loaded cached scran-normalised reference from ", ref_cache)
} else {
    # Reference logcounts is already scran-normalised from the HPC prep
    # script — nothing to recompute, just cache for reproducibility.
    saveRDS(reference, ref_cache)
}

if (file.exists(query_cache)) {
    query <- readRDS(query_cache)
    message("Loaded cached scran-normalised query from ", query_cache)
} else {
    query <- scranTransf(query)
    saveRDS(query, query_cache)
}

assayNames(reference)
assayNames(query)
```

Both objects now carry a `logcounts` assay, and that is what Φ-Space will
use below (`refAssay = "logcounts"`).

### Keep common genes and remove zero-expression genes

```{r common-genes}
common <- intersect(rownames(reference), rownames(query))
cat("Reference genes:", nrow(reference), "\n")
cat("Query genes:    ", nrow(query), "\n")
cat("Common:         ", length(common), "\n")

reference <- reference[common, ]
query     <- query[common, ]
query     <- zeroFeatQC(query)
dim(query)
```

The gene-symbol overlap is comfortably above ten thousand, which is plenty for
PLS.

### Balancing the reference — a recap

Two levers keep the reference balanced: **splitting** dense majority classes
into biologically meaningful subclusters (what we did above with CM and
Fibroblast), and **subsampling** with `PhiSpace::subsample(key = ...)` to cap
the per-class cell count. The Gao 2026 reference was already down-sampled
during preparation (see
[`HPC_reference_processing.md`](../references/HPC_reference_processing.md)),
capping rare types at ~100 and abundant types at a few thousand — so between
that and our fine split, we are in a reasonable spot without further action.

## Running Φ-Space

We now fit the PLS model on the reference and project the query into the
phenotype space. We ask for two phenotype layers at once:

- `cell_type_fine` — the split label we built above (CM / Fibroblast
  replaced by their Gao subclusters, everything else on its broad label).
- `age_group` (4 levels) — where on the maturation axis does it sit?

We also pass `cellTypeThreshold = 20` so that phenotype levels represented
by fewer than 20 reference cells are dropped before PLS fitting — this
trims the long tail of rare CM / Fibroblast subclusters without touching
the well-populated ones.

This step is the computational bottleneck of the module (roughly one minute
on a laptop), so we cache the output.

```{r run-phispace}
phispace_cache <- file.path(results_dir, "05_phispace_query.rds")

if (file.exists(phispace_cache)) {
    query <- readRDS(phispace_cache)
    message("Loaded cached PhiSpace result from ", phispace_cache)
} else {
    query <- PhiSpace(
        reference         = reference,
        query             = query,
        phenotypes        = c("cell_type_fine", "age_group"),
        refAssay          = "logcounts",
        regMethod         = "PLS",
        center            = TRUE,
        scale             = FALSE,
        cellTypeThreshold = 20
    )
    # Keep logcounts in the cache — Module 6 needs it for PCA on the
    # expression side of its pseudotime comparison.
    saveRDS(query, phispace_cache)
}
```

The phenotype-space embedding is stored as a `reducedDim`:

```{r inspect-scores}
phi_scores <- reducedDim(query, "PhiSpace")
dim(phi_scores)
colnames(phi_scores)
round(phi_scores[1:4, 1:6], 3)
```

Each row is a query cell, each column is a reference phenotype level. Values
are column-wise normalised so that the median of each column is zero and the
maximum absolute deviation is one — high positive values mean the cell
resembles that phenotype, values near zero mean average, and negative values
mean the cell is confidently *not* that phenotype.

### On the number of components

Φ-Space defaults `ncomp` to the total number of phenotype levels (here 17).
This is a sensible starting point that we have found works well in practice.
For a more principled choice — especially with large references — you can
cross-validate `ncomp` and feature count with `tunePhiSpace()`; we skip it
here in the interest of time, but the function is a one-liner when you need
it.

## Exploring the Phenotype Space

### Heatmap of scores grouped by developmental stage

A good first look is a heatmap with query cells as rows (grouped by
developmental stage) and reference phenotypes as columns. The expectation is
clear: cells from the foetal group should score high on foetal-related
reference phenotypes, and similarly for young and adult.

```{r heatmap-scores, fig.height=7}
age_levels <- unique(reference$age_group)
split_phenotype <- ifelse(
    colnames(phi_scores) %in% age_levels, "age group", "cell type"
)

ht <- Heatmap(
    phi_scores,
    name             = "Φ-Space score",
    row_split        = query$group,
    column_split     = split_phenotype,
    cluster_row_slices = FALSE,
    show_row_names   = FALSE,
    show_row_dend    = FALSE,
    show_column_dend = FALSE,
    column_names_rot = 45,
    row_title_rot    = 0
)
draw(ht, heatmap_legend_side = "right")
```

Two patterns should jump out:

- Within the `age group` columns, the diagonal block structure tells you that
  Φ-Space has correctly recovered the developmental stage of each query cell
  without ever being told what stage it was from.
- Within the `cell type` columns, the dominant column for most cells is
  `Cardiomyocyte`, as expected for a heart sample — but a subset of cells
  score high on `Fibroblast`, `Endothelial`, `Myeloid`, etc.

### PCA biplot of the phenotype space

The phenotype-space embedding is 17-dimensional. To get a two-dimensional view
that still reflects the geometry of the embedding, we apply PCA to the score
matrix. Φ-Space ships a `getPC()` helper that returns scores, loadings and
variance-explained in one call.

```{r phispace-pca}
pc_res <- getPC(phi_scores, ncomp = 5, center = TRUE, scale = FALSE)
round(pc_res$props * 100, 1)
```

PC1 and PC2 together explain a large fraction of the variance in phenotype
space. We now plot the embedding as a **biplot**: one panel shows the query
cells (the `scores`), the other panel shows the reference phenotypes (the
`loadings`) in the same 2D basis. Reading the two panels together tells you
*why* cells sit where they do — a cell in the positive-PC1 / positive-PC2
quadrant is being pulled there by the phenotypes that load positively on both.

```{r biplot, fig.width=14, fig.height=6}
scores_df <- as.data.frame(pc_res$scores[, 1:2])
colnames(scores_df) <- c("PC1", "PC2")
scores_df$group <- query$group

loadings_df <- as.data.frame(pc_res$loadings[, 1:2])
colnames(loadings_df) <- c("PC1", "PC2")
loadings_df$phenotype <- rownames(loadings_df)
loadings_df$layer <- ifelse(
    loadings_df$phenotype %in% unique(reference$age_group),
    "age group", "cell type"
)

pc1_lab <- sprintf("PC1 (%.1f%%)", pc_res$props[1] * 100)
pc2_lab <- sprintf("PC2 (%.1f%%)", pc_res$props[2] * 100)

p_scores <- ggplot(scores_df, aes(PC1, PC2, colour = group)) +
    geom_point(size = 0.6, alpha = 0.7) +
    scale_colour_brewer(palette = "Set1") +
    labs(title = "Query cells", x = pc1_lab, y = pc2_lab, colour = "Stage") +
    theme_bw(base_size = 13) +
    theme(legend.position = "bottom") +
    guides(colour = guide_legend(override.aes = list(size = 3, alpha = 1)))

p_loadings <- ggplot(loadings_df, aes(PC1, PC2, colour = layer)) +
    geom_segment(aes(x = 0, y = 0, xend = PC1, yend = PC2),
                 arrow = arrow(length = unit(0.015, "npc")), alpha = 0.5) +
    geom_text(aes(label = phenotype), size = 3.5, vjust = -0.4) +
    scale_colour_manual(values = c("age group" = "#d95f02", "cell type" = "#1b9e77")) +
    labs(title = "Reference phenotypes (loadings)",
         x = pc1_lab, y = pc2_lab, colour = "Phenotype layer") +
    theme_bw(base_size = 13) +
    theme(legend.position = "bottom")

p_scores + p_loadings
```

This biplot is the single most useful visualisation in Φ-Space. Read it like
this:

- **Query cells cluster by stage.** The three stage colours should separate
  along the PC that is loaded most heavily by `fetal`, `young`, `middle`,
  `old` — that is the maturation axis we will hand off to slingshot in
  Module 6.
- **Cell-type phenotypes point to where each lineage lives.** A query cell
  near a `Cardiomyocyte:*` arrow is a cardiomyocyte; one near a `Fibroblast:*`
  arrow is a fibroblast. Cells in between the two arrows are transitional /
  ambiguous and would be assigned inconsistently by a hard classifier.
- **The foetal→adult axis is visible even within cardiomyocytes.** Look at
  where the `Cardiomyocyte:Fetal_enriched_*` loadings sit relative to the
  adult-stage CM subclusters — query cardiomyocytes should spread along
  that direction, which is precisely the trajectory Module 6 will trace.

### Query cells coloured by broad cell type

The stage panel tells us *where each donor group lands* in phenotype space,
but not *which lineage* each cell belongs to. To see that, we recolour the
same query PCA by the broad cell type derived from the fine Φ-Space call
(collapsing every `Fibroblast:*` back to `Fibroblast`, every
`Cardiomyocyte:*` to `Cardiomyocyte`, etc.). We use `RColorBrewer`'s
`Set1` / `Dark2` — pooled to get enough distinct, high-contrast hues for
up to ~12 broad lineages without relying on subtle shade differences.

We colour by the *broad* label rather than the fine one because the fine
label has 20+ levels — more than any qualitative palette can distinguish
cleanly. If we ever need to visualise at the fine resolution, a
`facet_wrap(~ broad)` with the fine label inside each panel is much more
legible than a single plot with 20+ colours.

```{r biplot-broad, fig.width=8, fig.height=6}
age_levels_biplot <- unique(reference$age_group)
ct_cols_biplot    <- setdiff(colnames(phi_scores), age_levels_biplot)
fine_call         <- getClass(phi_scores[, ct_cols_biplot])
scores_df$broad   <- sub(":.*$", "", fine_call)

n_broad <- length(unique(scores_df$broad))
broad_palette <- c(
    RColorBrewer::brewer.pal(9, "Set1"),
    RColorBrewer::brewer.pal(8, "Dark2")
)[seq_len(n_broad)]

ggplot(scores_df, aes(PC1, PC2, colour = broad)) +
    geom_point(size = 0.6, alpha = 0.8) +
    scale_colour_manual(values = broad_palette) +
    labs(title = "Query cells coloured by broad cell type",
         x = pc1_lab, y = pc2_lab, colour = "Broad cell type") +
    theme_bw(base_size = 13) +
    theme(legend.position = "right") +
    guides(colour = guide_legend(override.aes = list(size = 3, alpha = 1)))
```

Read this panel against the loadings panel above: each broad-cell-type blob
should sit on top of (or adjacent to) the loading arrows of its own
subtypes. Cardiomyocyte cells should cluster around the `Cardiomyocyte:*`
arrows, fibroblasts around the `Fibroblast:*` arrows, and so on.

## Hard vs Soft Classification

Sometimes you genuinely do need a single label per cell — for example, to
tabulate cell-type composition per donor. Φ-Space provides `getClass()` for
exactly this: it picks the column with the highest score for each cell.

```{r hard-labels}
age_cols        <- unique(reference$age_group)
cell_type_cols  <- setdiff(colnames(phi_scores), age_cols)

query$PhiCellType <- getClass(phi_scores[, cell_type_cols])
query$PhiAge      <- getClass(phi_scores[, age_cols])

# Broad label: collapse "CellType:Subtype" back to "CellType".
query$PhiCellTypeBroad <- sub(":.*$", "", query$PhiCellType)

table(query$PhiCellType)
table(query$PhiCellTypeBroad)
table(query$group, query$PhiAge)
```

The `group` × `PhiAge` cross-table is a useful sanity check: most foetal
query cells should land in the `fetal` reference bin, and most adult query
cells in `old` — but the off-diagonal entries are where the teaching is.
Those cells are the ones whose continuous scores straddle two age groups.

Let's look at the cardiomyocytes more closely — they are the main interest
for Module 6:

```{r cm-score-distribution, fig.width=10, fig.height=4}
cm_scores <- phi_scores[query$PhiCellTypeBroad == "Cardiomyocyte", age_cols, drop = FALSE]
cm_df <- as.data.frame(cm_scores) %>%
    tibble::rownames_to_column("cell") %>%
    mutate(group = query$group[match(cell, colnames(query))]) %>%
    tidyr::pivot_longer(all_of(age_cols), names_to = "age_phenotype",
                        values_to = "score") %>%
    mutate(age_phenotype = factor(age_phenotype,
                                  levels = c("fetal", "young", "middle", "old")))

ggplot(cm_df, aes(age_phenotype, score, fill = group)) +
    geom_violin(scale = "width", alpha = 0.8) +
    scale_fill_brewer(palette = "Set1") +
    labs(x = "Reference age group", y = "Φ-Space score",
         fill = "Query stage",
         title = "Age-group scores for query cardiomyocytes") +
    theme_bw(base_size = 13)
```

Notice that even within a single query stage there is substantial spread
along the age-group axis. A hard classifier hides this spread; the continuous
scores are exactly what slingshot will exploit to trace a maturation
trajectory.

## Handing Off to Module 6

We save the annotated query object so that Module 6 can pick up where we
leave off.

```{r save-for-module-6}
final_cache <- file.path(results_dir, "05_phispace_query.rds")
saveRDS(query, final_cache)
cat("Annotated query at", final_cache, "\n")
```

The object now carries:

- `reducedDim(query, "PhiSpace")` — the score matrix (fine cell-type
  columns + age-group columns).
- `colData(query)$PhiCellType` — discrete *fine* cell-type label
  (CM / Fibroblast resolved to subtype).
- `colData(query)$PhiCellTypeBroad` — discrete broad cell-type label
  (CM and Fibroblast subtypes collapsed back).
- `colData(query)$PhiAge` — discrete maturation-stage label.
- all original metadata (`sample`, `group`, `sex`, QC metrics).

In Module 6 we will **subset this object to cardiomyocytes** (using
`PhiCellTypeBroad == "Cardiomyocyte"`) and apply **slingshot** to the
PhiSpace score matrix. The continuous age-group scores will act as the
biological anchor that tells slingshot which end of the trajectory is
foetal and which end is adult.

## Summary

- Φ-Space replaces a hard cell-type label with a continuous score per
  reference phenotype per cell — a natural fit for a dataset that spans a
  developmental gradient.
- The workflow is four steps: match normalisation (we used scran
  normalisation on both reference and query), keep common genes, run
  `PhiSpace()` with one or more phenotype layers, inspect the scores.
- Continuous scores can always be collapsed to a discrete label with
  `getClass()`, but the inverse is not possible — preserve the scores for
  any downstream analysis that benefits from soft assignments (trajectory
  inference, co-phenotype analysis, gradient visualisation).
- The phenotype-space embedding we saved is the input to Module 6
  (pseudotime) and Module 7 (cell-specific co-expression networks).

## Session Info

```{r session-info}
sessionInfo()
```
