---
title: "Module 6: Pseudotime Trajectory Analysis"
author: "Single Cell Workshop"
date: "`r Sys.Date()`"
output:
  html_document:
    theme: flatly
    toc: true
    toc_depth: 2
    toc_float: true
    number_sections: true
    code_folding: show
    df_print: paged
vignette: >
  %\VignetteEncoding{UTF-8}
  %\VignetteIndexEntry{Module 6: Pseudotime Trajectory Analysis}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  markdown: 
    wrap: 72
---

```{r setup, include=FALSE}
# Module 6 consumes Module 5's PhiSpace output. If that cache is missing
# (e.g. fresh clone, or CI before the Zenodo upload), skip execution but
# still render the prose so the site build succeeds.
data_ok <- file.exists("../results/05_phispace_query.rds")

knitr::opts_chunk$set(
    echo = TRUE,
    warning = FALSE,
    message = FALSE,
    fig.align = "center",
    fig.width = 10,
    fig.height = 6,
    eval = data_ok
)
```

```{asis, echo = !data_ok}
> **Note.** This module's code chunks are not evaluated because
> `results/05_phispace_query.rds` is missing. Render Module 5 first, or
> download the Zenodo cache.
```

## Introduction

In Module 5 we annotated each query cell with a continuous score on
every reference phenotype. Those scores already *hint* at a foetal→adult
gradient within cardiomyocytes — recall the age-group scores climbing
smoothly from `fetal` to `old`. Module 6 turns that hint into a single
scalar per cell: **pseudotime**.

Pseudotime is a real number per cell such that cells that are close in
the underlying biological process have close values. It is not real
time, and the absolute scale is not identified — it is an *ordering*.
Direction (which end is "early") is a separate question, and for
principal-curve methods like slingshot it has to be supplied externally.
This is exactly where the Φ-Space age-group scores earn their keep: the
cluster with the highest mean `fetal` score is an obvious anchor for the
start of the cardiomyocyte maturation trajectory.

### Two methods, two embeddings, two clusterings

We run two families of trajectory method:

- **Slingshot** — a *principal-curve* method. Cluster cells, build a
  minimum spanning tree on the cluster centroids, fit a smooth curve
  through each lineage, project every cell onto the curve. Pseudotime =
  arc length.
- **DPT (diffusion pseudotime) via destiny** — a *diffusion / random
  walk* method. Build a Gaussian-kernel affinity graph, eigendecompose
  the associated Markov chain, and use accumulated transition
  probability from a root as pseudotime.

Each method runs on two embeddings — **Φ-Space scores** and **PCA of
expression**. Crossing method with embedding separates a *method effect*
from an *embedding effect*.

Slingshot additionally depends on the **clustering** passed in: the MST
is built on cluster centroids, so the clustering you choose is a
first-class modelling choice, not a preprocessing detail. We therefore
run slingshot **twice** per embedding — once with unsupervised Louvain
clusters on PCA, once with the PhiCellType labels from Module 5 — to
show how much the clustering moves the trajectory.

This gives six pseudotimes in total:

- Slingshot — 2 embeddings × 2 clusterings = 4 runs
  (`pt_sling_phi_unsup`, `pt_sling_pca_unsup`, `pt_sling_phi_phict`,
  `pt_sling_pca_phict`).
- DPT — 2 embeddings × 1 run each = 2 runs (`pt_dpt_phi`, `pt_dpt_pca`).

The four slingshot runs are used in a clustering-sensitivity figure; the
canonical four-way comparison at the end of the module uses the two
PhiCellType-clustered slingshot runs plus the two DPT runs.

Both methods can auto-detect endpoints if you omit the relevant argument
(slingshot infers `start.clus`/`end.clus` from the MST; destiny infers
`tips` from the diffusion-map geometry). We pin the root explicitly in
all runs — to the highest-`fetal`-score cluster / cell — so that the
pseudotimes are directly comparable.

### Learning objectives

By the end of this module you will be able to:

- Explain what pseudotime is, what it is not, and why it needs an
  external directionality anchor for principal-curve methods.
- Fit slingshot and DPT trajectories on two different embeddings (PCA
  and Φ-Space) and separate embedding effects from method effects.
- Visualise pseudotime and principal curves on a shared UMAP.
- Use a four-pseudotime correlation matrix as a sanity check on any
  single trajectory result.
- Save a pseudotime-annotated CM subset for Module 7.

## Load Libraries and Module 5 Output

```{r load-libraries}
library(SingleCellExperiment)
library(slingshot)
library(destiny)
library(scater)
library(ggplot2)
library(dplyr)
library(tibble)
library(patchwork)
library(RColorBrewer)
```

The handoff from Module 5 is a single file — the query SCE with PhiSpace
scores in `reducedDim`, scran logcounts in `assay(query, "logcounts")`,
and per-cell annotations (`PhiCellType`, `PhiCellTypeBroad`, `PhiAge`,
`group`, etc.) in `colData`.

```{r load-module5}
query <- readRDS("../results/05_phispace_query.rds")

cat("Assays:      ", paste(assayNames(query), collapse = ", "), "\n")
cat("reducedDims: ", paste(reducedDimNames(query), collapse = ", "), "\n")
cat("Cells:       ", ncol(query), "\n")
cat("Broad types: ", paste(sort(unique(query$PhiCellTypeBroad)),
                            collapse = ", "), "\n")
```

```{r cache-setup}
results_dir <- "../results"
dir.create(results_dir, showWarnings = FALSE, recursive = TRUE)
```

## Subset to Cardiomyocytes

Cardiomyocytes carry the foetal→adult maturation signal we want to
trace. Fibroblasts and the other lineages are interesting but secondary;
running pseudotime on the whole query would mix several biological
processes into one trajectory.

```{r subset-cm}
cm <- query[, query$PhiCellTypeBroad == "Cardiomyocyte"]
cat("Cardiomyocytes:", ncol(cm), "\n")
print(table(cm$group))
print(sort(table(cm$PhiCellType), decreasing = TRUE))
```

The fine Φ-Space labels give us biologically meaningful CM subtypes
(`Cardiomyocyte:Fetal_enriched_*`, `:Metabolic`, `:Stressed`, …). They
are reference-informed though, so we'll *not* feed them directly to
slingshot — instead we'll compute an **unsupervised** clustering on the
PCA embedding (section below) and use that as the slingshot input. The
PhiCellType labels still matter for interpretation and for anchor
picking via the reference age-group scores.

## Expression-Based Embedding: PCA on scran logcounts

For the "classical" slingshot run we need an embedding of gene
expression. We compute a standard HVG → PCA pipeline on the scran
logcounts we already have from Module 5.

```{r pca-expression}
pca_cache <- file.path(results_dir, "06_cm_pca.rds")

if (file.exists(pca_cache)) {
    cm <- readRDS(pca_cache)
    message("Loaded cached CM object with PCA from ", pca_cache)
} else {
    # Highly variable gene selection
    gene_var <- rowVars(as.matrix(logcounts(cm)))
    names(gene_var) <- rownames(cm)
    n_hvg <- min(2000, nrow(cm))
    hvg <- names(sort(gene_var, decreasing = TRUE))[seq_len(n_hvg)]

    pca <- prcomp(t(as.matrix(logcounts(cm)[hvg, ])),
                  center = TRUE, scale. = FALSE)
    reducedDim(cm, "PCA") <- pca$x[, 1:20]
    saveRDS(cm, pca_cache)
}

cat("PCA dims:      ", paste(dim(reducedDim(cm, "PCA")), collapse = " x "), "\n")
cat("PhiSpace dims: ", paste(dim(reducedDim(cm, "PhiSpace")),
                             collapse = " x "), "\n")
```

We keep 20 PCs the default.

## UMAP for Visualisation

We compute a single UMAP on the PCA embedding for visualisation. All
pseudotime panels end up displayed on this same layout, which makes
cross-method comparison a matter of eyeballing the same regions of the
same plot. `n_neighbors = 15` and `min_dist = 0.5` were picked from the
parameter sweep in `scripts/06_umap_param_sweep.R` — low `n_neighbors`
values fragmented the cardiomyocyte continuum into islands that had no
biological meaning.

None of the pseudotime methods below take UMAP as input; the layout is
strictly for plotting.

```{r umap}
umap_cache <- file.path(results_dir, "06_cm_umap.rds")

if (file.exists(umap_cache)) {
    reducedDim(cm, "UMAP") <- readRDS(umap_cache)
    message("Loaded cached UMAP")
} else {
    set.seed(42)
    cm <- scater::runUMAP(cm, dimred = "PCA", name = "UMAP",
                          n_neighbors = 15, min_dist = 0.5)
    saveRDS(reducedDim(cm, "UMAP"), umap_cache)
}
```

### Sanity check: does the UMAP make biological sense?

Four panels before we run any trajectory fitting:

1.  **Donor stage** (`group`) — we expect a smooth foetal → adult
    gradient if the embedding captures the maturation axis.
2.  **Fetal score** — the Φ-Space `fetal` coordinate on its own. Blue →
    white; the bluer a cell, the more fetal. This is also what we use to
    pick the slingshot anchor cluster.
3.  **Φ-Space pseudotime**, defined as `adult − fetal` on the reference
    age-group scores. The Gao 2026 reference has four age bins (`fetal`,
    `young`, `middle`, `old`); we take the adult score as the mean of
    the three post-natal bins. Their difference is a per-cell scalar
    that is already an ordering along the maturation axis — no
    trajectory model needed.
4.  **Φ-Space pseudotime (rescaled)** — the same score passed through a
    sign-preserving cube-root transform (`sign(x) · |x|^(1/3)`). The raw
    `adult − fetal` is heavy-tailed and the colour gradient gets
    dominated by a handful of extreme cells; cube-root compresses the
    tails without flipping any orderings.

The Φ-Space pseudotime (panels 3 and 4) is a useful anchor for the rest
of the module. It is the "trivially defined" pseudotime that slingshot
and DPT will be compared against. If none of the fitted pseudotimes
reproduce this gradient, something has gone wrong.

```{r umap-sanity, fig.width=13, fig.height=11}
umap_df0 <- as.data.frame(reducedDim(cm, "UMAP"))
colnames(umap_df0) <- c("UMAP1", "UMAP2")

# Φ-Space pseudotime: adult-vs-fetal contrast on the reference
# age-group scores. "adult" = mean of the three post-natal bins.
phi_scores_all <- reducedDim(cm, "PhiSpace")
adult_score    <- rowMeans(phi_scores_all[, c("young", "middle", "old")])
cm$pt_phi_score    <- adult_score - phi_scores_all[, "fetal"]
cm$pt_phi_score_cr <- sign(cm$pt_phi_score) * abs(cm$pt_phi_score)^(1/3)
cm$fetal_score     <- phi_scores_all[, "fetal"]

base_umap <- function(df, colour_col) {
    ggplot(df, aes(UMAP1, UMAP2, colour = .data[[colour_col]])) +
        geom_point(size = 0.4, alpha = 0.7) +
        theme_bw(base_size = 11) +
        theme(aspect.ratio = 1)
}

p_by_group <- base_umap(
    cbind(umap_df0, group = cm$group), "group"
) +
    scale_colour_brewer(palette = "Set1") +
    labs(title = "Coloured by donor stage", colour = "group")

p_by_fetal <- base_umap(
    cbind(umap_df0, fetal = cm$fetal_score), "fetal"
) +
    scale_colour_gradient(low = "white", high = "blue") +
    labs(title = "Fetal score", colour = "fetal")

p_by_phi_pt <- base_umap(
    cbind(umap_df0, pt = cm$pt_phi_score), "pt"
) +
    scale_colour_distiller(palette = "Spectral") +
    labs(title = "Φ-Space pseudotime (adult − fetal)",
         colour = "adult − fetal")

p_by_phi_pt_cr <- base_umap(
    cbind(umap_df0, pt_cr = cm$pt_phi_score_cr), "pt_cr"
) +
    scale_colour_distiller(palette = "Spectral") +
    labs(title = "Φ-Space pseudotime (cube-root rescaled)",
         colour = "sign·|·|^(1/3)")

(p_by_group | p_by_fetal) / (p_by_phi_pt | p_by_phi_pt_cr)
```

## Unsupervised Clustering

Slingshot (and monocle) are *clustering-based* trajectory methods: they
build a minimum spanning tree on cluster centroids and fit principal
curves through the MST. Your clustering is not a preprocessing detail
but a first-class modelling choice — two different clusterings of the
same cells can give two different backbone topologies and, in turn, two
different pseudotimes. This is arguably the biggest practical drawback
of principal-curve methods, and one reason diffusion-based methods (DPT
below) are useful as a cross-check.

We compute a standard unsupervised clustering — SNN graph on the PCA
embedding, Louvain communities — and use *that* as the slingshot input.
Using the unsupervised clusters rather than PhiCellType means the
slingshot fit is not directly circular with the Φ-Space anchoring (the
start cluster is still chosen via the Φ-Space `fetal` score, but the
cluster definitions themselves are reference-free).

```{r cluster-snn}
cluster_cache <- file.path(results_dir, "06_cm_clusters.rds")

if (file.exists(cluster_cache)) {
    cm$cluster <- readRDS(cluster_cache)
    message("Loaded cached clusters")
} else {
    set.seed(42)
    snn <- scran::buildSNNGraph(cm, use.dimred = "PCA", k = 20)
    cm$cluster <- factor(igraph::cluster_louvain(snn)$membership)
    saveRDS(cm$cluster, cluster_cache)
}

cat("Cluster sizes:\n")
print(table(cm$cluster))
```

```{r umap-clusters, fig.width=7, fig.height=5.5}
clust_levels <- levels(cm$cluster)
clust_pool <- c(brewer.pal(9, "Set1"),
                brewer.pal(8, "Dark2"),
                brewer.pal(8, "Set2"))
clust_pal <- clust_pool[seq_along(clust_levels)]
names(clust_pal) <- clust_levels

ggplot(
    cbind(umap_df0, cluster = cm$cluster),
    aes(UMAP1, UMAP2, colour = cluster)
) +
    geom_point(size = 0.4, alpha = 0.7) +
    scale_colour_manual(values = clust_pal) +
    theme_bw(base_size = 11) +
    theme(aspect.ratio = 1) +
    guides(colour = guide_legend(override.aes = list(size = 2))) +
    labs(title = "Unsupervised Louvain clusters (SNN on PCA)")
```

## Picking the Trajectory Anchors

For each slingshot run we need a `start.clus`. We pick the cluster whose
mean score on the reference `fetal` phenotype is highest — Φ-Space
chooses the direction, the clustering defines the units. Since we are
running slingshot with two different clusterings, we need two anchors.

```{r pick-anchor}
phi_scores  <- reducedDim(cm, "PhiSpace")
fetal_score <- phi_scores[, "fetal"]

mean_fetal_by <- function(cluster_vec) {
    tibble(cluster = cluster_vec, fetal = fetal_score) |>
        group_by(cluster) |>
        summarise(mean_fetal = mean(fetal), n = n(),
                  .groups = "drop") |>
        arrange(desc(mean_fetal))
}

anchor_unsup_tbl <- mean_fetal_by(cm$cluster)
anchor_phict_tbl <- mean_fetal_by(cm$PhiCellType)

cat("Anchor table — unsupervised clusters:\n")
print(anchor_unsup_tbl)
cat("\nAnchor table — PhiCellType:\n")
print(anchor_phict_tbl)

start_cluster_unsup <- as.character(anchor_unsup_tbl$cluster[1])
start_cluster_phict <- as.character(anchor_phict_tbl$cluster[1])

cat("\nstart.clus (unsupervised): ", start_cluster_unsup, "\n", sep = "")
cat("start.clus (PhiCellType):  ",  start_cluster_phict, "\n", sep = "")
```

For destiny we need a single cell, not a cluster. We decouple this
choice from any clustering and just pick the globally most fetal cell,
so that the two DPT runs don't depend on which Louvain partition we
happened to compute.

```{r root-cell}
root_cell <- which.max(fetal_score)
cat("Root cell index:", root_cell,
    " (fetal score =", round(fetal_score[root_cell], 3), ")\n")
```

## Slingshot: four runs

Two embeddings × two clusterings. The same `dist.method = "simple"` is
used throughout — the default (scaled Mahalanobis) is singular when
small PhiCellType subclusters push the within-cluster covariance below
the inversion threshold.

```{r slingshot-runs}
run_slingshot <- function(cache, dimred, clusters, start) {
    fp <- file.path(results_dir, cache)
    if (file.exists(fp)) {
        message("Loaded cached ", cache)
        return(readRDS(fp))
    }
    sds <- slingshot(
        reducedDim(cm, dimred),
        clusterLabels = clusters,
        start.clus    = start,
        dist.method   = "simple"
    )
    saveRDS(sds, fp)
    sds
}

sling_phi_unsup <- run_slingshot("06_slingshot_phi_unsup.rds",
                                 "PhiSpace", cm$cluster,
                                 start_cluster_unsup)
sling_pca_unsup <- run_slingshot("06_slingshot_pca_unsup.rds",
                                 "PCA",      cm$cluster,
                                 start_cluster_unsup)
sling_phi_phict <- run_slingshot("06_slingshot_phi_phict.rds",
                                 "PhiSpace", cm$PhiCellType,
                                 start_cluster_phict)
sling_pca_phict <- run_slingshot("06_slingshot_pca_phict.rds",
                                 "PCA",      cm$PhiCellType,
                                 start_cluster_phict)
```

Slingshot's `slingPseudotime()` returns one column per lineage; for a
predominantly linear CM maturation trajectory we expect one dominant
lineage. We average over lineages to get a single pseudotime per cell.

```{r slingshot-pts}
lineage_pt <- function(sds) rowMeans(slingPseudotime(sds), na.rm = TRUE)

cm$pt_sling_phi_unsup <- lineage_pt(sling_phi_unsup)
cm$pt_sling_pca_unsup <- lineage_pt(sling_pca_unsup)
cm$pt_sling_phi_phict <- lineage_pt(sling_phi_phict)
cm$pt_sling_pca_phict <- lineage_pt(sling_pca_phict)

cat("Lineages found per run:\n")
cat("  PhiSpace / unsupervised: ", length(slingLineages(sling_phi_unsup)), "\n")
cat("  PCA      / unsupervised: ", length(slingLineages(sling_pca_unsup)), "\n")
cat("  PhiSpace / PhiCellType:  ", length(slingLineages(sling_phi_phict)), "\n")
cat("  PCA      / PhiCellType:  ", length(slingLineages(sling_pca_phict)), "\n")
```

## DPT via destiny

For the diffusion branch of the 2 × 2 grid we build a diffusion map on
each embedding and compute DPT from the same root cell we pinned above.

### DPT on PCA

```{r destiny-dpt-pca}
dpt_cache <- file.path(results_dir, "06_dpt_pca.rds")

if (file.exists(dpt_cache)) {
    dpt <- readRDS(dpt_cache)
    message("Loaded cached DPT/PCA result")
} else {
    dm  <- DiffusionMap(as.matrix(reducedDim(cm, "PCA")),
                        n_eigs = 20, sigma = "local", verbose = FALSE)
    dpt <- DPT(dm, tips = root_cell)
    saveRDS(dpt, dpt_cache)
}

cm$pt_dpt_pca <- dpt$dpt
summary(cm$pt_dpt_pca)
```

### DPT on Φ-Space

Identical settings to the PCA run — same `n_eigs`, same `sigma`, same
root — so the two DPT pseudotimes differ in exactly one thing: the input
embedding.

```{r destiny-dpt-phi}
dpt_phi_cache <- file.path(results_dir, "06_dpt_phispace.rds")

if (file.exists(dpt_phi_cache)) {
    dpt_phi <- readRDS(dpt_phi_cache)
    message("Loaded cached DPT/PhiSpace result")
} else {
    dm_phi <- DiffusionMap(as.matrix(reducedDim(cm, "PhiSpace")),
                           n_eigs = 20, sigma = "local",
                           verbose = FALSE)
    dpt_phi <- DPT(dm_phi, tips = root_cell)
    saveRDS(dpt_phi, dpt_phi_cache)
}

cm$pt_dpt_phi <- dpt_phi$dpt
summary(cm$pt_dpt_phi)
```

Calling `DPT(dm)` with no `tips` would have let destiny auto-detect
trajectory endpoints from the diffusion-map geometry. We pass the root
explicitly here to keep destiny's pseudotime directly comparable with
the slingshot runs (which also received an explicit `start.clus`).

## Visualising Pseudotimes on UMAP

We colour cells by pseudotime using the `Spectral` ColorBrewer palette
(blue for low values, red for high) — an ordered diverging scale whose
two endpoints are visually distinct, unlike sequential palettes like
viridis. Every panel sits on the shared UMAP computed above; for
slingshot panels we overlay the principal curves using
`slingshot::embedCurves()`, which re-fits the curves in UMAP space given
the existing cell ordering.

```{r rescale-pt}
rescale01 <- function(x) {
    x <- x - min(x, na.rm = TRUE)
    x / max(x, na.rm = TRUE)
}

cm$pt_sling_phi_unsup_01 <- rescale01(cm$pt_sling_phi_unsup)
cm$pt_sling_pca_unsup_01 <- rescale01(cm$pt_sling_pca_unsup)
cm$pt_sling_phi_phict_01 <- rescale01(cm$pt_sling_phi_phict)
cm$pt_sling_pca_phict_01 <- rescale01(cm$pt_sling_pca_phict)
cm$pt_dpt_phi_01         <- rescale01(cm$pt_dpt_phi)
cm$pt_dpt_pca_01         <- rescale01(cm$pt_dpt_pca)
```

```{r umap-curves}
umap_df <- as.data.frame(reducedDim(cm, "UMAP"))
colnames(umap_df) <- c("UMAP1", "UMAP2")

# Embed slingshot curves from their native embeddings into the shared UMAP.
embed_curves_df <- function(sds, label) {
    sds_u  <- embedCurves(sds, reducedDim(cm, "UMAP"))
    curves <- slingCurves(sds_u)
    do.call(rbind, lapply(seq_along(curves), function(i) {
        crv <- curves[[i]]
        s   <- crv$s[crv$ord, , drop = FALSE]
        data.frame(UMAP1 = s[, 1], UMAP2 = s[, 2],
                   lineage = paste0(label, "_L", i))
    }))
}

curves_phi_unsup_df <- embed_curves_df(sling_phi_unsup, "phi_unsup")
curves_pca_unsup_df <- embed_curves_df(sling_pca_unsup, "pca_unsup")
curves_phi_phict_df <- embed_curves_df(sling_phi_phict, "phi_phict")
curves_pca_phict_df <- embed_curves_df(sling_pca_phict, "pca_phict")

pt_scale <- scale_colour_distiller(
    palette = "Spectral",
    name    = "Pseudotime"
)

umap_base <- function(pt_vec) {
    df <- cbind(umap_df, pt = pt_vec)
    ggplot(df, aes(UMAP1, UMAP2, colour = pt)) +
        geom_point(size = 0.4, alpha = 0.7) +
        pt_scale +
        theme_bw(base_size = 11) +
        theme(aspect.ratio = 1)
}

with_curves <- function(p, crv_df) {
    p + geom_path(data = crv_df,
                  aes(UMAP1, UMAP2, group = lineage),
                  inherit.aes = FALSE,
                  colour = "lightgray", linewidth = 0.8)
}
```

### Slingshot clustering sensitivity

The same slingshot call, run four times — two embeddings × two
clusterings. If the four panels paint different gradients on the same
UMAP, the clustering choice is not a preprocessing detail.

```{r umap-sling-sensitivity, fig.width=12, fig.height=10}
p_phi_unsup <- with_curves(umap_base(cm$pt_sling_phi_unsup_01),
                           curves_phi_unsup_df) +
    labs(title = "Φ-Space / unsupervised clusters")
p_pca_unsup <- with_curves(umap_base(cm$pt_sling_pca_unsup_01),
                           curves_pca_unsup_df) +
    labs(title = "PCA / unsupervised clusters")
p_phi_phict <- with_curves(umap_base(cm$pt_sling_phi_phict_01),
                           curves_phi_phict_df) +
    labs(title = "Φ-Space / PhiCellType")
p_pca_phict <- with_curves(umap_base(cm$pt_sling_pca_phict_01),
                           curves_pca_phict_df) +
    labs(title = "PCA / PhiCellType")

(p_phi_unsup | p_pca_unsup) / (p_phi_phict | p_pca_phict) +
    plot_layout(guides = "collect") +
    plot_annotation(
        title    = "Slingshot on two embeddings × two clusterings",
        subtitle = "Top row: unsupervised Louvain clusters on PCA. Bottom row: PhiCellType labels from Module 5."
    )
```

Reading the figure:

- **Top row vs bottom row** (clustering effect). With unsupervised
  Louvain clusters the trajectory tends to follow whatever the Louvain
  partition happened to emphasise on the PCA graph — including axes that
  have little to do with maturation. Switching to PhiCellType, where the
  clusters are defined by the biology we care about (fetal-enriched /
  metabolic / stressed / ...), pulls the principal curves onto the
  actual foetal → adult axis.
- **Left column vs right column** (embedding effect). Φ-Space tends to
  produce a smoother principal curve because the embedding *itself* is
  aligned with the reference age groups; PCA is free to detour through
  donor or batch structure before reaching the adult end.

For the canonical comparison with DPT below we use the
**PhiCellType-clustered** slingshot runs (bottom row).

### Canonical four-way: slingshot (PhiCellType) vs DPT

```{r umap-pt-plots, fig.width=12, fig.height=10}
p_sling_phi <- with_curves(umap_base(cm$pt_sling_phi_phict_01),
                           curves_phi_phict_df) +
    labs(title = "slingshot / Φ-Space")
p_sling_pca <- with_curves(umap_base(cm$pt_sling_pca_phict_01),
                           curves_pca_phict_df) +
    labs(title = "slingshot / PCA")
p_dpt_phi <- umap_base(cm$pt_dpt_phi_01) +
    labs(title = "DPT / Φ-Space")
p_dpt_pca <- umap_base(cm$pt_dpt_pca_01) +
    labs(title = "DPT / PCA")

(p_sling_phi | p_sling_pca) / (p_dpt_phi | p_dpt_pca) +
    plot_layout(guides = "collect") +
    plot_annotation(
        title = "Pseudotime on UMAP — two methods × two embeddings"
    )
```

If all four panels paint the same blue → red gradient across the same
UMAP regions, the trajectory is robust. If the two slingshot panels
agree but the two DPT panels disagree with them, that is a pure method
effect; if the two Φ-Space panels agree but the two PCA panels do not,
that is a pure embedding effect.

## Four-Way Comparison

For the quantitative comparison we compare slingshot run with
**PhiCellType clusters** — the variant that best recovered the foetal →
adult axis in the sensitivity figure above — against DPT on the same two
embeddings. This is the 2 × 2 grid of methods × embeddings we promised
in the introduction.

```{r pt-mat}
pt_mat <- as.data.frame(colData(cm)[,
    c("pt_sling_phi_phict_01", "pt_sling_pca_phict_01",
      "pt_dpt_phi_01",         "pt_dpt_pca_01")])
colnames(pt_mat) <- c("slingshot/PhiSpace", "slingshot/PCA",
                      "DPT/PhiSpace",       "DPT/PCA")
```

### Correlation matrix

```{r pt-cor}
round(cor(pt_mat, use = "complete.obs", method = "spearman"), 3)
```

Spearman (rank) correlation is the right scalar summary because each
method has a different absolute scale and we only care about the *order*
of cells.

### Pairs plot

```{r pt-pairs, fig.width=10, fig.height=10}
pt_long <- pt_mat |>
    rownames_to_column("cell") |>
    mutate(group = cm$group[match(cell, colnames(cm))])

make_pair <- function(x, y) {
    ggplot(pt_long, aes(.data[[x]], .data[[y]], colour = group)) +
        geom_point(size = 0.4, alpha = 0.5) +
        geom_abline(slope = 1, intercept = 0,
                    linetype = "dashed", alpha = 0.5) +
        scale_colour_brewer(palette = "Set1") +
        theme_bw(base_size = 11) +
        theme(legend.position = "none") +
        labs(x = x, y = y)
}

method_names <- c("slingshot/PhiSpace", "slingshot/PCA",
                  "DPT/PhiSpace",       "DPT/PCA")
pair_idx <- combn(method_names, 2, simplify = FALSE)
pairs <- lapply(pair_idx, function(p) make_pair(p[1], p[2]))

wrap_plots(pairs, ncol = 3) +
    plot_annotation(title = "Pairwise agreement between pseudotimes")
```

Read the 2 × 2 grid of methods diagonally:

- **Rows of the method table** (slingshot/PhiSpace ↔ slingshot/PCA, or
  DPT/PhiSpace ↔ DPT/PCA) isolate the *embedding* effect — same method,
  different embedding.
- **Columns of the method table** (slingshot/PhiSpace ↔ DPT/PhiSpace, or
  slingshot/PCA ↔ DPT/PCA) isolate the *method* effect — same embedding,
  different trajectory family.
- **Anti-diagonal pairs** (slingshot/PhiSpace ↔ DPT/PCA, slingshot/PCA ↔
  DPT/PhiSpace) vary both at once; agreement here is the strongest
  evidence the trajectory is not an artefact of either choice.

### Pseudotime vs donor stage

```{r pt-vs-group, fig.width=12, fig.height=4}
box_df <- pt_long |>
    tidyr::pivot_longer(
        cols = all_of(method_names),
        names_to = "method", values_to = "pseudotime"
    ) |>
    mutate(method = factor(method, levels = method_names))

ggplot(box_df, aes(group, pseudotime, fill = group)) +
    geom_violin(scale = "width", alpha = 0.8) +
    geom_boxplot(width = 0.15, alpha = 0.6, outlier.size = 0.3) +
    scale_fill_brewer(palette = "Set1") +
    facet_wrap(~ method, nrow = 1) +
    labs(x = "Donor stage", y = "Pseudotime (rescaled)",
         fill = "Stage",
         title = "Do pseudotimes recover the known foetal → adult ordering?") +
    theme_bw(base_size = 12) +
    theme(legend.position = "none")
```

The donor stage (`group`) is held-out information: none of the four
pseudotime methods saw it. If the violins step upward cleanly from
`fetal` to `adult` on all four panels, the trajectory is recovering the
real maturation order.

## Handing Off to Module 7

We save the CM object with the Φ-Space maturation score and all six
pseudotimes attached so Module 7 can use the same cardiomyocyte subset
for cell-specific co-expression network analysis.

```{r save-for-module-7}
final_cache <- file.path(results_dir, "06_cm_pseudotime.rds")
saveRDS(cm, final_cache)
cat("Annotated CM object at", final_cache, "\n")
```

The object carries, in addition to everything Module 5 handed us:

- `reducedDim(cm, "PCA")` — 20 PCs on scran logcounts of HVGs.
- `reducedDim(cm, "UMAP")` — 2D UMAP on the PCA embedding
  (`n_neighbors = 15`, `min_dist = 0.5`).
- `colData(cm)$cluster` — unsupervised Louvain clusters on the PCA SNN
  graph; used in the clustering-sensitivity slingshot runs.
- `colData(cm)$pt_phi_score` — Φ-Space pseudotime (`adult − fetal`
  reference score), a trajectory-free baseline.
- `colData(cm)$pt_sling_phi_unsup` / `pt_sling_pca_unsup` — slingshot
  with unsupervised clusters (clustering-sensitivity runs).
- `colData(cm)$pt_sling_phi_phict` / `pt_sling_pca_phict` — slingshot
  with PhiCellType clusters (canonical).
- `colData(cm)$pt_dpt_phi` / `pt_dpt_pca` — DPT pseudotimes from the two
  destiny runs.
- `colData(cm)$pt_*_01` — the same pseudotimes rescaled to `[0, 1]` for
  cross-method comparison.

For Module 7 we will primarily use `pt_phi_score` — the trajectory-free
Φ-Space maturation baseline — to select maturation-associated target
genes. The six fitted pseudotimes live in the object for optional
sensitivity analyses, but the NeighbourNet module is intentionally
decoupled from the slingshot/DPT comparison.

## Summary

- Pseudotime gives a scalar ordering of cells along a biological
  process; it is not real time, and its direction for principal-curve
  methods comes from an external anchor.
- We ran six pseudotimes: four slingshot runs crossing two embeddings
  with two clusterings, plus two DPT runs crossing PCA and Φ-Space.
- Slingshot is our principal-curve method: cluster centroids → MST →
  principal curves → arc-length pseudotime. We overlay its curves on the
  UMAP to make the trajectory visible.
- DPT via destiny is our diffusion cross-check. It adds two useful
  things: (i) automatic endpoint detection, and (ii) a fundamentally
  different statistical object (random-walk probabilities rather than
  arc length), which catches principal-curve artefacts.
- The four-pseudotime comparison is not bureaucratic — crossing method
  with embedding is what lets us separate embedding effects from method
  effects and decide whether the trajectory is robust to either.

## Session Info

```{r session-info}
sessionInfo()
```
