---
title: "Module 6: Pseudotime Trajectory Analysis"
author: "Single Cell Workshop"
date: "`r Sys.Date()`"
output:
  html_document:
    theme: flatly
    toc: true
    toc_depth: 2
    toc_float: true
    number_sections: true
    code_folding: show
    df_print: paged
vignette: >
  %\VignetteIndexEntry{Module 6: Pseudotime Trajectory Analysis}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
# Module 6 consumes Module 5's PhiSpace output. If that cache is missing
# (e.g. fresh clone, or CI before the Zenodo upload), skip execution but
# still render the prose so the site build succeeds.
data_ok <- file.exists("../results/05_phispace_query.rds")

knitr::opts_chunk$set(
    echo = TRUE,
    warning = FALSE,
    message = FALSE,
    fig.align = "center",
    fig.width = 10,
    fig.height = 6,
    eval = data_ok
)
```

```{asis, echo = !data_ok}
> **Note.** This module's code chunks are not evaluated because
> `results/05_phispace_query.rds` is missing. Render Module 5 first, or
> download the Zenodo cache.
```

## Introduction

In Module 5 we annotated each query cell with a continuous score on every
reference phenotype. Those scores already *hint* at a foetal→adult
gradient within cardiomyocytes — recall the age-group scores climbing
smoothly from `fetal` to `old`. Module 6 turns that hint into a single
scalar per cell: **pseudotime**.

Pseudotime is a real number per cell such that cells that are close in the
underlying biological process have close values. It is not real time, and
the absolute scale is not identified — it is an *ordering*. Direction
(which end is "early") is a separate question, and for principal-curve
methods like slingshot it has to be supplied externally. This is exactly
where the Φ-Space age-group scores earn their keep: the cluster with the
highest mean `fetal` score is an obvious anchor for the start of the
cardiomyocyte maturation trajectory.

### Two methods, two embeddings

We run two families of method, chosen to be maximally complementary:

- **Slingshot** — a *principal-curve* method. Cluster cells, build a
  minimum spanning tree on the cluster centroids, fit a smooth curve
  through each lineage, project every cell onto the curve. Pseudotime =
  arc length. We run slingshot **twice**: once on PCA of scran logcounts
  (the classical choice), once on the Φ-Space score matrix. Same cells,
  same clusters, same anchor. Any difference between the two pseudotimes
  is an *embedding effect*.
- **DPT (diffusion pseudotime) via destiny** — a *diffusion / random walk*
  method. Build a Gaussian-kernel affinity graph, eigendecompose the
  associated Markov chain, and use accumulated transition probability
  from a root as pseudotime. We run DPT once, on PCA of scran logcounts,
  as a cross-check: a different statistical machinery on the *same*
  embedding as the first slingshot run.

One thing destiny gets "for free" that slingshot does not: **endpoint
detection**. Without a `tips` argument, destiny picks trajectory tips
automatically from the diffusion-map geometry. We still choose to pin the
root at the highest-`fetal`-score cell, for comparability, but the
auto-detection is a genuine advantage of diffusion methods for
exploratory work where you do not yet know which end is which. The
flip side — less explicit control over endpoints — is a feature, not a
defect; it is just a different point on the "automated vs controllable"
axis.

### Learning objectives

By the end of this module you will be able to:

- Explain what pseudotime is, what it is not, and why it needs an
  external directionality anchor for principal-curve methods.
- Fit a slingshot trajectory on two different embeddings and interpret
  the difference.
- Fit a DPT trajectory via destiny and compare it against slingshot.
- Use a simple multi-pseudotime correlation matrix as a sanity check on
  any single trajectory result.
- Save a pseudotime-annotated CM subset for Module 7.

## Load Libraries and Module 5 Output

```{r load-libraries}
library(SingleCellExperiment)
library(slingshot)
library(destiny)
library(scater)
library(ggplot2)
library(dplyr)
library(tibble)
library(patchwork)
library(RColorBrewer)
```

The handoff from Module 5 is a single file — the query SCE with PhiSpace
scores in `reducedDim`, scran logcounts in `assay(query, "logcounts")`,
and per-cell annotations (`PhiCellType`, `PhiCellTypeBroad`, `PhiAge`,
`group`, etc.) in `colData`.

```{r load-module5}
query <- readRDS("../results/05_phispace_query.rds")

cat("Assays:      ", paste(assayNames(query), collapse = ", "), "\n")
cat("reducedDims: ", paste(reducedDimNames(query), collapse = ", "), "\n")
cat("Cells:       ", ncol(query), "\n")
cat("Broad types: ", paste(sort(unique(query$PhiCellTypeBroad)),
                            collapse = ", "), "\n")
```

```{r cache-setup}
results_dir <- "../results"
dir.create(results_dir, showWarnings = FALSE, recursive = TRUE)
```

## Subset to Cardiomyocytes

Cardiomyocytes carry the foetal→adult maturation signal we want to trace.
Fibroblasts and the other lineages are interesting but secondary; running
pseudotime on the whole query would mix several biological processes
into one trajectory.

```{r subset-cm}
cm <- query[, query$PhiCellTypeBroad == "Cardiomyocyte"]
cat("Cardiomyocytes:", ncol(cm), "\n")
print(table(cm$group))
print(sort(table(cm$PhiCellType), decreasing = TRUE))
```

The fine Φ-Space labels give us biologically meaningful CM clusters
(`Cardiomyocyte:Fetal_enriched_*`, `:Metabolic`, `:Stressed`, …) without
any extra Leiden pass. We'll reuse them as the clustering input to
slingshot below.

## Expression-Based Embedding: PCA on scran logcounts

For the "classical" slingshot run we need an embedding of gene
expression. We compute a standard HVG → PCA pipeline on the scran
logcounts we already have from Module 5.

```{r pca-expression}
pca_cache <- file.path(results_dir, "06_cm_pca.rds")

if (file.exists(pca_cache)) {
    cm <- readRDS(pca_cache)
    message("Loaded cached CM object with PCA from ", pca_cache)
} else {
    # Highly variable gene selection
    gene_var <- rowVars(as.matrix(logcounts(cm)))
    names(gene_var) <- rownames(cm)
    n_hvg <- min(2000, nrow(cm))
    hvg <- names(sort(gene_var, decreasing = TRUE))[seq_len(n_hvg)]

    pca <- prcomp(t(as.matrix(logcounts(cm)[hvg, ])),
                  center = TRUE, scale. = FALSE)
    reducedDim(cm, "PCA") <- pca$x[, 1:20]
    saveRDS(cm, pca_cache)
}

cat("PCA dims:      ", paste(dim(reducedDim(cm, "PCA")), collapse = " x "), "\n")
cat("PhiSpace dims: ", paste(dim(reducedDim(cm, "PhiSpace")),
                             collapse = " x "), "\n")
```

We keep 20 PCs. That is a defensible default for single-cell data of this
size; tunePhiSpace-style cross-validation of the component count is
possible but not the point here.

## Picking the Trajectory Anchor

For slingshot we need a `start.clus`. Rather than hand-picking a root
cell, we pick the CM *cluster* whose mean score on the reference `fetal`
phenotype is highest — a fully data-driven anchor.

```{r pick-anchor}
phi_scores <- reducedDim(cm, "PhiSpace")
fetal_score <- phi_scores[, "fetal"]

anchor_tbl <- tibble(
    cluster = cm$PhiCellType,
    fetal   = fetal_score
) |>
    group_by(cluster) |>
    summarise(mean_fetal = mean(fetal), n = n(),
              .groups = "drop") |>
    arrange(desc(mean_fetal))

print(anchor_tbl)

start_cluster <- anchor_tbl$cluster[1]
cat("\nAnchor (start.clus):", start_cluster, "\n")
```

We also pre-compute the per-cell root index for destiny, which wants a
cell rather than a cluster.

```{r root-cell}
start_cells <- which(cm$PhiCellType == start_cluster)
root_cell   <- start_cells[which.max(fetal_score[start_cells])]
cat("Root cell index:", root_cell, "\n")
```

## Slingshot on Φ-Space Scores

```{r slingshot-phispace}
sling_phi_cache <- file.path(results_dir, "06_slingshot_phispace.rds")

if (file.exists(sling_phi_cache)) {
    sling_phi <- readRDS(sling_phi_cache)
    message("Loaded cached slingshot-PhiSpace result")
} else {
    sling_phi <- slingshot(
        reducedDim(cm, "PhiSpace"),
        clusterLabels = cm$PhiCellType,
        start.clus    = start_cluster,
        dist.method   = "simple"
    )
    saveRDS(sling_phi, sling_phi_cache)
}

cat("Lineages found:\n")
print(slingLineages(sling_phi))
```

Slingshot's `slingPseudotime()` returns one column per lineage; for a
predominantly linear CM maturation trajectory we expect one dominant
lineage.

```{r pt-phi}
pt_phi <- slingPseudotime(sling_phi)
cm$pt_sling_phi <- rowMeans(pt_phi, na.rm = TRUE)
summary(cm$pt_sling_phi)
```

## Slingshot on PCA of Expression

Same clusters, same anchor, different embedding.

```{r slingshot-pca}
sling_pca_cache <- file.path(results_dir, "06_slingshot_pca.rds")

if (file.exists(sling_pca_cache)) {
    sling_pca <- readRDS(sling_pca_cache)
    message("Loaded cached slingshot-PCA result")
} else {
    sling_pca <- slingshot(
        reducedDim(cm, "PCA"),
        clusterLabels = cm$PhiCellType,
        start.clus    = start_cluster,
        dist.method   = "simple"
    )
    saveRDS(sling_pca, sling_pca_cache)
}

pt_pca <- slingPseudotime(sling_pca)
cm$pt_sling_pca <- rowMeans(pt_pca, na.rm = TRUE)
summary(cm$pt_sling_pca)
```

## DPT via destiny on PCA

For the diffusion cross-check we build a diffusion map on the PCA
embedding and compute DPT from the same root cell we used above.

```{r destiny-dpt}
dpt_cache <- file.path(results_dir, "06_dpt_pca.rds")

if (file.exists(dpt_cache)) {
    dpt <- readRDS(dpt_cache)
    message("Loaded cached DPT result")
} else {
    dm  <- DiffusionMap(as.matrix(reducedDim(cm, "PCA")),
                        n_eigs = 20, sigma = "local", verbose = FALSE)
    dpt <- DPT(dm, tips = root_cell)
    saveRDS(dpt, dpt_cache)
}

cm$pt_dpt_pca <- dpt$dpt
summary(cm$pt_dpt_pca)
```

A note on what we *did not* pass: if we had called `DPT(dm)` with no
`tips`, destiny would have auto-detected trajectory endpoints from the
diffusion-map geometry — that is the "endpoints for free" property of
diffusion methods. We pass the root explicitly here only to keep
destiny's pseudotime directly comparable with the two slingshot runs.

## Three-Way Comparison

All three pseudotimes live on different absolute scales, so we rescale
each to `[0, 1]` before comparing.

```{r rescale-pt}
rescale01 <- function(x) {
    x <- x - min(x, na.rm = TRUE)
    x / max(x, na.rm = TRUE)
}

cm$pt_sling_phi_01 <- rescale01(cm$pt_sling_phi)
cm$pt_sling_pca_01 <- rescale01(cm$pt_sling_pca)
cm$pt_dpt_pca_01   <- rescale01(cm$pt_dpt_pca)

pt_mat <- as.data.frame(colData(cm)[,
    c("pt_sling_phi_01", "pt_sling_pca_01", "pt_dpt_pca_01")])
colnames(pt_mat) <- c("slingshot/PhiSpace",
                      "slingshot/PCA",
                      "DPT/PCA")
```

### Correlation matrix

```{r pt-cor}
round(cor(pt_mat, use = "complete.obs", method = "spearman"), 3)
```

Spearman (rank) correlation is the right scalar summary because each
method has a different absolute scale and we only care about the *order*
of cells.

### Pairs plot

```{r pt-pairs, fig.width=9, fig.height=9}
pt_long <- pt_mat |>
    rownames_to_column("cell") |>
    mutate(group = cm$group[match(cell, colnames(cm))])

make_pair <- function(x, y) {
    ggplot(pt_long, aes(.data[[x]], .data[[y]], colour = group)) +
        geom_point(size = 0.4, alpha = 0.5) +
        geom_abline(slope = 1, intercept = 0,
                    linetype = "dashed", alpha = 0.5) +
        scale_colour_brewer(palette = "Set1") +
        theme_bw(base_size = 11) +
        theme(legend.position = "none") +
        labs(x = x, y = y)
}

pairs <- list(
    make_pair("slingshot/PhiSpace", "slingshot/PCA"),
    make_pair("slingshot/PhiSpace", "DPT/PCA"),
    make_pair("slingshot/PCA",      "DPT/PCA")
)

wrap_plots(pairs, ncol = 2) +
    plot_annotation(title = "Pairwise agreement between pseudotimes")
```

Read the three panels as answers to three different questions:

- **slingshot/PhiSpace vs slingshot/PCA** — pure *embedding* effect.
  Same method, same clusters, same anchor. Disagreement here says the
  choice of embedding moves the trajectory.
- **slingshot vs DPT on PCA** — pure *method* effect. Same embedding,
  different trajectory family. Disagreement says the result is
  geometry-specific.
- **slingshot/PhiSpace vs DPT/PCA** — the hardest diagonal; if this
  one agrees, the trajectory is probably robust.

### Pseudotime vs donor stage

```{r pt-vs-group, fig.width=12, fig.height=4}
box_df <- pt_long |>
    tidyr::pivot_longer(
        cols = c("slingshot/PhiSpace", "slingshot/PCA", "DPT/PCA"),
        names_to = "method", values_to = "pseudotime"
    ) |>
    mutate(method = factor(method,
        levels = c("slingshot/PhiSpace", "slingshot/PCA", "DPT/PCA")))

ggplot(box_df, aes(group, pseudotime, fill = group)) +
    geom_violin(scale = "width", alpha = 0.8) +
    geom_boxplot(width = 0.15, alpha = 0.6, outlier.size = 0.3) +
    scale_fill_brewer(palette = "Set1") +
    facet_wrap(~ method, nrow = 1) +
    labs(x = "Donor stage", y = "Pseudotime (rescaled)",
         fill = "Stage",
         title = "Do pseudotimes recover the known foetal → adult ordering?") +
    theme_bw(base_size = 12) +
    theme(legend.position = "none")
```

The donor stage (`group`) is held-out information: none of the three
pseudotime methods saw it. If the violins step upward cleanly from
`fetal` to `adult` on all three panels, the trajectory is recovering the
real maturation order.

## Handing Off to Module 7

We save the CM object with all three pseudotimes attached so Module 7
can use them as the time axis for gene-regulatory-network inference.

```{r save-for-module-7}
final_cache <- file.path(results_dir, "06_cm_pseudotime.rds")
saveRDS(cm, final_cache)
cat("Annotated CM object at", final_cache, "\n")
```

The object carries, in addition to everything Module 5 handed us:

- `reducedDim(cm, "PCA")` — 20 PCs on scran logcounts of HVGs.
- `colData(cm)$pt_sling_phi` / `pt_sling_pca` / `pt_dpt_pca` — raw
  pseudotimes from the three runs.
- `colData(cm)$pt_*_01` — the same pseudotimes rescaled to `[0, 1]`
  for cross-method comparison.

For Module 7 we will primarily use `pt_sling_phi` — the Φ-Space
slingshot pseudotime — as the canonical time axis. The other two live
in the object for optional sensitivity analyses.

## Summary

- Pseudotime gives a scalar ordering of cells along a biological
  process; it is not real time, and its direction for principal-curve
  methods comes from an external anchor.
- Slingshot is our primary method: cluster centroids → MST → principal
  curves → arc-length pseudotime. We ran it on both a PCA-of-expression
  embedding and on the Φ-Space score matrix to isolate the effect of
  the embedding.
- DPT via destiny is our diffusion cross-check on the PCA embedding.
  It adds two useful things: (i) automatic endpoint detection, and
  (ii) a fundamentally different statistical object (random-walk
  probabilities rather than arc length), which catches principal-curve
  artefacts.
- The three-pseudotime comparison is not bureaucratic — it tells you
  whether your trajectory is robust to embedding choice, method choice,
  or both.

## Session Info

```{r session-info}
sessionInfo()
```
