---
title: "Module 6: Pseudotime Trajectory Analysis"
author: "Single Cell Workshop"
date: "`r Sys.Date()`"
output:
  html_document:
    theme: flatly
    toc: true
    toc_depth: 2
    toc_float: true
    number_sections: true
    code_folding: show
    df_print: paged
vignette: >
  %\VignetteIndexEntry{Module 6: Pseudotime Trajectory Analysis}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
# Module 6 consumes Module 5's PhiSpace output. If that cache is missing
# (e.g. fresh clone, or CI before the Zenodo upload), skip execution but
# still render the prose so the site build succeeds.
data_ok <- file.exists("../results/05_phispace_query.rds")

knitr::opts_chunk$set(
    echo = TRUE,
    warning = FALSE,
    message = FALSE,
    fig.align = "center",
    fig.width = 10,
    fig.height = 6,
    eval = data_ok
)
```

```{asis, echo = !data_ok}
> **Note.** This module's code chunks are not evaluated because
> `results/05_phispace_query.rds` is missing. Render Module 5 first, or
> download the Zenodo cache.
```

## Introduction

In Module 5 we annotated each query cell with a continuous score on every
reference phenotype. Those scores already *hint* at a foetal→adult
gradient within cardiomyocytes — recall the age-group scores climbing
smoothly from `fetal` to `old`. Module 6 turns that hint into a single
scalar per cell: **pseudotime**.

Pseudotime is a real number per cell such that cells that are close in the
underlying biological process have close values. It is not real time, and
the absolute scale is not identified — it is an *ordering*. Direction
(which end is "early") is a separate question, and for principal-curve
methods like slingshot it has to be supplied externally. This is exactly
where the Φ-Space age-group scores earn their keep: the cluster with the
highest mean `fetal` score is an obvious anchor for the start of the
cardiomyocyte maturation trajectory.

### Two methods × two embeddings

We run two families of method on two different embeddings — four
pseudotimes in total, laid out on a 2 × 2 grid:

|                | Φ-Space scores  | PCA of expression |
|----------------|-----------------|-------------------|
| **slingshot**  | `pt_sling_phi`  | `pt_sling_pca`    |
| **DPT**        | `pt_dpt_phi`    | `pt_dpt_pca`      |

- **Slingshot** — a *principal-curve* method. Cluster cells, build a
  minimum spanning tree on the cluster centroids, fit a smooth curve
  through each lineage, project every cell onto the curve. Pseudotime =
  arc length.
- **DPT (diffusion pseudotime) via destiny** — a *diffusion / random
  walk* method. Build a Gaussian-kernel affinity graph, eigendecompose
  the associated Markov chain, and use accumulated transition
  probability from a root as pseudotime.

Crossing the two methods with two embeddings separates what is otherwise
confounded: a disagreement along a row of the table is an *embedding
effect* (same method, different embedding), a disagreement along a
column is a *method effect* (same embedding, different method), and a
disagreement on the diagonal is both. The comparison section below
reads out all four cells of this grid.

One thing destiny gets "for free" that slingshot does not: **endpoint
detection**. Without a `tips` argument, destiny picks trajectory tips
automatically from the diffusion-map geometry. We still choose to pin the
root at the highest-`fetal`-score cell, for comparability, but the
auto-detection is a genuine advantage of diffusion methods for
exploratory work where you do not yet know which end is which. The
flip side — less explicit control over endpoints — is a feature, not a
defect; it is just a different point on the "automated vs controllable"
axis.

### Learning objectives

By the end of this module you will be able to:

- Explain what pseudotime is, what it is not, and why it needs an
  external directionality anchor for principal-curve methods.
- Fit slingshot and DPT trajectories on two different embeddings
  (PCA and Φ-Space) and separate embedding effects from method effects.
- Visualise pseudotime and principal curves on a shared UMAP.
- Use a four-pseudotime correlation matrix as a sanity check on any
  single trajectory result.
- Save a pseudotime-annotated CM subset for Module 7.

## Load Libraries and Module 5 Output

```{r load-libraries}
library(SingleCellExperiment)
library(slingshot)
library(destiny)
library(scater)
library(ggplot2)
library(dplyr)
library(tibble)
library(patchwork)
library(RColorBrewer)
```

The handoff from Module 5 is a single file — the query SCE with PhiSpace
scores in `reducedDim`, scran logcounts in `assay(query, "logcounts")`,
and per-cell annotations (`PhiCellType`, `PhiCellTypeBroad`, `PhiAge`,
`group`, etc.) in `colData`.

```{r load-module5}
query <- readRDS("../results/05_phispace_query.rds")

cat("Assays:      ", paste(assayNames(query), collapse = ", "), "\n")
cat("reducedDims: ", paste(reducedDimNames(query), collapse = ", "), "\n")
cat("Cells:       ", ncol(query), "\n")
cat("Broad types: ", paste(sort(unique(query$PhiCellTypeBroad)),
                            collapse = ", "), "\n")
```

```{r cache-setup}
results_dir <- "../results"
dir.create(results_dir, showWarnings = FALSE, recursive = TRUE)
```

## Subset to Cardiomyocytes

Cardiomyocytes carry the foetal→adult maturation signal we want to trace.
Fibroblasts and the other lineages are interesting but secondary; running
pseudotime on the whole query would mix several biological processes
into one trajectory.

```{r subset-cm}
cm <- query[, query$PhiCellTypeBroad == "Cardiomyocyte"]
cat("Cardiomyocytes:", ncol(cm), "\n")
print(table(cm$group))
print(sort(table(cm$PhiCellType), decreasing = TRUE))
```

The fine Φ-Space labels give us biologically meaningful CM clusters
(`Cardiomyocyte:Fetal_enriched_*`, `:Metabolic`, `:Stressed`, …) without
any extra Leiden pass. We'll reuse them as the clustering input to
slingshot below.

## Expression-Based Embedding: PCA on scran logcounts

For the "classical" slingshot run we need an embedding of gene
expression. We compute a standard HVG → PCA pipeline on the scran
logcounts we already have from Module 5.

```{r pca-expression}
pca_cache <- file.path(results_dir, "06_cm_pca.rds")

if (file.exists(pca_cache)) {
    cm <- readRDS(pca_cache)
    message("Loaded cached CM object with PCA from ", pca_cache)
} else {
    # Highly variable gene selection
    gene_var <- rowVars(as.matrix(logcounts(cm)))
    names(gene_var) <- rownames(cm)
    n_hvg <- min(2000, nrow(cm))
    hvg <- names(sort(gene_var, decreasing = TRUE))[seq_len(n_hvg)]

    pca <- prcomp(t(as.matrix(logcounts(cm)[hvg, ])),
                  center = TRUE, scale. = FALSE)
    reducedDim(cm, "PCA") <- pca$x[, 1:20]
    saveRDS(cm, pca_cache)
}

cat("PCA dims:      ", paste(dim(reducedDim(cm, "PCA")), collapse = " x "), "\n")
cat("PhiSpace dims: ", paste(dim(reducedDim(cm, "PhiSpace")),
                             collapse = " x "), "\n")
```

We keep 20 PCs. That is a defensible default for single-cell data of this
size; tunePhiSpace-style cross-validation of the component count is
possible but not the point here.

## UMAP for Visualisation

We compute a UMAP on the PCA embedding purely for visualisation — none
of the pseudotime methods below take UMAP as input. A single shared
UMAP lets us overlay the four pseudotimes, and the slingshot curves,
on the same two-dimensional layout.

```{r umap}
umap_cache <- file.path(results_dir, "06_cm_umap.rds")

if (file.exists(umap_cache)) {
    reducedDim(cm, "UMAP") <- readRDS(umap_cache)
    message("Loaded cached UMAP")
} else {
    set.seed(42)
    cm <- scater::runUMAP(cm, dimred = "PCA", name = "UMAP")
    saveRDS(reducedDim(cm, "UMAP"), umap_cache)
}
```

## Picking the Trajectory Anchor

For slingshot we need a `start.clus`. Rather than hand-picking a root
cell, we pick the CM *cluster* whose mean score on the reference `fetal`
phenotype is highest — a fully data-driven anchor.

```{r pick-anchor}
phi_scores <- reducedDim(cm, "PhiSpace")
fetal_score <- phi_scores[, "fetal"]

anchor_tbl <- tibble(
    cluster = cm$PhiCellType,
    fetal   = fetal_score
) |>
    group_by(cluster) |>
    summarise(mean_fetal = mean(fetal), n = n(),
              .groups = "drop") |>
    arrange(desc(mean_fetal))

print(anchor_tbl)

start_cluster <- anchor_tbl$cluster[1]
cat("\nAnchor (start.clus):", start_cluster, "\n")
```

We also pre-compute the per-cell root index for destiny, which wants a
cell rather than a cluster.

```{r root-cell}
start_cells <- which(cm$PhiCellType == start_cluster)
root_cell   <- start_cells[which.max(fetal_score[start_cells])]
cat("Root cell index:", root_cell, "\n")
```

## Slingshot on Φ-Space Scores

```{r slingshot-phispace}
sling_phi_cache <- file.path(results_dir, "06_slingshot_phispace.rds")

if (file.exists(sling_phi_cache)) {
    sling_phi <- readRDS(sling_phi_cache)
    message("Loaded cached slingshot-PhiSpace result")
} else {
    sling_phi <- slingshot(
        reducedDim(cm, "PhiSpace"),
        clusterLabels = cm$PhiCellType,
        start.clus    = start_cluster,
        dist.method   = "simple"
    )
    saveRDS(sling_phi, sling_phi_cache)
}

cat("Lineages found:\n")
print(slingLineages(sling_phi))
```

Slingshot's `slingPseudotime()` returns one column per lineage; for a
predominantly linear CM maturation trajectory we expect one dominant
lineage.

```{r pt-phi}
pt_phi <- slingPseudotime(sling_phi)
cm$pt_sling_phi <- rowMeans(pt_phi, na.rm = TRUE)
summary(cm$pt_sling_phi)
```

## Slingshot on PCA of Expression

Same clusters, same anchor, different embedding.

```{r slingshot-pca}
sling_pca_cache <- file.path(results_dir, "06_slingshot_pca.rds")

if (file.exists(sling_pca_cache)) {
    sling_pca <- readRDS(sling_pca_cache)
    message("Loaded cached slingshot-PCA result")
} else {
    sling_pca <- slingshot(
        reducedDim(cm, "PCA"),
        clusterLabels = cm$PhiCellType,
        start.clus    = start_cluster,
        dist.method   = "simple"
    )
    saveRDS(sling_pca, sling_pca_cache)
}

pt_pca <- slingPseudotime(sling_pca)
cm$pt_sling_pca <- rowMeans(pt_pca, na.rm = TRUE)
summary(cm$pt_sling_pca)
```

## DPT via destiny

For the diffusion branch of the 2 × 2 grid we build a diffusion map on
each embedding and compute DPT from the same root cell we pinned
above.

### DPT on PCA

```{r destiny-dpt-pca}
dpt_cache <- file.path(results_dir, "06_dpt_pca.rds")

if (file.exists(dpt_cache)) {
    dpt <- readRDS(dpt_cache)
    message("Loaded cached DPT/PCA result")
} else {
    dm  <- DiffusionMap(as.matrix(reducedDim(cm, "PCA")),
                        n_eigs = 20, sigma = "local", verbose = FALSE)
    dpt <- DPT(dm, tips = root_cell)
    saveRDS(dpt, dpt_cache)
}

cm$pt_dpt_pca <- dpt$dpt
summary(cm$pt_dpt_pca)
```

### DPT on Φ-Space

Φ-Space gives us only 17 coordinates (13 cell types + 4 age groups), so
we cap `n_eigs` accordingly; everything else is identical to the PCA
run.

```{r destiny-dpt-phi}
dpt_phi_cache <- file.path(results_dir, "06_dpt_phispace.rds")

if (file.exists(dpt_phi_cache)) {
    dpt_phi <- readRDS(dpt_phi_cache)
    message("Loaded cached DPT/PhiSpace result")
} else {
    n_eigs_phi <- min(15, ncol(reducedDim(cm, "PhiSpace")) - 1)
    dm_phi <- DiffusionMap(as.matrix(reducedDim(cm, "PhiSpace")),
                           n_eigs = n_eigs_phi, sigma = "local",
                           verbose = FALSE)
    dpt_phi <- DPT(dm_phi, tips = root_cell)
    saveRDS(dpt_phi, dpt_phi_cache)
}

cm$pt_dpt_phi <- dpt_phi$dpt
summary(cm$pt_dpt_phi)
```

A note on what we *did not* pass: if we had called `DPT(dm)` with no
`tips`, destiny would have auto-detected trajectory endpoints from the
diffusion-map geometry — that is the "endpoints for free" property of
diffusion methods. We pass the root explicitly here only to keep
destiny's pseudotime directly comparable with the slingshot runs.

## Visualising Pseudotimes on UMAP

Before any quantitative comparison, it helps to *see* the four
pseudotimes on a shared layout. We colour cells by pseudotime using a
blue → yellow → red scale — low values blue, mid yellow, high red —
which is the most intuitive encoding for an ordered scalar (reversible
methods like viridis wash out the two endpoints against each other).

For the two slingshot panels we overlay the principal curves on top of
the UMAP using `slingshot::embedCurves()`, which re-fits the curves in
any reduced-dimensional space given the existing cell ordering. DPT is
a per-cell scalar with no curve object, so those two panels show only
the pseudotime colouring.

```{r rescale-pt}
rescale01 <- function(x) {
    x <- x - min(x, na.rm = TRUE)
    x / max(x, na.rm = TRUE)
}

cm$pt_sling_phi_01 <- rescale01(cm$pt_sling_phi)
cm$pt_sling_pca_01 <- rescale01(cm$pt_sling_pca)
cm$pt_dpt_phi_01   <- rescale01(cm$pt_dpt_phi)
cm$pt_dpt_pca_01   <- rescale01(cm$pt_dpt_pca)
```

```{r umap-curves}
umap_df <- as.data.frame(reducedDim(cm, "UMAP"))
colnames(umap_df) <- c("UMAP1", "UMAP2")

# Embed slingshot curves from their native embeddings into UMAP space.
sling_phi_umap <- embedCurves(sling_phi, reducedDim(cm, "UMAP"))
sling_pca_umap <- embedCurves(sling_pca, reducedDim(cm, "UMAP"))

curve_df <- function(sds, label) {
    curves <- slingCurves(sds)
    do.call(rbind, lapply(seq_along(curves), function(i) {
        crv <- curves[[i]]
        s   <- crv$s[crv$ord, , drop = FALSE]
        data.frame(UMAP1 = s[, 1], UMAP2 = s[, 2],
                   lineage = paste0(label, "_L", i))
    }))
}

curves_phi_df <- curve_df(sling_phi_umap, "phi")
curves_pca_df <- curve_df(sling_pca_umap, "pca")
```

```{r umap-pt-plots, fig.width=12, fig.height=10}
pt_scale <- scale_colour_gradientn(
    colours = c("blue", "yellow", "red"),
    name    = "Pseudotime"
)

umap_base <- function(pt_vec) {
    df <- cbind(umap_df, pt = pt_vec)
    ggplot(df, aes(UMAP1, UMAP2, colour = pt)) +
        geom_point(size = 0.4, alpha = 0.7) +
        pt_scale +
        theme_bw(base_size = 11) +
        theme(aspect.ratio = 1)
}

p_sling_phi <- umap_base(cm$pt_sling_phi_01) +
    geom_path(data = curves_phi_df,
              aes(UMAP1, UMAP2, group = lineage),
              inherit.aes = FALSE,
              colour = "black", linewidth = 0.8) +
    labs(title = "slingshot / Φ-Space")

p_sling_pca <- umap_base(cm$pt_sling_pca_01) +
    geom_path(data = curves_pca_df,
              aes(UMAP1, UMAP2, group = lineage),
              inherit.aes = FALSE,
              colour = "black", linewidth = 0.8) +
    labs(title = "slingshot / PCA")

p_dpt_phi <- umap_base(cm$pt_dpt_phi_01) +
    labs(title = "DPT / Φ-Space")

p_dpt_pca <- umap_base(cm$pt_dpt_pca_01) +
    labs(title = "DPT / PCA")

(p_sling_phi | p_sling_pca) / (p_dpt_phi | p_dpt_pca) +
    plot_layout(guides = "collect") +
    plot_annotation(title = "Pseudotime on UMAP — two methods × two embeddings")
```

Read the figure column-by-column and row-by-row:

- **Rows** (method). Slingshot curves emphasise a single backbone;
  DPT spreads cells along whichever geometry the diffusion kernel
  picks out.
- **Columns** (embedding). Φ-Space forces the trajectory through
  cells whose cell-type and age scores change smoothly. PCA can pick
  up other axes of variation (donor, batch, technical) that Φ-Space
  never saw.

If all four panels paint the same blue → red gradient across the same
UMAP regions, the trajectory is robust. If the two slingshot panels
agree but the two DPT panels disagree with them, that is a pure method
effect; if the two PhiSpace panels agree but the PCA panels do not,
that is a pure embedding effect.

## Four-Way Comparison

```{r pt-mat}
pt_mat <- as.data.frame(colData(cm)[,
    c("pt_sling_phi_01", "pt_sling_pca_01",
      "pt_dpt_phi_01",   "pt_dpt_pca_01")])
colnames(pt_mat) <- c("slingshot/PhiSpace", "slingshot/PCA",
                      "DPT/PhiSpace",       "DPT/PCA")
```

### Correlation matrix

```{r pt-cor}
round(cor(pt_mat, use = "complete.obs", method = "spearman"), 3)
```

Spearman (rank) correlation is the right scalar summary because each
method has a different absolute scale and we only care about the *order*
of cells.

### Pairs plot

```{r pt-pairs, fig.width=10, fig.height=10}
pt_long <- pt_mat |>
    rownames_to_column("cell") |>
    mutate(group = cm$group[match(cell, colnames(cm))])

make_pair <- function(x, y) {
    ggplot(pt_long, aes(.data[[x]], .data[[y]], colour = group)) +
        geom_point(size = 0.4, alpha = 0.5) +
        geom_abline(slope = 1, intercept = 0,
                    linetype = "dashed", alpha = 0.5) +
        scale_colour_brewer(palette = "Set1") +
        theme_bw(base_size = 11) +
        theme(legend.position = "none") +
        labs(x = x, y = y)
}

method_names <- c("slingshot/PhiSpace", "slingshot/PCA",
                  "DPT/PhiSpace",       "DPT/PCA")
pair_idx <- combn(method_names, 2, simplify = FALSE)
pairs <- lapply(pair_idx, function(p) make_pair(p[1], p[2]))

wrap_plots(pairs, ncol = 3) +
    plot_annotation(title = "Pairwise agreement between pseudotimes")
```

Read the 2 × 2 grid of methods diagonally:

- **Rows of the method table** (slingshot/PhiSpace ↔ slingshot/PCA, or
  DPT/PhiSpace ↔ DPT/PCA) isolate the *embedding* effect — same method,
  different embedding.
- **Columns of the method table** (slingshot/PhiSpace ↔ DPT/PhiSpace,
  or slingshot/PCA ↔ DPT/PCA) isolate the *method* effect — same
  embedding, different trajectory family.
- **Anti-diagonal pairs** (slingshot/PhiSpace ↔ DPT/PCA, slingshot/PCA
  ↔ DPT/PhiSpace) vary both at once; agreement here is the strongest
  evidence the trajectory is not an artefact of either choice.

### Pseudotime vs donor stage

```{r pt-vs-group, fig.width=12, fig.height=4}
box_df <- pt_long |>
    tidyr::pivot_longer(
        cols = all_of(method_names),
        names_to = "method", values_to = "pseudotime"
    ) |>
    mutate(method = factor(method, levels = method_names))

ggplot(box_df, aes(group, pseudotime, fill = group)) +
    geom_violin(scale = "width", alpha = 0.8) +
    geom_boxplot(width = 0.15, alpha = 0.6, outlier.size = 0.3) +
    scale_fill_brewer(palette = "Set1") +
    facet_wrap(~ method, nrow = 1) +
    labs(x = "Donor stage", y = "Pseudotime (rescaled)",
         fill = "Stage",
         title = "Do pseudotimes recover the known foetal → adult ordering?") +
    theme_bw(base_size = 12) +
    theme(legend.position = "none")
```

The donor stage (`group`) is held-out information: none of the four
pseudotime methods saw it. If the violins step upward cleanly from
`fetal` to `adult` on all four panels, the trajectory is recovering the
real maturation order.

## Handing Off to Module 7

We save the CM object with all four pseudotimes attached so Module 7
can use them as the time axis for gene-regulatory-network inference.

```{r save-for-module-7}
final_cache <- file.path(results_dir, "06_cm_pseudotime.rds")
saveRDS(cm, final_cache)
cat("Annotated CM object at", final_cache, "\n")
```

The object carries, in addition to everything Module 5 handed us:

- `reducedDim(cm, "PCA")` — 20 PCs on scran logcounts of HVGs.
- `reducedDim(cm, "UMAP")` — 2D UMAP on the PCA embedding, for plotting.
- `colData(cm)$pt_sling_phi` / `pt_sling_pca` / `pt_dpt_phi` /
  `pt_dpt_pca` — raw pseudotimes from the four runs.
- `colData(cm)$pt_*_01` — the same pseudotimes rescaled to `[0, 1]`
  for cross-method comparison.

For Module 7 we will primarily use `pt_sling_phi` — the Φ-Space
slingshot pseudotime — as the canonical time axis. The other three
live in the object for optional sensitivity analyses.

## Summary

- Pseudotime gives a scalar ordering of cells along a biological
  process; it is not real time, and its direction for principal-curve
  methods comes from an external anchor.
- We ran two methods on two embeddings — slingshot and DPT, each on
  PCA and on Φ-Space — giving four pseudotimes in a 2 × 2 grid.
- Slingshot is our principal-curve method: cluster centroids → MST →
  principal curves → arc-length pseudotime. We overlay its curves on
  the UMAP to make the trajectory visible.
- DPT via destiny is our diffusion cross-check. It adds two useful
  things: (i) automatic endpoint detection, and (ii) a fundamentally
  different statistical object (random-walk probabilities rather than
  arc length), which catches principal-curve artefacts.
- The four-pseudotime comparison is not bureaucratic — crossing method
  with embedding is what lets us separate embedding effects from method
  effects and decide whether the trajectory is robust to either.

## Session Info

```{r session-info}
sessionInfo()
```
